---
title: "Imputation with beagle"
author: "Christen Bossu"
date: "10/21/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Programs you will need
beagle: for imputation
bcftools: for counting variants per scaffold
vcftools: for filtering out scaffolds that only have one variant

## Bash and unix scripting
Most of this will be done with bash or scripting in the terminal. I'm using Rmarkdown because I like the code chunks. But you really won't be using R (except for maybe plotting something).

For the imputation, we want to remove any scaffolds or chromosomes that have a small number of loci on them (this will make the imputation bad). Another way to count the number of variants per chromosome is with bcftools. First, you need to bgzip our vcfile and index it with bcftools. I have bcftools loaded in a conda environment.

```{bash}
vcf="/scratch/alpine/ericacnr@colostate.edu/BANS/01.fastq_processing/mega-non-model-wgs-snakeflow/results/bqsr-round-0/vcf/filtered_HC/BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.recode.vcf"

conda activate bioinf

bgzip $vcf

gzvcf="/scratch/alpine/ericacnr@colostate.edu/BANS/01.fastq_processing/mega-non-model-wgs-snakeflow/results/bqsr-round-0/vcf/filtered_HC/BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.recode.vcf.gz"

#from 02.imputation
bcftools index $gzvcf

bcftools index -s "$gzvcf" | awk '$3 >= 10 {print $1"\t1\t"$2}' > keep_scaffolds.regions.txt

bcftools view \
  -R keep_scaffolds.regions.txt \
  -Oz \
  -o BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.rm_scaf_lth10.vcf.gz \
  "$gzvcf"
  
bcftools index -t BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.rm_scaf_lth10.vcf.gz 
```

# Problem Shooting
Okay so the imputation in repeatedly failing at the same SNP (CM026855.1	15386922), so I'm going to try removing that SNP and see if it can complete the imputation properly. I can't see anything wrong with this SNP that might be causing it to fail but I need this done!!

```{bash}
bcftools view \
   --exclude 'CHROM="CM026855.1" && POS=15386922' \
   -Oz -o BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.rm_scaf_lth10.recode.gdsnps.vcf.gz \
   BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.rm_scaf_lth10.recode.vcf.gz

#verify that it worked
bcftools index BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.rm_scaf_lth10.recode.gdsnps.vcf.gz
bcftools view -r CM026855.1:15386922 BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.rm_scaf_lth10.recode.gdsnps.vcf.gz
```

Trying again.
```{bash}
BEAGLE_JAR="/projects/ericacnr@colostate.edu/mambaforge/envs/bioinf/java/beagle.27Jan18.7e1.jar"

vcf="BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.rm_scaf_lth10.recode.gdsnps.vcf.gz"

java -Xss5m -Xmx90g -jar $BEAGLE_JAR \
  gtgl=$vcf \
  out=BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.imputed4.1.gdsnps.vcf \
  nthreads=25
```

Now I'm getting a haploid error for this SNP: CM026855.1	23395646
```{r}
bcftools view \
   --exclude 'CHROM="CM026855.1" && POS=23395646' \
   -Oz -o BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.rm_scaf_lth10.recode.gdsnps2.vcf.gz \
   BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.rm_scaf_lth10.recode.gdsnps.vcf.gz
```


## splitting up by scaffolds
The above didn't work, it just failed one snp over.

```{bash}
IN_VCF="BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.rm_scaf_lth10.recode.vcf.gz"

bcftools query -f '%CHROM\n' "$IN_VCF" | uniq -u > chr_list.txt
```

making big_scaffolds

cat <<EOF > big_scaffolds.txt
...
EOF

cat <<EOF > small_scaffolds.txt
...
EOF

within a sub directory split_vcf...
```{bash}
while read -r CHR; do
  echo "Creating VCF for big scaffold $CHR ..."
  bcftools view -r "$CHR" "$IN_VCF" -Oz -o BANS.big.${CHR}.vcf.gz
  tabix -p vcf BANS.big.${CHR}.vcf.gz
done < big_scaffolds.txt

bcftools view -r small_scaffolds.txt "$IN_VCF" -Oz -o BANS.small_scaffolds.vcf.gz
tabix -p vcf BANS.small_scaffolds.vcf.gz
```

Now building a loop
```{bash}
#!/bin/bash
#
#SBATCH --job-name=impute
#SBATCH --output=impute.%j.out
#SBATCH --error=impute.%j.err
#SBATCH -t 1:00:00
#SBATCH --partition=amilan
#SBATCH --qos=normal
#SBATCH --nodes=1
#SBATCH --ntasks-per-node 11
#SBATCH --mem=40G
#SBATCH --mail-type=BEGIN,END,FAIL
#SBATCH  --mail-user=ericacnr@colostate.edu

#echo commands to stdout

set -x
source ~/.bashrc

#for chrom in `ls BANS.big.* | cut -f3 -d'.'`; do echo $chrom; sbatch imputation.sbatch $chrom; done

BEAGLE_JAR="/projects/ericacnr@colostate.edu/mambaforge/envs/bioinf/java/beagle.27Jan18.7e1.jar"

chrom=$1

java -Xss5m -Xmx90g -jar $BEAGLE_JAR \
  gtgl=BANS.big."$chrom".1.vcf.gz \
  out=BANS."$chrom" \
  nthreads=25
```

and the small chroms
```{bash}
#!/bin/bash
#
#SBATCH --job-name=impute
#SBATCH --output=impute.%j.out
#SBATCH --error=impute.%j.err
#SBATCH -t 1:30:00
#SBATCH --partition=amilan
#SBATCH --qos=normal
#SBATCH --nodes=1
#SBATCH --ntasks-per-node 11
#SBATCH --mem=40G
#SBATCH --mail-type=BEGIN,END,FAIL
#SBATCH  --mail-user=ericacnr@colostate.edu

#echo commands to stdout

set -x
source ~/.bashrc

BEAGLE_JAR="/projects/ericacnr@colostate.edu/mambaforge/envs/bioinf/java/beagle.27Jan18.7e1.jar"

vcf="BANS.small_scaffolds.vcf.gz"

java -Xss5m -Xmx90g -jar $BEAGLE_JAR \
  gtgl=$vcf \
  out=BANS.small \
  nthreads=25
```

make a list of the files called imputed_vcfs.list

```{bash}
bcftools concat \
  -Oz \
  -o BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.imputed4.1.vcf.gz \
  $(cat imputed_vcfs.list)
  
tabix -p vcf BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.imputed4.1.vcf.gz
bcftools stats BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.imputed4.1.vcf.gz > BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.imputed4.1.stats
```

Hilarious, now I've got more snps than expected! Originally 3123809, now 4765431. Gained 1641622??

Checking for dups
```{bash}
bcftools query -f '%CHROM\t%POS\n' BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.imputed4.1.vcf.gz \
  | sort \
  | uniq -d > duplicated_positions.txt
  
wc -l duplicated_positions.txt
```

Hmmm

```{bash}
FASTA="/scratch/alpine/ericacnr@colostate.edu/BANS/01.fastq_processing/mega-non-model-wgs-snakeflow/resources/genome.fasta"
bcftools norm \
  -d all \
  -f "$FASTA" \
  BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.imputed4.1.vcf.gz \
  -Oz -o BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.imputed4.1_nodup.vcf.gz
```

So, checking each separate file.
```{bash}
for orig in BANS.big.CM*.vcf.gz; do
  # strip path and .vcf.gz
  base=$(basename "$orig" .vcf.gz)

  # base looks like: BANS.big.CM026856.1
  # extract scaffold with suffix (.1) and without
  scaf_with_suffix=${base#BANS.big.}      # CM026856.1
  scaf=${scaf_with_suffix%.1}             # CM026856

  # corresponding imputed file name (adjust if yours differ)
  imp="BANS.${scaf}.vcf.gz"

  if [ ! -f "$imp" ]; then
    echo "WARNING: missing imputed file for $scaf: $imp" >&2
    echo -e "${scaf}\tNA\tNA\t${orig}\tMISSING_${imp}" >> scaffold_mismatches.tsv
    continue
  fi

  # count non-header variants
  n_orig=$(bcftools view -H "$orig" | wc -l)
  n_imp=$(bcftools view -H "$imp" | wc -l)

  if [ "$n_orig" -ne "$n_imp" ]; then
    echo "Mismatch for $scaf: orig=$n_orig, imputed=$n_imp"
    echo -e "${scaf}\t${n_orig}\t${n_imp}\t${orig}\t${imp}" >> scaffold_mismatches.tsv
  fi
done
```


Mismatch for CM026848: orig=364028, imputed=728056
Mismatch for CM026849: orig=342789, imputed=685578
Mismatch for CM026851: orig=225746, imputed=451492
Mismatch for CM026852: orig=204428, imputed=408856
Mismatch for CM026854: orig=112070, imputed=224140
Mismatch for CM026855: orig=105645, imputed=97000
Mismatch for CM026856: orig=94010, imputed=188020
Mismatch for CM026857: orig=76220, imputed=152440
Mismatch for CM026858: orig=78175, imputed=156350
Mismatch for CM026859: orig=74241, imputed=148482
Mismatch for CM026861: orig=60609, imputed=121218
Mismatch for CM026869: orig=21765, imputed=43530
Mismatch for CM026870: orig=17118, imputed=34236
Mismatch for CM026871: orig=16637, imputed=33274
Mismatch for CM026872: orig=12843, imputed=25686
Mismatch for CM026873: orig=16679, imputed=33358
Mismatch for CM026874: orig=9354, imputed=18708
Mismatch for CM026884: orig=24280, imputed=48560

So most of them are fucked. Most of them are doubled, which explains the higher post cat count. Removing them all, redoing them all.

Haha we're back to the cursed Chrom CM026855.1! 23395646 is showing as haploid for at least one indv... I stg if that was the only problem and I did all this...

```{r}
bcftools view \
   --exclude 'CHROM="CM026855.1" && POS=23395646' \
   -Oz -o BANS.big.CM026855.1.2.vcf.gz \
   BANS.big.CM026855.1.vcf.gz

bcftools index BANS.big.CM026855.1.2.vcf.gz
bcftools view -r CM026855.1:23395646 BANS.big.CM026855.1.2.vcf.gz
```

Finally! After reimputing that one, I think everything is as it should be (minus that one weird SNP).

```{bash}
bcftools concat \
  -Oz \
  -o BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.imputed4.1.final.vcf.gz \
  $(cat imputed_vcfs.list)
  
tabix -p vcf BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.imputed4.1.final.vcf.gz
bcftools stats BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.imputed4.1.final.vcf.gz > BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.imputed4.1.final.stats

#double check for missingness
vcftools --gzvcf BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.imputed4.1.final.vcf.gz --missing-indv
bcftools view -H -i 'GT="mis"' BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.imputed4.1.final.vcf.gz | wc -l
```

Incredible! Redid the cat and I finally have 3123808 SNPs. Unclear why removing that weird SNP didn't solve the issue before breaking up the chroms, or maybe it did and I missed it somehow. Anyway, onward!

/scratch/alpine/ericacnr@colostate.edu/BANS/02.imputation/BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.imputed4.1.vcf.gz



