---
title: "03.a.initial_popstructure"
output: html_document
---

# Libraries
```{r}
library(tidyverse)
library(data.table)
library(dplyr)
library(rnaturalearth)
library(geosphere)
library(ggplot2)
library(rnaturalearth)
library(raster)
library(terra)
```

# Pruning
```{bash}
#GIVING SNP IDS
vcf="/scratch/alpine/ericacnr@colostate.edu/BANS/02.imputation/BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.imputed4.1.final.vcf.gz"

conda activate GWAS2
plink --vcf $vcf --aec --recode --out BANS.all.recode

cat BANS.all.recode.map | awk -F"\t" '{split($1, a, "__"); print $1"\t"a[1]"."$4"\t"$3"\t"$4}' > BANS.all.recode_loc.map

plink --ped BANS.all.recode.ped \
--map BANS.all.recode_loc.map \
--make-bed --aec \
--out BANS.all.recode

bfile="BANS.all.recode"

plink --bfile $bfile --aec \
--indep-pairwise 25 10 0.5 --out PC1_allindv
#278091 of 3123808 variants removed.

plink --bfile $bfile --aec --extract PC1_allindv.prune.in \
  --make-bed --out pruned_allindv
#2845717 variants and 209 people pass filters and QC.

plink --bfile pruned_allindv --aec --allow-no-sex --pca 208 \
  --out PC1_allindv.ld25-10-0.5
#2845717 variants and 209 people pass filters and QC.
```

```{bash}
rsync -avzP ericacnr@colostate.edu@login.rc.colorado.edu:/scratch/alpine/ericacnr@colostate.edu/BANS/02.5.neutral_pop_struc/PC1_allindv.ld25-10-0.5*eigen* /Users/ericarobertson/Desktop/BANS_adaptive_units/analysis/02.5.neutral_pop_struc/
```
# Time to plot the PCA.
## grouping pops
The actual group pops is missing indv that I filtered for missingness, so this is a quick temporary one...
```{r}
meta <- read.csv("../../../raw_data/BANS_all_sample_data_shifted.csv")
coords <- meta %>% dplyr::select(Long, Lat)
dist_matrix <- distm(as.matrix(coords), fun = distHaversine)

libraries <- read.delim("../../01.fastq_processing/snakemake/config_files/BANS.Plate1.2.3.units.tsv") %>% rename(BGP_ID=sample)

lib.meta <- left_join(meta, libraries %>% dplyr::select(library, BGP_ID), by="BGP_ID")

# clustering
hclust_groups <- hclust(as.dist(dist_matrix), method = "complete") # complete method here avoids chaining, ensuring all samples in a cluster are within the specified distance of each other. If you want some flexibility you could try "average"

# cut tree at 111.111 km to assign spatial clusters
meta$GroupID <- cutree(hclust_groups, h = 111111)

# create population labels and filter for minimum cluster size
meta.sub <- meta %>%
  mutate(Pop = paste0("Cluster_", GroupID)) %>%
  group_by(Pop) %>%
  filter(n() > 4) %>%  # keep clusters with ≥4 individuals
  ungroup()

pops <- data.frame(BGP_ID = meta.sub$BGP_ID, Group = meta.sub$Pop)

pops %>% write_csv("../scratch/pops.csv")

map <- ne_states(country = c("United States of America", "Canada", "Mexico"), 
                 returnclass = 'sf')

# create colors for clusters
n_clusters <- length(unique(meta.sub$Pop))
cluster_colors <- rainbow(n_clusters)
names(cluster_colors) <- unique(meta.sub$Pop)

meta.b <- meta.sub %>%
  group_by(Pop) %>%
  mutate(cluster_size = n()) %>%
  ungroup()

# plot it
p <- ggplot() +
  geom_sf(data = map, fill = "white", color = "black", size = 0.1) +
  geom_point(data = meta.b, aes(x = Long, y = Lat, color = Pop),
             size = 1.5, position = position_jitter(width = 0.1, height = 0.1)) +
  geom_text(data = meta.b,
            aes(x = Long, y = Lat, label = Pop),
            size = 2, vjust = -0.6) +
  scale_colour_manual(values = cluster_colors, name = "Spatial Clusters", guide = "none") +
  coord_sf(xlim = c(-145, -55), ylim = c(15, 80)) + 
  theme_minimal() +
  labs(title = "All clusters with ≥4 individuals",
       x = "Longitude",
       y = "Latitude")
p
ggsave("../outputs/original_cluster_plot.png")
```

## plotting PCA
### color scheme
Quick, somewhat neutral color scheme based on geography alone...
```{r}
# # Define clusters by region
# West_clusters    <- c("Cluster_1", "Cluster_5", "Cluster_16", "Cluster_4", "Cluster_9", "Cluster_12", "Cluster_2", "Cluster_20")
# Central_clusters <- c("Cluster_3", "Cluster_8", "Cluster_17", "Cluster_21", "Cluster_18", "Cluster_19")
# East_clusters    <- c("Cluster_6", "Cluster_14", "Cluster_11", "Cluster_10", "Cluster_13", "Cluster_15", "Cluster_7")
# 
# # Assign colors by cluster
# library(RColorBrewer)
# W.cols <- setNames(brewer.pal(n = length(West_clusters), "Reds"), West_clusters)
# C.cols <- setNames(brewer.pal(n = length(Central_clusters), "Blues"), Central_clusters)
# E.cols <- setNames(brewer.pal(n = length(East_clusters), "Greens"), East_clusters)
# bg <- c(W.cols, C.cols, E.cols)
# saveRDS(bg, file = "../figures/colors/bg_colors.rds")
bg <- readRDS("../figures/colors/bg_colors.rds")
```
### plotting
```{r}
plink_pca <- read.table("../scratch/PCA/PC1_allindv.ld25-10-0.5.eigenvec", header = FALSE) %>% dplyr::select(-V2)

#just saving a sample list real quick...
plink_pca <- read.table("../scratch/PCA/PC1_allindv.ld25-10-0.5.eigenvec", header = FALSE) %>% dplyr::select(V1) %>% rename(BGP_ID=V1) %>% write_csv("../scratch/BANS_allind_sample_list.csv")

eigenval <- scan("../scratch/PCA/PC1_allindv.ld25-10-0.5.eigenval")

pve05 <- data.frame(PC = 1:208, pve = eigenval/sum(eigenval)*100)

# set names
pca05 <- plink_pca
colnames(pca05)[-1] <- paste0("PC", 1:(ncol(pca05) - 1))
names(pca05)[1] <- "ind"
nrow(pca05)
#correct number of samples

pops <- read.csv("../scratch/pops.csv")

# Join PCA data
pca.cluster <- left_join(pca05, pops %>% rename(ind=BGP_ID), by="ind")  # left_join PCA first to preserve PC columns

# Factor for cluster order
region_order <- c(
  "Cluster_1", "Cluster_5", "Cluster_16", "Cluster_4", "Cluster_9", "Cluster_12", "Cluster_2", "Cluster_20",
  "Cluster_3", "Cluster_8", "Cluster_17", "Cluster_21", "Cluster_18", "Cluster_19",
  "Cluster_6", "Cluster_14", "Cluster_11", "Cluster_10", "Cluster_13", "Cluster_15", "Cluster_7"
)
pca.cluster$Cluster <- factor(pca.cluster$Group, levels = region_order)

# Plot
b05 <- ggplot(pca.cluster, aes(PC1, PC2, col = Cluster)) +
  geom_point(size = 3) +
  scale_color_manual(values = bg) +
  theme_light() +
  xlab(paste0("PC1 (", signif(pve05$pve[1], 3), "%)")) +
  ylab(paste0("PC2 (", signif(pve05$pve[2], 3), "%)"))

b05
ggsave("../figures/pca_plots/BANS_allind_overall_PCA.png", width=8)
```

Useful place names
```{r}
group_key <- tibble(
  Group = unique(pops$Group),
  Group_label = c(
    "Fairbanks, AK",
    "Royal City, WA",
    "Brightwater Res, SK",
    "Peace River, BC",
    "Whitehorse, YT",
    "Port Bruce, ON", #6 is pulling some PE stuff closer to Ontario into the same cluster
    "Wood Islands, PE",
    "Fort Ellice, MB",
    "Edmonton, AB",
    "Sackville, NB",
    "Meteghan, NS",
    "Shuswap Band Lands, BC",
    "D'Aiguillon, QC",
    "Odanak, QC",
    "Parson's Pond, NL",
    "Behchoko, NT",
    "Oak Hammock, MB",
    "Buffalo, WY",
    "Cheyenne, WY",
    "Missoula, MT",
    "Jamestown, ND"
  )
)
saveRDS(group_key, "../scratch/geo_informative_labels.RDS")
```

```{r}
meta.b.plot <- meta.b %>%
  rename(Group = Pop) %>%   # only needed if meta.b uses Pop instead of Group
  left_join(group_key, by = "Group")

bg <- readRDS("../figures/colors/bg_colors.rds")
p <- ggplot() +
  geom_sf(data = map, fill = "white", color = "black", linewidth = 0.1) +
  geom_point(
    data = meta.b.plot,
    aes(x = Long, y = Lat, color = Group),
    size = 1.5,
    position = position_jitter(width = 0.1, height = 0.1)
  ) +
  geom_text(
    data = meta.b.plot,
    aes(x = Long, y = Lat, label = Group_label),
    size = 2,
    vjust = -0.6
  ) +
  scale_colour_manual(values = bg, name = "Spatial Clusters", guide = "none") +
  coord_sf(xlim = c(-145, -55), ylim = c(15, 80)) +
  theme_minimal() +
  labs(
    title = "All clusters with ≥4 individuals",
    x = "Longitude",
    y = "Latitude"
  )

p

ggsave("../figures/BANS_allind_cluster_map_bg.png", plot = p, width = 10, height = 7, dpi = 300)
```

# Remove Related Individuals
## NGSrelate
Using NGSrelate
```{bash, label="ngsRelate install"}
conda activate bioinf
#stuff I need to make the libraries install correctly
mamba install -c conda-forge bzip2 zlib xz libcurl

git clone --recursive https://github.com/SAMtools/htslib
#this misses a necessary thing libhts.a so gotta do it again
cd /projects/ericacnr@colostate.edu/htslib
make clean

# make sure it uses conda include/lib
make -j2 libhts.a \
  CPPFLAGS="-I$CONDA_PREFIX/include" \
  LDFLAGS="-L$CONDA_PREFIX/lib" \
  LIBS="-lz -lbz2 -llzma -lcurl"
  
# get ngsRelate
git clone https://github.com/ANGSD/ngsRelate
cd htslib/;make -j2;cd ../ngsRelate;make HTSSRC=../htslib/
```

Running on just the one big chromosome.
```{bash}
vcf="/scratch/alpine/ericacnr@colostate.edu/BANS/02.imputation/BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.imputed4.1.final.vcf.gz"

chr="CM026847.1"

bcftools $vcf
bcftools view -r $chr -Oz -o BANS.${chr}.vcf.gz $vcf

bcftools index -t BANS.${chr}.vcf.gz
```

```{bash}
#!/bin/bash
#
#SBATCH --job-name=relate
#SBATCH --output=relate.%j.out
#SBATCH --error=relate.%j.err
#################
#SBATCH -t 2:00:00
#SBATCH -p amilan
#SBATCH --qos=normal
#SBATCH --nodes=1
#SBATCH --ntasks-per-node 5
#SBATCH --mem=12G
#################
#SBATCH --mail-type=END
#SBATCH  --mail-user=ericacnr@colostate.edu
##################
#echo commands to stdout

path="/projects/ericacnr@colostate.edu/ngsRelate/"
chr="CM026847.1"

$path/ngsRelate -h BANS.${chr}.vcf.gz -T GT -c 1 -O vcf.${chr}.res
```

```{bash}
rsync -avzP ericacnr@colostate.edu@login.rc.colorado.edu:/scratch/alpine/ericacnr@colostate.edu/BANS/02.5.neutral_pop_struc/vcf.CM026847.1-2.res /Users/ericarobertson/Desktop/BANS_Pub/analysis/03.delineate_ESUs/scratch/ngsRelate?
```

```{r}
res <- fread("../scratch/ngsRelate/vcf.CM026847.1-2.res")

colnames(res)

range(res$rab)
range(res$KING)
```

Checking for which indv I think are related.
```{r}
library(plotly)
rel <- ggplot(pca.cluster, aes(
    x = PC1, 
    y = PC2, 
    col = Group,
    text = paste0(
        "ID: ", ind, "<br>",
        "Group: ", Group, "<br>",
        "PC1: ", round(PC1, 4), "<br>",
        "PC2: ", round(PC2, 4)
    )
)) +
  geom_point(size = 3) +
  scale_color_manual(values = bg) +
  theme_light() +
  xlab(paste0("PC1 (", signif(pve05$pve[1], 3), "%)")) +
  ylab(paste0("PC2 (", signif(pve05$pve[2], 3), "%)"))

ggplotly(rel)
```
Most of Cluster16...
25N03069 - 25N03072
25N03076
25N03078 - 25N03068
25N03074 - 25N03070

Cluster 20 
25N17432
25N17424 - 25N17421

Cluster 21
25N19612

Checking relatedness values for those.. three options for relatedness metrics:
res$rab
res$theta
res$KING

Rules of thumb for diploids (unadmixed, reasonably good data):
Unrelated: θ ~ 0, rab ~ 0
3rd-degree (first cousins): θ ~ 0.0625, rab ~ 0.125
2nd-degree (half-sibs / avuncular / grandparent): θ ~ 0.125, rab ~ 0.25
1st-degree (parent–offspring / full sib): θ ~ 0.25, rab ~ 0.5
```{r}
hist(res$rab)
#top rab pairs
top.rab <- res %>% filter(rab >=0.125)
nrow(top.rab)
top.rab

unique(top.rab$ida)
unique(top.rab$idb)

samp_list <- read_csv("../scratch/BANS_allind_sample_list.csv") %>% rename(IID=BGP_ID)

samp_unrel <- samp_list %>% filter(!IID %in% unique(top.rab$ida))# %>% filter(!IID %in% unique(top.rab$idb))
nrow(samp_unrel)

samp_unrel %>% write_delim("../scratch/BANS_unrel_sample_list_short.txt")
```

If we remove those individuals, does it improve the PCA?

```{bash}
conda activate bioinf
vcf="/scratch/alpine/ericacnr@colostate.edu/BANS/02.imputation/BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.imputed4.1.final.vcf.gz"

bcftools view -S BANS_unrel_sample_list_short.txt -Oz -o BANS.filtered.unrel.vcf.gz "$vcf"
bcftools index -t BANS.filtered.unrel.vcf.gz
```

Also going to try using PCANGSD for the PCA because apparently they often look better.
```{bash, label="installing pcangst to projects"}
conda create -n pcangsd python=3.10 -y
conda activate pcangsd

mamba install -c conda-forge cython numpy scipy pandas h5py -y

git clone https://github.com/Rosemeis/pcangsd.git
cd pcangsd/
python setup.py build_ext --inplace
pip install --user -r requirements.txt

```

```{bash, label="making input bedfile"}
conda activate GWAS2
vcf="BANS.filtered.unrel.vcf.gz"
plink --vcf $vcf --aec --recode --out BANS.filtered.unrel
#3123808 variants and 201 people pass filters and QC.

cat BANS.filtered.unrel.map | awk -F"\t" '{split($1, a, "__"); print $1"\t"a[1]"."$4"\t"$3"\t"$4}' > BANS.filtered.unrel_loc.map

plink --ped BANS.filtered.unrel.ped \
--map BANS.filtered.unrel_loc.map \
--make-bed --aec \
--out BANS.filtered.unrel

bfile="BANS.filtered.unrel"

plink --bfile $bfile --aec \
--indep-pairwise 25 10 0.5 --out BANS.filtered.unrel
#278863 of 3123808 variants removed.

plink --bfile $bfile --aec --extract BANS.filtered.unrel.prune.in \
  --make-bed --out BANS.filtered.unrel_pruned
#2844945 variants and 201 people pass filters and QC.

plink --bfile BANS.filtered.unrel_pruned --aec --allow-no-sex --pca 189 \
  --out BANS.filtered.unrel.ld25-10-0.5
#2844945 variants and 201 people pass filters and QC.
```

```{bash}
#!/bin/bash
#
#SBATCH --job-name=pcadmix
#SBATCH --output=pcadmix.%j.out
#SBATCH --error=pcadmix.%j.err
#SBATCH -t 5:00:00
#SBATCH --partition=amilan
#SBATCH --qos=normal
#SBATCH --nodes=1
#SBATCH --ntasks-per-node 11
#SBATCH --mem=40G
#SBATCH --mail-type=BEGIN,END,FAIL
#SBATCH  --mail-user=ericacnr@colostate.edu
##################
#echo commands to stdout

set -x

source ~/.bashrc
conda activate pcangsd

pcangsd_path="/projects/ericacnr@colostate.edu/pcangsd"

export PYTHONPATH="/projects/ericacnr@colostate.edu/pcangsd:$PYTHONPATH"

for eig in {1..5}
do
  for r in {1..5}
         do
          $CONDA_PREFIX/bin/python $pcangsd_path/pcangsd/main.py -p BANS.filtered.unrel_pruned -t 4 --admix --admix-seed $RANDOM -e $eig --admix-alpha 10000 -o BANS.filtered.unrel_pruned.alpha10K.e"$eig".run"$r"
        done
done
```

```{bash}
rsync -avzP ericacnr@colostate.edu@login.rc.colorado.edu:/scratch/alpine/ericacnr@colostate.edu/BANS/02.5.neutral_pop_struc/BANS.filtered.unrel.ld25-10-0.5.eigen* /Users/ericarobertson/Desktop/BANS_adaptive_units/analysis/02.5.delineate_ESUs/
```

```{r}
plink_pca <- read.table("../scratch/PCA/BANS.filtered.unrel.ld25-10-0.5.eigenvec", header = FALSE) %>% dplyr::select(-V2)
eigenval <- scan("../scratch/PCA/BANS.filtered.unrel.ld25-10-0.5.eigenval")

pve05 <- data.frame(PC = 1:189, pve = eigenval/sum(eigenval)*100)
ggplot(pve05[1:20, ], aes(x = PC, y = pve)) +
  geom_col(width = 0.9) +
  labs(x = "PC", y = "PVE (%)") +
  theme_bw()

# set names
pca05 <- plink_pca
colnames(pca05)[-1] <- paste0("PC", 1:(ncol(pca05) - 1))
names(pca05)[1] <- "ind"
nrow(pca05)
#correct number of samples

bg <- readRDS("../figures/colors/bg_colors.rds")

pops <- read.csv("../scratch/pops.csv") %>% filter(BGP_ID %in% pca05$ind)

# Join PCA data
pca.cluster <- left_join(pca05, pops %>% rename(ind=BGP_ID), by="ind")  # left_join PCA first to preserve PC columns

samp_unrel  <-  fread("../scratch/BANS_unrel_sample_list_short.txt")
pca.cluster.norel <- pca.cluster %>% filter(ind %in% samp_unrel$IID)
nrow(pca.cluster.norel)

geo_key <- readRDS("../scratch/geo_informative_labels.RDS")
geo_letters <- read.csv("../scratch/cluster_letter_key_geo_lonlat.csv")
letter_order <- sort(unique(pca.cluster.norel$cluster_letter))

pca.cluster.norel <- left_join(pca.cluster.norel, geo_key, by="Group") %>% left_join(geo_letters, by="Group")

rel.nr <- ggplot(pca.cluster.norel, aes(
    x = PC1, 
    y = PC2, 
    col = Group,
    text = paste0(
        "ID: ", ind, "<br>",
        "Location: ", cluster_letter, "<br>",
        "PC1: ", round(PC1, 4), "<br>",
        "PC2: ", round(PC2, 4)
    )
)) +
  geom_point(size = 3) +
  scale_color_manual(
  values = bg,
  breaks = letter_order,
  labels = setNames(pca.cluster.norel$cluster_letter, pca.cluster.norel$Group)
)+
  theme_light() +
  xlab(paste0("PC1 (", signif(pve05$pve[1], 3), "%)")) +
  ylab(paste0("PC2 (", signif(pve05$pve[2], 3), "%)"))

ggplotly(rel.nr)
rel.nr

ggsave("../figures/pca_plots/BANS_PCA_norel_regionlabels_ordered.png", width = 8.5, height = 7)
```

```{r}
library(dplyr)
library(ggplot2)

# key: one row per Group with its cluster letter
key <- pca.cluster.norel %>%
  distinct(Group, cluster_letter)

# order Groups by alphabetic cluster letter
group_breaks <- key %>%
  arrange(cluster_letter) %>%
  pull(Group)

# labels: named by Group, values are cluster letters
group_labels <- setNames(key$cluster_letter, key$Group)

rel.nr <- ggplot(pca.cluster.norel, aes(
  x = PC1,
  y = PC2,
  col = Group,
  text = paste0(
    "ID: ", ind, "<br>",
    "Location: ", cluster_letter, "<br>",
    "PC1: ", round(PC1, 4), "<br>",
    "PC2: ", round(PC2, 4)
  )
)) +
  geom_point(size = 3) +
  scale_color_manual(
    values = bg,
    breaks = group_breaks,
    labels = group_labels,
    name = "Cluster"   # optional legend title
  ) +
  theme_light() +
  xlab(paste0("PC1 (", signif(pve05$pve[1], 3), "%)")) +
  ylab(paste0("PC2 (", signif(pve05$pve[2], 3), "%)"))

rel.nr
ggsave("../figures/pca/BANS_PCA_norel_letter_labels_ordered.png", width = 8.5, height = 7)

```

Yay, that looks much cleaner. Let's see if ADMIX likes it.


# ADMIXTURE
## old way
ADMIXTURE does not accept chromosome names that are not human chromosomes. We will thus just exchange the first column by 0

```{bash}
awk '{$1="0";print $0}' BANS.filtered.unrel_pruned.bim > BANS.filtered.unrel_pruned.bim.tmp
mv BANS.filtered.unrel_pruned.bim.tmp BANS.filtered.unrel_pruned.bim
```

Running admixture with k=1-3

```{bash, label="bash"}
#!/bin/bash
#
#SBATCH --job-name=admix
#SBATCH --output=admix.%j.out
#SBATCH --error=admix.%j.err
#################
#SBATCH -t 2:00:00
#SBATCH -p amilan
#SBATCH --qos=normal
#SBATCH --nodes=1
#SBATCH --ntasks-per-node 5
#SBATCH --mem=12G
#################
#SBATCH --mail-type=END
#SBATCH  --mail-user=ericacnr@colostate.edu
##################
#echo commands to stdout

source ~/.bashrc

num=$1
conda activate GWAS2
admixture --cv BANS.filtered.unrel_pruned.bed $num > BANS.filtered.unrel_pruned.log"$num".out
```

```{bash}
rsync -avzP ericacnr@colostate.edu@login.rc.colorado.edu:/scratch/alpine/ericacnr@colostate.edu/BANS/02.5.neutral_pop_struc/BANS.filtered.unrel_pruned*.Q /Users/ericarobertson/Desktop/BANS_adaptive_units/analysis/02.5.delineate_ESUs
```

To identify the best value of k clusters which is the value with lowest cross-validation error, we need to collect the cv errors. 

```{bash}
awk '/CV/ {print $3,$4}' BANS.filtered.unrel_pruned*out | cut -c 4,7-20
```
1 0.51944
2 0.53064
3 0.55535

So, this is suggesting that there isn't a strong signal of population structure an k=1 is the best supported. This is reasonable considering the strong IBD signal. So, let's plot the admix but also do some feems to see if there are strong barrier to migration present that might help us delineate stuff.

```{r}
samplelist <- pca05$ind

all_data <- tibble(sample=character(),
                   k=numeric(),
                   Q=character(),
                   value=numeric())

for (k in 1:4){
    data <- read_delim(paste0("admix_output/BANS.filtered.unrel_pruned.",k,".Q"),
                       col_names = paste0("Q",seq(1:k)),
                       delim=" ")
    data$sample <- samplelist
    data$k <- k
    
    #This step converts from wide to long.
    data %>% gather(Q, value, -sample,-k) -> data
    all_data <- rbind(all_data,data)
}

head(all_data)

ADMIX_plot <- all_data %>%
  filter(k %in% 1:4) %>%
  ggplot(aes(x = sample, y = value, fill = factor(Q))) +
  geom_bar(stat = "identity", width = 1) +
  facet_wrap(~ k, ncol = 1, scales = "free_y") +
  xlab("Sample") +
  ylab("Ancestry proportion") +
  scale_fill_brewer(palette = "Set1", name = "Cluster") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1),
    panel.spacing = unit(0.3, "lines")
  )
ggsave("plots/final_ESU/BANS_ADMIX_k1-4.png",width = 10)
ADMIX_plot
```

```{r}
all_k <- all_data %>%
  ggplot(aes(x = sample, y = value, fill = factor(Q))) + 
  geom_bar(stat = "identity", position = "stack") +
  xlab("Sample") + ylab("Ancestry") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 5)) +  # Adjust size here
  scale_fill_brewer(palette = "Set1", name = "K", labels = seq(1:5)) +
  facet_wrap(~k, ncol = 1)

ggplotly(all_k)
```

## pcangsd admix
for eig in {1..5}
do
  for r in {1..5}
         do
          $CONDA_PREFIX/bin/python $pcangsd_path/pcangsd/main.py -p BANS.filtered.unrel_pruned -t 4 --admix --admix-seed $RANDOM -e $eig --admix-alpha 10 -o BANS.filtered.unrel_pruned.alpha10.e"$eig".run"$r"
        done
done
```{bash}
rsync -avzP ericacnr@colostate.edu@login.rc.colorado.edu:/scratch/alpine/ericacnr@colostate.edu/BANS/02.5.neutral_pop_struc/BANS.filtered.unrel_pruned.alpha0.*extratest*.Q /Users/ericarobertson/Desktop/BANS_adaptive_units/analysis/02.5.delineate_ESUs/intermediates/admix_output/pcadmix_output/
```

```{r, label="plotting_with_pophelper", message=FALSE, warning=FALSE}
library(pophelper)
library(data.table)
library(dplyr)
library(stringr)
library(gridExtra)
library(grid)

# Inputs
q_dir   <- "pcadmix_output"
q_pat   <- "^BANS.filtered.unrel_pruned.alpha5.*\\.Q$"
famfile <- "BANS.filtered.unrel.fam"
metafile <- "../03.GEA/results/BANS_topRDA_vars1mil.csv"

# cluster colors (use bg if it exists; otherwise define a fallback)
bg <- c(
  "#E69F00", # orange
  "#56B4E9", # sky blue
  "#009E73", # bluish green
  "#F0E442", # yellow
  "#0072B2", # blue
  "#D55E00", # vermillion
  "#CC79A7", # reddish purple
  "#999999", # gray
  "#44AA99", # teal
  "#AA4499"  # magenta
)

region_order <- c(
  "Cluster_1","Cluster_5","Cluster_16","Cluster_4","Cluster_9","Cluster_12","Cluster_2","Cluster_20",
  "Cluster_3","Cluster_8","Cluster_17","Cluster_21","Cluster_18","Cluster_19",
  "Cluster_6","Cluster_14","Cluster_11","Cluster_10","Cluster_13","Cluster_15","Cluster_7"
)


# Read Q files + add sample IDs
sfiles <- list.files(path = q_dir, pattern = q_pat, full.names = TRUE)
stopifnot(length(sfiles) > 0)

slist <- readQ(files = sfiles, filetype = "basic", indlabfromfile = FALSE)

inorder <- fread(famfile, header = FALSE)[, .(IID = V1)]  # your IDs are in V1

# assign rownames across all runs (only if dimensions match)
if (length(unique(sapply(slist, nrow))) == 1 && nrow(slist[[1]]) == nrow(inorder)) {
  slist <- lapply(slist, `rownames<-`, inorder$IID)
} else {
  stop("Row counts differ among Q files and/or don't match the .fam; can't safely assign IID rownames.")
}


# Build Region labels aligned to Q rows
meta <- read.csv(metafile) %>% filter(BGP_ID %in% inorder$IID)

Region_vec <- meta$Pop[match(rownames(slist[[1]]), meta$BGP_ID)]
Region_df <- data.frame(Region = Region_vec, stringsAsFactors = FALSE)
rownames(Region_df) <- rownames(slist[[1]])
Region_df$Region <- factor(Region_df$Region, levels = region_order)
Region_df$Region <- as.character(Region_df$Region)


# Parse K/run from filenames (safer to use sfiles, not names(slist))

Kvec   <- as.integer(str_extract(basename(sfiles), "(?<=admix\\.)\\d+"))
Runvec <- as.integer(str_extract(basename(sfiles), "(?<=run)\\d+"))

# choose which Ks to plot
Ks <- sort(unique(Kvec[!is.na(Kvec)]))
stopifnot(length(Ks) > 0)


# Build plots (use the first run for each K)

g_list <- vector("list", length(Ks))
names(g_list) <- paste0("K", Ks)

for (K in Ks) {
  idx <- which(Kvec == K)
  idx <- idx[order(Runvec[idx])]
  idx1 <- idx[1]  # first run for this K

  # pophelper wants a list (even if single run); slist[idx1] already is a list
  p <- plotQ(
    alignK(slist[idx1]),
    grplab = Region_df,
    showgrplab = TRUE,
    clustercol = unname(bg[seq_len(K)]),
    returnplot = TRUE,
    exportplot = FALSE,
    ordergrp = TRUE,
    grplabsize = 1.5,
    linesize = 0.5,
    pointsize = 2,
    grplabangle = 45,
    grplabheight = 14
  )

  g_list[[paste0("K", K)]] <- p$plot[[1]]
}


# Save multi-panel PDF (2 cols; auto height)


pdf("plots/final_ESU/BANS_pcadmix_filtered.unrel.pruned_alpha5_K2-6_RDAregions_admixture.pdf",
    width = 10, height = 2.7 * nrow, useDingbats = FALSE)
plots_to_show <- g_list
grid.arrange(
  grobs = plots_to_show,
  layout_matrix = matrix(seq_along(plots_to_show), byrow = FALSE)
)
dev.off()
```

## color coding this with letters:
```{r}
library(pophelper)
library(data.table)
library(dplyr)
library(stringr)
library(readr)
library(tidyr)
library(ggplot2)
library(gridExtra)
library(grid)

# Inputs
q_dir   <- "../intermediates/admix_output/pcadmix_output"
q_pat <- "^BANS\\.filtered\\.unrel_pruned\\.alpha0.*extratest.*\\.Q$"
famfile <- "../intermediates/BANS.filtered.unrel.fam"

popsfile   <- "../intermediates/pops.csv"             # BGP_ID, Group
letterfile <- "../intermediates/cluster_letter_key_geo_lonlat.csv"           # Group/Pop, cluster_letter

out_pdf    <- "../figures/admix/BANS_pcadmix_overall_alpha0_allK_letters_extratest_GGPLOT.pdf"

# Read Q files + add sample IDs

sfiles <- list.files(path = q_dir, pattern = q_pat, full.names = TRUE)
stopifnot(length(sfiles) > 0)

slist <- readQ(files = sfiles, filetype = "basic", indlabfromfile = FALSE)

inorder <- fread(famfile, header = FALSE)[, .(IID = V1)]
stopifnot(length(unique(sapply(slist, nrow))) == 1)
stopifnot(nrow(slist[[1]]) == nrow(inorder))

slist <- lapply(slist, `rownames<-`, inorder$IID)


# Build cluster_letter per individual (aligned to Q rownames)

pops <- read_csv(popsfile, show_col_types = FALSE) %>%
  transmute(BGP_ID, Group)

letters <- read_csv(letterfile, show_col_types = FALSE)

if ("Pop" %in% names(letters) && !"Group" %in% names(letters)) {
  letters <- letters %>% rename(Group = Pop)
}
stopifnot(all(c("Group", "cluster_letter") %in% names(letters)))

pops.cluster.letter <- pops %>%
  left_join(letters %>% select(Group, cluster_letter), by = "Group")

# vector of letters in the same order as Q rows
Letter_vec <- pops.cluster.letter$cluster_letter[
  match(rownames(slist[[1]]), pops.cluster.letter$BGP_ID)
]

# letter order (stable)
letter_levels <- letters %>%
  distinct(cluster_letter) %>%
  pull(cluster_letter) %>%
  trimws()


# Parse K/run from filenames

Kvec   <- as.integer(str_extract(basename(sfiles), "(?<=admix\\.)\\d+"))
Runvec <- as.integer(str_extract(basename(sfiles), "(?<=run)\\d+"))

Ks <- sort(unique(Kvec[!is.na(Kvec)]))
stopifnot(length(Ks) > 0)


# Component color helper (per K)

get_comp_cols <- function(K) {

  # fixed 10-color palette
  base_cols <- c(
    "#E69F00", # orange
    "#56B4E9", # sky blue
    "#009E73", # bluish green
    "#F0E442", # yellow
    "#0072B2", # blue
    "#D55E00", # vermillion
    "#CC79A7", # reddish purple
    "#999999", # gray
    "#44AA99", # teal
    "#AA4499"  # magenta
  )


  # if K > 10, interpolate additional colors
  if (K <= length(base_cols)) {
    cols <- base_cols[1:K]
  } else {
    cols <- colorRampPalette(base_cols)(K)
  }

  names(cols) <- paste0("ESU", seq_len(K))
  cols
}

# Build ggplots + assignments

g_list <- vector("list", length(Ks))
names(g_list) <- paste0("K", Ks)

assign_list <- vector("list", length(Ks))
names(assign_list) <- paste0("K", Ks)

for (K in Ks) {

  # pick the first run for this K
  idx <- which(Kvec == K)
  idx <- idx[order(Runvec[idx])]
  idx1 <- idx[1]

  # wide Q (rows = individuals)
  Qwide <- as.data.frame(slist[[idx1]])
  stopifnot(ncol(Qwide) == K)

  colnames(Qwide) <- paste0("ESU", seq_len(K))
  Qwide$ind <- rownames(Qwide)

  # add cluster_letter + (optional) Group
  meta_ind <- tibble(
    ind = rownames(slist[[idx1]]),
    cluster_letter = trimws(as.character(Letter_vec))
  ) %>%
    mutate(cluster_letter = factor(cluster_letter, levels = letter_levels))

  # long for ggplot
  q_long <- Qwide %>%
    left_join(meta_ind, by = "ind") %>%
    filter(!is.na(cluster_letter)) %>%
    pivot_longer(cols = starts_with("ESU"),
                 names_to = "popGroup",
                 values_to = "prob") %>%
    mutate(
      popGroup = factor(popGroup, levels = paste0("ESU", seq_len(K)))
    )

  # ggplot for this K
  comp_cols <- get_comp_cols(K)

  g_list[[paste0("K", K)]] <-
    ggplot(q_long, aes(x = ind, y = prob, fill = popGroup)) +
    geom_col(color = "gray", linewidth = 0.1) +
    facet_grid(~cluster_letter, switch = "x", scales = "free_x", space = "free_x") +
    theme_minimal() +
    labs(x = "", y = "Ancestry", title = paste0("K = ", K)) +
    scale_y_continuous(expand = c(0, 0)) +
    scale_x_discrete(expand = expansion(add = 1)) +
    theme(
      panel.spacing.x = unit(0.001, "lines"),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      strip.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5, size = 8),
      panel.grid = element_blank()
    ) +
    scale_fill_manual(values = comp_cols, guide = "none")

}

pdf(out_pdf, width = 10, height = 2.2 * length(g_list), useDingbats = FALSE)
grid.arrange(
  grobs = g_list,
  layout_matrix = matrix(seq_along(g_list), ncol = 1, byrow = FALSE)
)
dev.off()

message("Saved PDF: ", out_pdf)
message("Saved assignments: ", out_assign)
```
#### double checking convergence
```{r}
library(tidyverse)

#log-likelihoods pulled from the output file for alpha=0: awk '/Log-likelihood:/ {print $NF}' pcadmix.22816373.out

# awk '
# /Log-likelihood:/ {
#   vals[n++] = $NF
# }
# END {
#   printf("loglik_vals <- c(\n  ")
#   for (i = 0; i < n; i++) {
#     printf("%s", vals[i])
#     if (i < n-1) {
#       if ((i+1) % 5 == 0)
#         printf(",\n  ")
#       else
#         printf(", ")
#     }
#   }
#   printf("\n)\n")
# }
# ' pcadmix.23335757.out

#pcadmix.23335757.out
loglik_vals <- c(
  -317379697.37537, -317379813.54059, -317379678.38043, -317379760.91672, -317379713.25396,
  -314899814.78644, -314899716.79290, -314899793.68513, -314899574.60226, -314899754.03714,
  -312851042.25835, -312850828.66974, -312851082.17316, -312851247.93471, -312850893.01092,
  -310629491.65431, -310629568.51323, -310629372.21268, -310629284.03814, -310629277.99590,
  -308942526.42675, -308942091.66666, -308942166.14086, -308942631.48534, -308942846.93230
)

Ks
runs_per_k <- 5

ll_df <- tibble(
  K = rep(Ks, each = runs_per_k),
  run = rep(1:runs_per_k, times = length(Ks)),
  logLik = loglik_vals
)

ll_sum <- ll_df %>%
  group_by(K) %>%
  summarise(
    meanLL = mean(logLik),
    sdLL   = sd(logLik),
    .groups = "drop"
  )

p_ll <- ggplot(ll_sum, aes(x = K, y = meanLL)) +
  geom_line() +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = meanLL - sdLL, ymax = meanLL + sdLL), width = 0.15) +
  labs(x = "K", y = "Mean log-likelihood (± SD across runs)") +
  theme_bw()

p_ll

p_ll_runs <- ggplot(ll_df, aes(x = K, y = logLik, group = run)) +
  geom_point(alpha = 0.8) +
  geom_line(alpha = 0.4) +
  theme_bw() +
  labs(x = "K", y = "Log-likelihood (each run)")

p_ll_runs
```