---
title: "03.a.initial_popstructure"
output: html_document
---

# Libraries
```{r}
library(tidyverse)
library(data.table)
library(dplyr)
library(rnaturalearth)
library(geosphere)
library(ggplot2)
library(rnaturalearth)
library(raster)
library(terra)
```

# Pruning
```{bash}
#GIVING SNP IDS
vcf="/scratch/alpine/ericacnr@colostate.edu/BANS/02.imputation/BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.rm_scaf_lth10.imputed4.1.vcf.gz"

conda activate GWAS2
plink --vcf $vcf --aec --recode --out BANS.all.recode

cat BANS.all.recode.map | awk -F"\t" '{split($1, a, "__"); print $1"\t"a[1]"."$4"\t"$3"\t"$4}' > BANS.all.recode_loc.map

plink --ped BANS.all.recode.ped \
--map BANS.all.recode_loc.map \
--make-bed --aec \
--out BANS.all.recode

bfile="BANS.all.recode"

plink --bfile $bfile --aec \
--indep-pairwise 25 10 0.5 --out PC1_allindv
#602317 of 5219142 variants removed

plink --bfile $bfile --aec --extract PC1_allindv.prune.in \
  --make-bed --out pruned_allindv
#4616825 variants and 206 people pass filters and QC.

plink --bfile pruned_allindv --aec --allow-no-sex --pca 205 \
  --out PC1_allindv.ld25-10-0.5
```

```{bash}
rsync -avzP ericacnr@colostate.edu@login.rc.colorado.edu:/scratch/alpine/ericacnr@colostate.edu/BANS/02.5.neutral_pop_struc/PC1_allindv.ld25-10-0.5*eigen* /Users/ericarobertson/Desktop/BANS_Pub/analysis/03.delineate_ESUs/scratch/PCA/
```
# Time to plot the PCA.
## grouping pops
The actual group pops is missing indv that I filtered for missingness, so this is a quick temporary one...
```{r}
library(dplyr)
library(readr)
library(geosphere)
library(leaflet)
library(htmlwidgets)
library(viridisLite)

meta <- read.csv("../../../raw_data/BANS_all_sample_data_shifted.csv")

coords <- meta %>%
  dplyr::select(Long, Lat)

dist_matrix <- distm(as.matrix(coords), fun = distHaversine)

libraries <- read.delim("../../01.fastq_processing/snakemake/BANS.units.tsv") %>%
  rename(BGP_ID = sample)

lib.meta <- left_join(
  meta,
  libraries %>% dplyr::select(library, BGP_ID),
  by = "BGP_ID"
)

hclust_groups <- hclust(as.dist(dist_matrix), method = "complete")

meta$GroupID <- cutree(hclust_groups, h = 50000)

meta.sub <- meta %>%
  mutate(Pop = paste0("Cluster_", GroupID))

pops <- data.frame(
  BGP_ID = meta.sub$BGP_ID,
  Group = meta.sub$Pop,
  Lat = meta.sub$Lat,
  Long = meta.sub$Long
)

pops %>% write_csv("../scratch/pops.csv")

# cluster colors
clusters <- sort(unique(meta.sub$Pop))

cluster_pal <- colorFactor(
  palette = viridisLite::turbo(length(clusters)),
  domain = clusters
)

# interactive map
cluster_map <- leaflet(meta.sub) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(
    lng = ~Long,
    lat = ~Lat,
    color = ~cluster_pal(Pop),
    fillColor = ~cluster_pal(Pop),
    fillOpacity = 0.8,
    opacity = 1,
    radius = 5,
    stroke = TRUE,
    weight = 1,
    popup = ~paste0(
      "<b>BGP_ID:</b> ", BGP_ID, "<br>",
      "<b>Cluster:</b> ", Pop, "<br>",
      "<b>Lat:</b> ", Lat, "<br>",
      "<b>Long:</b> ", Long
    ),
    label = ~Pop
  ) %>%
  addLegend(
    position = "bottomright",
    pal = cluster_pal,
    values = ~Pop,
    title = "Spatial clusters",
    opacity = 1
  )

cluster_map

######## SAVING PLOT
n_clusters <- length(unique(meta.sub$Pop))
cluster_colors <- rainbow(n_clusters)
names(cluster_colors) <- unique(meta.sub$Pop)

meta.b <- meta.sub %>%
  group_by(Pop) %>%
  mutate(cluster_size = n()) %>%
  ungroup()

# plot it
p <- ggplot() +
  geom_sf(data = map, fill = "white", color = "black", size = 0.1) +
  geom_point(data = meta.b, aes(x = Long, y = Lat, color = Pop),
             size = 1.5, position = position_jitter(width = 0.1, height = 0.1)) +
  geom_text(data = meta.b,
            aes(x = Long, y = Lat, label = Pop),
            size = 2, vjust = -0.6) +
  scale_colour_manual(values = cluster_colors, name = "Spatial Clusters", guide = "none") +
  coord_sf(xlim = c(-145, -55), ylim = c(15, 80)) + 
  theme_minimal() +
  labs(title = "All clusters with ≥4 individuals",
       x = "Longitude",
       y = "Latitude")
p
ggsave("../outputs/random_color_cluster_plot.png")
```

## plotting PCA

Useful place names
```{r}
group_key <- tribble(
  ~Group,       ~Group_label,
  "Cluster_1",  "AK",
  "Cluster_2",  "WA",
  "Cluster_3",  "SK",
  "Cluster_4",  "BC_1",
  "Cluster_5",  "YT",
  "Cluster_6",  "PE",
  "Cluster_7",  "MB_1",
  "Cluster_8",  "AB",
  "Cluster_9",  "NB",
  "Cluster_10", "NS_1",
  "Cluster_11", "NS_2",
  "Cluster_12", "ON",
  "Cluster_13", "BC_2",
  "Cluster_14", "QC_1",
  "Cluster_15", "QC_2",
  "Cluster_16", "NL",
  "Cluster_17", "NT",
  "Cluster_18", "MB_2",
  "Cluster_19", "WY_1",
  "Cluster_20", "WY_2",
  "Cluster_21", "MT",
  "Cluster_22", "ND"
)

saveRDS(group_key, "../scratch/geo_informative_labels.RDS")

library(sf)

cluster.sf <- left_join(cluster_geo, pops, by = "Group") %>%
  st_as_sf(coords = c("Long", "Lat"), crs = 4326)

cluster.centroids <- cluster.sf %>%
  group_by(Group, Group_label) %>%
  summarise() %>%
  st_centroid()

cluster.centroids <- cluster.centroids %>%
  cbind(st_coordinates(.)) %>% dplyr::select(-geometry)

saveRDS(cluster.centroids, "../scratch/geo_informative_labels_centroids.RDS")
```

### color scheme
Quick, somewhat neutral color scheme based on geography alone...
```{r}
# Define clusters by region
West_clusters    <- c("Cluster_1", "Cluster_5", "Cluster_17", "Cluster_4", "Cluster_8", "Cluster_13", "Cluster_2", "Cluster_21")
Central_clusters <- c("Cluster_3", "Cluster_7", "Cluster_18", "Cluster_19", "Cluster_22", "Cluster_20")
East_clusters    <- c("Cluster_12", "Cluster_15", "Cluster_14", "Cluster_10", "Cluster_9", "Cluster_11", "Cluster_6", "Cluster_16")

# Assign colors by cluster
library(RColorBrewer)
W.cols <- setNames(brewer.pal(n = length(West_clusters), "Reds"), West_clusters)
C.cols <- setNames(brewer.pal(n = length(Central_clusters), "Blues"), Central_clusters)
E.cols <- setNames(brewer.pal(n = length(East_clusters), "Greens"), East_clusters)
bg <- c(W.cols, C.cols, E.cols)
saveRDS(bg, file = "../figures/colors/bg_colors_new.rds")
bg <- readRDS("../figures/colors/bg_colors_new.rds")
```
### plotting
```{r}
#just saving a sample list real quick...
read.table("../scratch/PCA/PC1_allindv.ld25-10-0.5.eigenvec", header = FALSE) %>% dplyr::select(V1) %>% rename(BGP_ID=V1) %>% write_csv("../scratch/BANS_allind_sample_list.csv")

plink_pca <- read.table("../scratch/PCA/PC1_allindv.ld25-10-0.5.eigenvec", header = FALSE) %>% dplyr::select(-V2)

eigenval <- scan("../scratch/PCA/PC1_allindv.ld25-10-0.5.eigenval")

pve05 <- data.frame(PC = 1:205, pve = eigenval/sum(eigenval)*100)

# set names
pca05 <- plink_pca
colnames(pca05)[-1] <- paste0("PC", 1:(ncol(pca05) - 1))
names(pca05)[1] <- "ind"
nrow(pca05)
#correct number of samples

pops <- read.csv("../scratch/pops.csv")

# Join PCA data
pca.cluster <- left_join(pca05, pops %>% rename(ind=BGP_ID), by="ind")  # left_join PCA first to preserve PC columns

# Factor for cluster order
region_order <- c(
  "Cluster_1", "Cluster_5", "Cluster_17", "Cluster_4", "Cluster_8", "Cluster_13", "Cluster_2", "Cluster_21",
  "Cluster_3", "Cluster_7", "Cluster_18", "Cluster_19", "Cluster_22", "Cluster_20",
  "Cluster_12", "Cluster_15", "Cluster_14", "Cluster_10", "Cluster_9", "Cluster_11", "Cluster_6", "Cluster_16"
)
pca.cluster$Cluster <- factor(pca.cluster$Group, levels = region_order)

# Plot
b05 <- ggplot(pca.cluster, aes(PC1, PC2, col = Cluster)) +
  geom_point(size = 3) +
  scale_color_manual(values = bg) +
  theme_light() +
  xlab(paste0("PC1 (", signif(pve05$pve[1], 3), "%)")) +
  ylab(paste0("PC2 (", signif(pve05$pve[2], 3), "%)"))

b05
ggsave("../figures/pca_plots/BANS_allind_overall_PCA_2.png", width=8)
```

```{r}
meta.b.plot <- meta.b %>%
  rename(Group = Pop) %>%   # only needed if meta.b uses Pop instead of Group
  left_join(group_key, by = "Group")

label_df <- meta.b.plot %>%
  group_by(Group, Group_label) %>%
  summarise(
    Long = mean(Long),
    Lat = mean(Lat),
    .groups = "drop"
  )

bg <- readRDS("../figures/colors/bg_colors_new.rds")
p <- ggplot() +
  geom_sf(data = map, fill = "grey95", color = "black", linewidth = 0.1) +
  geom_point(
  data = meta.b.plot,
  aes(x = Long, y = Lat, fill = Group),
  shape = 21,
  color = "black",
  stroke = 0.3,
  size = 2,
  position = position_jitter(width = 0.1, height = 0.1)
) +
scale_fill_manual(values = bg, guide = "none")+
  ggrepel::geom_label_repel(
    data = label_df,
    aes(x = Long, y = Lat, label = Group_label),
    nudge_y=1.5,
    size=2,
    label.size=.15
  ) +
  scale_colour_manual(values = bg, name = "Spatial Clusters", guide = "none") +
  coord_sf(xlim = c(-160, -55), ylim = c(15, 80)) +
  theme_minimal() +
  labs(
    title = "All clusters with ≥4 individuals",
    x = "Longitude",
    y = "Latitude"
  )

p

ggsave("../figures/BANS_allind_cluster_map_bg.png", plot = p, width = 10, height = 7, dpi = 300)
```

# Remove Related Individuals
## NGSrelate
Using NGSrelate
```{bash, label="ngsRelate install"}
conda activate bioinf
#stuff I need to make the libraries install correctly
mamba install -c conda-forge bzip2 zlib xz libcurl

git clone --recursive https://github.com/SAMtools/htslib
#this misses a necessary thing libhts.a so gotta do it again
cd /projects/ericacnr@colostate.edu/htslib
make clean

# make sure it uses conda include/lib
make -j2 libhts.a \
  CPPFLAGS="-I$CONDA_PREFIX/include" \
  LDFLAGS="-L$CONDA_PREFIX/lib" \
  LIBS="-lz -lbz2 -llzma -lcurl"
  
# get ngsRelate
git clone https://github.com/ANGSD/ngsRelate
cd htslib/;make -j2;cd ../ngsRelate;make HTSSRC=../htslib/
```

Running on just the one big chromosome.
```{bash}
conda activate bioinf

vcf="/scratch/alpine/ericacnr@colostate.edu/BANS/02.imputation/BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.rm_scaf_lth10.imputed4.1.vcf.gz"

chr="CM026847.1"

bcftools view -r $chr -Oz -o BANS.${chr}.vcf.gz $vcf

bcftools index -t BANS.${chr}.vcf.gz
```

```{bash, label="ngs_relate.sbatch"}
#!/bin/bash
#
#SBATCH --job-name=relate
#SBATCH --output=relate.%j.out
#SBATCH --error=relate.%j.err
#################
#SBATCH -t 2:00:00
#SBATCH -p amilan
#SBATCH --qos=normal
#SBATCH --nodes=1
#SBATCH --ntasks-per-node 5
#SBATCH --mem=12G
#################
#SBATCH --mail-type=END
#SBATCH  --mail-user=ericacnr@colostate.edu
##################
#echo commands to stdout

path="/projects/ericacnr@colostate.edu/ngsRelate/"
chr="CM026847.1"

$path/ngsRelate -h BANS.${chr}.vcf.gz -T GT -c 1 -O vcf.${chr}.res
```

```{bash}
rsync -avzP ericacnr@colostate.edu@login.rc.colorado.edu:/scratch/alpine/ericacnr@colostate.edu/BANS/02.5.neutral_pop_struc/vcf.CM026847.1.res /Users/ericarobertson/Desktop/BANS_Pub/analysis/03.delineate_ESUs/scratch/ngsRelate
```

```{r}
fam <- fread("../scratch/BANS.all.recode.fam")
res <- fread("../scratch/ngsRelate/vcf.CM026847.1.res")

colnames(res)

range(res$rab)
range(res$KING)

id_lookup <- data.table(
  idx = 0:(nrow(fam) - 1),
  IID = fam$V2
)

# replace a column
res <- merge(
  res,
  id_lookup,
  by.x = "a",
  by.y = "idx",
  all.x = TRUE
)

setnames(res, "IID", "a_ID")

# replace b column
res <- merge(
  res,
  id_lookup,
  by.x = "b",
  by.y = "idx",
  all.x = TRUE
)

setnames(res, "IID", "b_ID")
```

Checking for which indv I think are related.
```{r}
library(plotly)
rel <- ggplot(pca.cluster, aes(
    x = PC1, 
    y = PC2, 
    col = Group,
    text = paste0(
        "ID: ", ind, "<br>",
        "Group: ", Group, "<br>",
        "PC1: ", round(PC1, 4), "<br>",
        "PC2: ", round(PC2, 4)
    )
)) +
  geom_point(size = 3) +
  scale_color_manual(values = bg) +
  theme_light() +
  xlab(paste0("PC1 (", signif(pve05$pve[1], 3), "%)")) +
  ylab(paste0("PC2 (", signif(pve05$pve[2], 3), "%)"))

ggplotly(rel)
```
Checking relatedness values for those.. three options for relatedness metrics:
res$rab
res$theta
res$KING

Rules of thumb for diploids (unadmixed, reasonably good data):
Unrelated: θ ~ 0, rab ~ 0
3rd-degree (first cousins): θ ~ 0.0625, rab ~ 0.125
2nd-degree (half-sibs / avuncular / grandparent): θ ~ 0.125, rab ~ 0.25
1st-degree (parent–offspring / full sib): θ ~ 0.25, rab ~ 0.5
```{r}
hist(res$rab)
#top rab pairs
top.rab <- res %>% filter(rab >=0.125)
nrow(top.rab)
top.rab

unique(top.rab$a_ID)
unique(top.rab$b_ID)

samp_list <- fam %>% rename(IID=V1) %>% select(IID)

samp_unrel <- samp_list %>% filter(!IID %in% unique(top.rab$a_ID))# %>% filter(!IID %in% unique(top.rab$idb))
nrow(samp_unrel)

samp_unrel %>% write_delim("../scratch/BANS_unrel_sample_list_short_2.txt", col_names = F)
```

If we remove those individuals, does it improve the PCA?

```{bash}
conda activate bioinf
vcf="/scratch/alpine/ericacnr@colostate.edu/BANS/02.imputation/BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.rm_scaf_lth10.imputed4.1.vcf.gz"

bcftools view -S BANS_unrel_sample_list.txt -Oz -o BANS.filtered.unrel.vcf.gz "$vcf"
bcftools index -t BANS.filtered.unrel.vcf.gz
```

Also going to try using PCANGSD for the PCA because apparently they often look better.
```{bash, label="installing pcangst to projects"}
conda create -n pcangsd python=3.10 -y
conda activate pcangsd

mamba install -c conda-forge cython numpy scipy pandas h5py -y

git clone https://github.com/Rosemeis/pcangsd.git
cd pcangsd/
python setup.py build_ext --inplace
pip install --user -r requirements.txt

```

```{bash, label="making input bedfile"}
conda activate GWAS2
vcf="BANS.filtered.unrel.vcf.gz"
plink --vcf $vcf --aec --recode --out BANS.filtered.unrel

cat BANS.filtered.unrel.map | awk -F"\t" '{split($1, a, "__"); print $1"\t"a[1]"."$4"\t"$3"\t"$4}' > BANS.filtered.unrel_loc.map

plink --ped BANS.filtered.unrel.ped \
--map BANS.filtered.unrel_loc.map \
--make-bed --aec \
--out BANS.filtered.unrel

bfile="BANS.filtered.unrel"

plink --bfile $bfile --aec \
--indep-pairwise 25 10 0.2 --out BANS.filtered.unrel

plink --bfile $bfile --aec --extract BANS.filtered.unrel.prune.in \
  --make-bed --out BANS.filtered.unrel_pruned

plink --bfile BANS.filtered.unrel_pruned --aec --allow-no-sex --pca 199 \
  --out BANS.filtered.unrel.ld25-10-0.2
```

R2 of 0.5: 4613981 variants and 200 people pass filters and QC.
R2 of 0.2: 3802168 variants and 200 people pass filters and QC.

```{bash, label="pcadmix"}
#!/bin/bash
#
#SBATCH --job-name=pcadmix
#SBATCH --output=pcadmix.%j.out
#SBATCH --error=pcadmix.%j.err
#SBATCH -t 5:00:00
#SBATCH --partition=amilan
#SBATCH --qos=normal
#SBATCH --nodes=1
#SBATCH --ntasks-per-node 11
#SBATCH --mem=40G
#SBATCH --mail-type=BEGIN,END,FAIL
#SBATCH  --mail-user=ericacnr@colostate.edu
##################
#echo commands to stdout

set -x

source ~/.bashrc
conda activate pcangsd

pcangsd_path="/projects/ericacnr@colostate.edu/pcangsd"

export PYTHONPATH="/projects/ericacnr@colostate.edu/pcangsd:$PYTHONPATH"

for eig in {1..5}
do
  for r in {1..5}
         do
          $CONDA_PREFIX/bin/python $pcangsd_path/pcangsd/main.py -p BANS.filtered.unrel_pruned -t 4 --admix --admix-seed $RANDOM -e $eig --admix-alpha 10000 -o BANS.filtered.unrel_pruned0.2.alpha10K.e"$eig".run"$r"
        done
done
```

```{bash}
rsync -avzP ericacnr@colostate.edu@login.rc.colorado.edu:/scratch/alpine/ericacnr@colostate.edu/BANS/02.5.neutral_pop_struc/BANS.filtered.unrel.ld25-10-0.2.eigen* /Users/ericarobertson/Desktop/BANS_Pub/analysis/03.delineate_ESUs/scratch/PCA
```

```{r}
plink_pca <- read.table("../scratch/PCA/BANS.filtered.unrel.ld25-10-0.2.eigenvec", header = FALSE) %>% dplyr::select(-V2)
eigenval <- scan("../scratch/PCA/BANS.filtered.unrel.ld25-10-0.2.eigenval")

pve05 <- data.frame(PC = 1:199, pve = eigenval/sum(eigenval)*100)
ggplot(pve05[1:20, ], aes(x = PC, y = pve)) +
  geom_col(width = 0.9) +
  labs(x = "PC", y = "PVE (%)") +
  theme_bw()

# set names
pca05 <- plink_pca
colnames(pca05)[-1] <- paste0("PC", 1:(ncol(pca05) - 1))
names(pca05)[1] <- "ind"
nrow(pca05)
#correct number of samples

bg <- readRDS("../figures/colors/bg_colors_new.rds")

pops <- read.csv("../scratch/pops.csv") %>% filter(BGP_ID %in% pca05$ind)

# Join PCA data
pca.cluster <- left_join(pca05, pops %>% rename(ind=BGP_ID), by="ind")  # left_join PCA first to preserve PC columns

samp_unrel  <-  fread("../scratch/BANS_unrel_sample_list_short_2.txt", header = F)
pca.cluster.norel <- pca.cluster %>% filter(ind %in% samp_unrel$V1)
nrow(pca.cluster.norel)

geo_key <- readRDS("../scratch/geo_informative_labels.RDS")
region_order <- c(
  "Cluster_1", "Cluster_5", "Cluster_17", "Cluster_4", "Cluster_8", "Cluster_13", "Cluster_2", "Cluster_21",
  "Cluster_3", "Cluster_7", "Cluster_18", "Cluster_19", "Cluster_22", "Cluster_20",
  "Cluster_12", "Cluster_15", "Cluster_14", "Cluster_10", "Cluster_9", "Cluster_11", "Cluster_6", "Cluster_16"
)

pca.cluster.norel <- pca.cluster.norel %>%
  left_join(geo_key, by = "Group") %>%
  mutate(
    Group = factor(Group, levels = region_order)
  )

# make one label per Group, in the same order
group_labels <- pca.cluster.norel %>%
  distinct(Group, Group_label) %>%
  arrange(Group)

label_vec <- setNames(group_labels$Group_label, group_labels$Group)

# make sure colors are in the same order
bg_use <- bg[region_order]

rel.nr <- ggplot(
  pca.cluster.norel,
  aes(
    x = PC1,
    y = PC2,
    color = Group,
    text = paste0(
      "ID: ", ind, "<br>",
      "Location: ", Group_label, "<br>",
      "PC1: ", round(PC1, 4), "<br>",
      "PC2: ", round(PC2, 4)
    )
  )
) +
  geom_point(size = 3) +
  scale_color_manual(
    values = bg_use,
    breaks = region_order,
    labels = label_vec[region_order],
    name = "Location"
  ) +
  theme_light() +
  xlab(paste0("PC1 (", signif(pve05$pve[1], 3), "%)")) +
  ylab(paste0("PC2 (", signif(pve05$pve[2], 3), "%)"))

rel.nr

ggsave("../figures/pca_plots/BANS_PCA_norel_regionlabels_ordered_2.png", width = 8.5, height = 7)
```

```{r}
ggsave("../figures/pca_plots/BANS_PCA_norel_geo_labels_ordered.png", width = 8.5, height = 7)
```

Yay, that looks much cleaner. Let's see if ADMIX likes it.

# ADMIXTURE
## pcangsd admix
for eig in {1..5}
do
  for r in {1..5}
         do
          $CONDA_PREFIX/bin/python $pcangsd_path/pcangsd/main.py -p BANS.filtered.unrel_pruned -t 4 --admix --admix-seed $RANDOM -e $eig --admix-alpha 10 -o BANS.filtered.unrel_pruned.alpha10.e"$eig".run"$r"
        done
done
```{bash}
rsync -avzP ericacnr@colostate.edu@login.rc.colorado.edu:/scratch/alpine/ericacnr@colostate.edu/BANS/02.5.neutral_pop_struc/BANS.filtered.unrel_pruned0.2.alpha0*.Q /Users/ericarobertson/Desktop/BANS_Pub/analysis/03.delineate_ESUs/scratch/admix_output/pcadmix_output/
```

```{r}
library(pophelper)
library(data.table)
library(dplyr)
library(stringr)
library(readr)
library(tidyr)
library(ggplot2)
library(gridExtra)
library(grid)

# Inputs
q_dir   <- "../scratch/admix_output/pcadmix_output"
q_pat   <- "^BANS.filtered.unrel_pruned0.2.alpha0.*\\.Q$"
famfile <- "../scratch/BANS.filtered.unrel.fam"

popsfile   <- "../scratch/pops.csv"             # BGP_ID, Group
letterfile <- "../scratch/geo_informative_labels.RDS"           # Group/Pop, cluster_letter

out_pdf    <- "../figures/admix_plots/BANS_pcadmix_overall_alpha0_allK_labels.pdf"

# Read Q files + add sample IDs

sfiles <- list.files(path = q_dir, pattern = q_pat, full.names = TRUE)
stopifnot(length(sfiles) > 0)

slist <- readQ(files = sfiles, filetype = "basic", indlabfromfile = FALSE)

inorder <- fread(famfile, header = FALSE)[, .(IID = V1)]
stopifnot(length(unique(sapply(slist, nrow))) == 1)
stopifnot(nrow(slist[[1]]) == nrow(inorder))

slist <- lapply(slist, `rownames<-`, inorder$IID)


# Build cluster_letter per individual (aligned to Q rownames)

pops <- read_csv(popsfile, show_col_types = FALSE) %>%
  transmute(BGP_ID, Group)

letters <- readRDS(letterfile)

pops.cluster.letter <- pops %>%
  left_join(letters, by = "Group")

# vector of letters in the same order as Q rows
Letter_vec <- pops.cluster.letter$Group_label[
  match(rownames(slist[[1]]), pops.cluster.letter$BGP_ID)
]

# letter order (stable)
letter_levels <- letters %>%
  distinct(Group_label) %>%
  pull(Group_label) %>%
  trimws()


# Parse K/run from filenames

Kvec   <- as.integer(str_extract(basename(sfiles), "(?<=admix\\.)\\d+"))
Runvec <- as.integer(str_extract(basename(sfiles), "(?<=run)\\d+"))

Ks <- sort(unique(Kvec[!is.na(Kvec)]))
stopifnot(length(Ks) > 0)


# Component color helper (per K)

get_comp_cols <- function(K) {

  # fixed 10-color palette
  base_cols <- c(
    "#E69F00", # orange
    "#56B4E9", # sky blue
    "#009E73", # bluish green
    "#F0E442", # yellow
    "#0072B2", # blue
    "#D55E00", # vermillion
    "#CC79A7", # reddish purple
    "#999999", # gray
    "#44AA99", # teal
    "#AA4499"  # magenta
  )


  # if K > 10, interpolate additional colors
  if (K <= length(base_cols)) {
    cols <- base_cols[1:K]
  } else {
    cols <- colorRampPalette(base_cols)(K)
  }

  names(cols) <- paste0("ESU", seq_len(K))
  cols
}

# Build ggplots + assignments

g_list <- vector("list", length(Ks))
names(g_list) <- paste0("K", Ks)

assign_list <- vector("list", length(Ks))
names(assign_list) <- paste0("K", Ks)

for (K in Ks) {

  # pick the first run for this K
  idx <- which(Kvec == K)
  idx <- idx[order(Runvec[idx])]
  idx1 <- idx[1]

  # wide Q (rows = individuals)
  Qwide <- as.data.frame(slist[[idx1]])
  stopifnot(ncol(Qwide) == K)

  colnames(Qwide) <- paste0("ESU", seq_len(K))
  Qwide$ind <- rownames(Qwide)

  # add cluster_letter + (optional) Group
  meta_ind <- tibble(
    ind = rownames(slist[[idx1]]),
    Group_label = trimws(as.character(Letter_vec))
  ) %>%
    mutate(Group_label = factor(Group_label, levels = letter_levels))

  # long for ggplot
  q_long <- Qwide %>%
    left_join(meta_ind, by = "ind") %>%
    filter(!is.na(Group_label)) %>%
    pivot_longer(cols = starts_with("ESU"),
                 names_to = "popGroup",
                 values_to = "prob") %>%
    mutate(
      popGroup = factor(popGroup, levels = paste0("ESU", seq_len(K)))
    )

  # ggplot for this K
  comp_cols <- get_comp_cols(K)

  g_list[[paste0("K", K)]] <-
    ggplot(q_long, aes(x = ind, y = prob, fill = popGroup)) +
    geom_col(color = "gray", linewidth = 0.1) +
    facet_grid(~Group_label, switch = "x", scales = "free_x", space = "free_x") +
    theme_minimal() +
    labs(x = "", y = "Ancestry", title = paste0("K = ", K)) +
    scale_y_continuous(expand = c(0, 0)) +
    scale_x_discrete(expand = expansion(add = 1)) +
    theme(
      panel.spacing.x = unit(0.001, "lines"),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      strip.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
      panel.grid = element_blank()
    ) +
    scale_fill_manual(values = comp_cols, guide = "none")

}

pdf(out_pdf, width = 10, height = 2.2 * length(g_list), useDingbats = FALSE)
grid.arrange(
  grobs = g_list,
  layout_matrix = matrix(seq_along(g_list), ncol = 1, byrow = FALSE)
)
dev.off()
print(out_pdf)
```
#### double checking convergence
```{r}
library(tidyverse)

#log-likelihoods pulled from the output file for alpha=0: awk '/Log-likelihood:/ {print $NF}' pcadmix.22816373.out

# awk '
# /Log-likelihood:/ {
#   vals[n++] = $NF
# }
# END {
#   printf("loglik_vals <- c(\n  ")
#   for (i = 0; i < n; i++) {
#     printf("%s", vals[i])
#     if (i < n-1) {
#       if ((i+1) % 5 == 0)
#         printf(",\n  ")
#       else
#         printf(", ")
#     }
#   }
#   printf("\n)\n")
# }
# ' pcadmix.23335757.out

#pcadmix.23335757.out
loglik_vals <- c(
  -317379697.37537, -317379813.54059, -317379678.38043, -317379760.91672, -317379713.25396,
  -314899814.78644, -314899716.79290, -314899793.68513, -314899574.60226, -314899754.03714,
  -312851042.25835, -312850828.66974, -312851082.17316, -312851247.93471, -312850893.01092,
  -310629491.65431, -310629568.51323, -310629372.21268, -310629284.03814, -310629277.99590,
  -308942526.42675, -308942091.66666, -308942166.14086, -308942631.48534, -308942846.93230
)

Ks
runs_per_k <- 5

ll_df <- tibble(
  K = rep(Ks, each = runs_per_k),
  run = rep(1:runs_per_k, times = length(Ks)),
  logLik = loglik_vals
)

ll_sum <- ll_df %>%
  group_by(K) %>%
  summarise(
    meanLL = mean(logLik),
    sdLL   = sd(logLik),
    .groups = "drop"
  )

p_ll <- ggplot(ll_sum, aes(x = K, y = meanLL)) +
  geom_line() +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = meanLL - sdLL, ymax = meanLL + sdLL), width = 0.15) +
  labs(x = "K", y = "Mean log-likelihood (± SD across runs)") +
  theme_bw()

p_ll

p_ll_runs <- ggplot(ll_df, aes(x = K, y = logLik, group = run)) +
  geom_point(alpha = 0.8) +
  geom_line(alpha = 0.4) +
  theme_bw() +
  labs(x = "K", y = "Log-likelihood (each run)")

p_ll_runs
```