---
title: "03.b.grouping_scheme_outputs"
output: html_document
---

# LIBRARIES
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggthemes)
library(patchwork)
library(cowplot)
library(ggpubr)
library(sf)
library(raster)
library(terra)
library(rnaturalearth)
library(tess3r)
library(genoscapeRtools)
library(mignette)
library(ggplot2)
library(ggrepel)
library(pophelper)
library(data.table)
library(stringr)
library(gridExtra)
library(grid)
```

# INTRODUCTION
After evaluating the basic population structure information we can potentially make multiple different genetically based ESU groupings. For each of these options, we need a genoscape, and admixture plot for that K value. Potential schemes are:

G4 - K of 4
G5 - K of 5
G6 - K of 6

The following will be the initial code walk-through for G4, then implementation of functions to tidy up the rest and make iteration easier. The K values used for the following admixture and genoscapes are calculated using ANGST PCA functions, as described in 03.a.initial_popstructure.Rmd.

I read back in the genoscape outputs for each grouping scheme and run them in mignette (custom R package designed by Matt Desaix) to get conservation unit polygons that go into the spatial analysis later.

# GENOMICS ONLY ESU DELINEATION
## G4
### Genoscape
To start, load the .Q file. Load the admixture proportions and merge with a sample list you (same order as admix input). I did run multiple iterations for each K value, but I found that there was significant overlap across runs showing good convergence so for this analysis I will just be reading in Run 1 for the appropriate K value. I also testing different alpha values with no significant differences between alpha 0 and alpha 10, so I will again be using only the alpha0 iterations for these figures and subsequent analysis. I'm also using a custom color pallet established here and repeated throughout all additional analyses.

The admixture outputs (.Q files) don't have sample associated with them. Because I used the plink ped files as my input for ANGST when making these the sample order will be uniform across .Q files and the overall PCA eigen files. So, we can use the eigen files to pull the sample IDs.
```{r, message=FALSE, warning=FALSE}
Q_tibble <- read.table("../scratch/admix_output/pcadmix_output/BANS.filtered.unrel_pruned.alpha0.e3.run1.admix.4.Q") %>%
  setNames(paste0("ESU", 1:4))
pca_df <- read.table("../scratch/PCA/BANS.filtered.unrel.ld25-10-0.2.eigenvec", header = FALSE) %>% dplyr::select(-V2)

colnames(pca_df)[-1] <- paste0("PC", 1:(ncol(pca_df) - 1))
names(pca_df)[1] <- "ind"

Q_tibble$Sample <- pca_df$ind

K <- ncol(Q_tibble)-1

# Making the color pallet
cols <-  c(
  "ESU1" = "#C1839F",
  "ESU2" = "#FF5A5F",
  "ESU3" = "#8ACB88",
  "ESU4" = "#FFBF46")
barplot(
  rep(1, length(cols)),
  col = cols,
  border = NA,
  axes = FALSE,
  space = 0,
  names.arg = names(cols),
  las = 2
)
```
Now we pull in the sample data to get the lat long so we have spatial data for all the k values. This allows for the genoscape interpolation later.

```{r, message=FALSE, warning=FALSE, label="meta"}
all <- read_csv("../../../raw_data/BANS_all_sample_data_shifted.csv") %>% filter(BGP_ID %in% pca_df$ind)

ind_filtered <- all %>%
  mutate(Region_site=paste(State, Lat, Long, sep="_"), Region_pop=State)

LatLong_tibble <- Q_tibble %>%
  dplyr::select(Sample) %>%
  left_join(ind_filtered, by = c("Sample" = "BGP_ID")) %>%
  dplyr::select(Sample, Lat, Long)
```

Now we need the breading range, but in this bird that's a global map (from eBird), so we crop it down to just North America.
```{r, label="base map"}
map <- ne_states(country = c("United States of America", "Canada"), 
                 returnclass = 'sf')
na_union <- st_union(map)

breeding.sf <- st_read("../../../raw_data/spatial_files/banswa_range_2023.gpkg") %>% filter(season=="breeding")  %>% st_transform(crs(map))

#clip BANS range to just NA and make a single, clean polygon
breeding_na <- st_intersection(breeding.sf, na_union)

plot(st_geometry(breeding_na))
```

Alright, step on of the genoscape is to make the genoscape "brick". This is basically just the individual ESU K value interpolated across space. So there's a layer for each ESU with green showing where the K values are higher and white/pink being the lower values.
```{r, label="brick"}
# prepare the data for tess3Q_map_rasters
long_lat_tibble <- Q_tibble %>%
  dplyr::select(Sample) %>%
  left_join(LatLong_tibble, by = "Sample") %>%
  dplyr::select(Long, Lat) 

long_lat_matrix <- long_lat_tibble %>%
  as.matrix()

# make a matrix of Q values
Q_matrix <- Q_tibble %>%
  dplyr::select(-Sample) %>%
  as.matrix()

bad <- which(!is.finite(long_lat_matrix[,1]) | !is.finite(long_lat_matrix[,2]))
length(bad)

# interpolate the Q-values by kriging using tess3r::tess3Q_map_rasters()
genoscape_brick <- tess3r::tess3Q_map_rasters(
  x = Q_matrix, 
  coord = long_lat_matrix,  
  map.polygon = breeding_na,
  window = extent(breeding_na)[1:4],
  resolution = c(1000,1000), # if you want more cells in your raster, set higher
  # this next lines need to be here, but don't do much...
  col.palette = tess3r::CreatePalette(cols, K), 
  method = "map.max", 
  interpol = tess3r::FieldsKrigModel(10),  
  main = "Ancestry coefficients",
  xlab = "Longitude", 
  ylab = "Latitude", 
  cex = .4
)

# after that, we need to add names of the clusters back onto this raster brick
names(genoscape_brick) <- names(Q_tibble)[-(K+1)]
names(genoscape_brick)

# we write this out to use later for the spatial polygons
writeRaster(genoscape_brick, "../scratch/genoscape_output/BANS_overall_genoscape_brick_pcadmix_G4.tif", overwrite = TRUE)
```

Then we scale that and give it a coordinate system. This also kicks it to red, green, blue and alpha values which is what get's layered to make the colors we actually want for the final genoscape.
```{r}
# scaling and cleaning the genoscape_brick
genoscape_rgba <- genoscapeRtools::qprob_rando_raster(
  TRB = genoscape_brick,
  cols = cols, 
  alpha_scale = 3, 
  abs_thresh = .5, 
  alpha_exp = 1.55, 
  alpha_chop_max = 230
)

# at this point, we must be explicit about adding a projection to the raster.
# this adds the info for a regular lat-long projection
crs(genoscape_rgba) <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"
```

Now we need some spatial landmarks to make this resulting plot easier to interpret in space.
```{r, message=FALSE, warning=FALSE, results='hide'}
# get shapefiles from natural earth
coastlines <- ne_download(scale = 10, type = "coastline", category = "physical", returnclass = "sf")
countries <- ne_download(scale = 10, type = "admin_0_boundary_lines_land", category = "cultural", returnclass = "sf")
states <- ne_download(scale = 10, type = "admin_1_states_provinces_lines", category = "cultural", returnclass = "sf")
ocean <- ne_download(scale = 10, type = "ocean", category = "physical", returnclass = "sf")
lakes <- ne_download(scale = 10, type = "lakes", category = "physical", returnclass = "sf")
```

Finally, the plotting.
```{r, message=FALSE, warning=FALSE}
# reproject to a Lambert conic projection
lamproj <- "+proj=lcc +lat_1=20 +lat_2=60 +lat_0=40 +lon_0=-100 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"
genoscape_spat <- rast(genoscape_rgba)

genoscape_projected <- terra::project(genoscape_spat, lamproj, method = "near")

# generate bounding box to crop the above rnaturalearth layers to the same area as the genoscape
bb <- st_bbox(breeding_na) %>% st_transform(lamproj)

coords_sf <- st_as_sf(
  LatLong_tibble,
  coords = c("Long", "Lat"),
  crs = 4326   # WGS84 (lon/lat)
)

# then we want to get the sample points as coded by letter, which I established previously
cluster_geo <- readRDS("../scratch/geo_informative_labels_centroids.RDS")

## make sf points from lon/lat
cluster_geo_sf <- st_as_sf(
  cluster_geo,
  coords = c("Y", "X"),
  crs = 4326,      # lon/lat
  remove = FALSE
)

## transform to your map CRS (lamproj must be a valid CRS/proj string)
cluster_geo_sf <- st_transform(cluster_geo_sf, st_crs(lamproj))

# plot it
genoscape <- ggplot() +
  ggspatial::layer_spatial(genoscape_projected) +
  geom_sf(data = breeding_na, fill = NA, linewidth = 0.1225 , alpha = 0.6) +
  geom_sf(data = countries, fill = NA, linewidth = 0.25) +
  geom_sf(data = coastlines, fill = NA, linewidth = 0.1225) +
  geom_sf(data = states, fill = NA, linewidth = 0.1225) +
  geom_sf(data = ocean, fill = "#bddbe6") +
  geom_sf(data = lakes, fill = "#bddbe6") + #bddbe6
  #geom_spatial_point(data = long_lat_tibble, mapping = aes(x = Long, y = Lat), size=1, alpha = 0.7) + 
  theme(panel.border = element_rect(color = "black", fill = NA, linewidth = 1)) +
  geom_sf(data=cluster_geo_sf)+
  coord_sf(
  crs = st_crs(lamproj),
  xlim = c(bb["xmin"], bb["xmax"]),
  ylim = c(bb["ymin"], bb["ymax"]),
  expand = FALSE)+
  geom_label_repel(
    data = cluster_geo_sf,
    aes(label = Group_label, geometry = geometry),
    stat = "sf_coordinates",
    size = 2,            # increase text size (try 2.2–3.0
    fontface = "bold",
    label.size = .2,
    fill = "white",
    alpha = 1,
    box.padding = 0.25,
    point.padding = 0.2,
    segment.color = "black",
    segment.size = 0.5,
    min.segment.length = 0
)
genoscape
```
Save as both a png and pdf
```{r}
ggsave(plot=genoscape, "../figures/genoscape_plots/BANS_genoscape_G4_overall.png")

pdf("../figures/genoscape_plots/BANS_genoscape_G4_overall.pdf")
print(genoscape)
dev.off()
```

















### Admixture Plots
Alright, now we just want an appropriately colored admixture plot for each K value we've selected.
```{r}
K_target <- 4

# Read Q file
Q_tibble <- read.table(
  "../scratch/admix_output/pcadmix_output/BANS.filtered.unrel_pruned0.2.alpha0.e3.run1.admix.4.Q"
) %>%
  rename(
    ESU1 = V1,
    ESU2 = V2,
    ESU3 = V3,
    ESU4 = V4
  )

# Read PCA file
pca_df <- read.table(
  "../scratch/PCA/BANS.filtered.unrel.ld25-10-0.2.eigenvec",
  header = FALSE
) %>%
  dplyr::select(-V2)

colnames(pca_df)[-1] <- paste0("PC", 1:(ncol(pca_df) - 1))
names(pca_df)[1] <- "ind"

# Make sure the row orders match in length
stopifnot(nrow(Q_tibble) == nrow(pca_df))

# Add individual IDs
Q_tibble <- Q_tibble %>%
  mutate(ind = pca_df$ind)

# Colors
component_colors <- readRDS("../figures/colors/CU_grouping_colors_2_G4.rds")
names(component_colors) <- paste0("ESU", 1:K_target)

# Cluster membership and letter
pops <- read.csv("../scratch/pops.csv") %>%
  transmute(ind = BGP_ID, Group)

letter_key <- read.csv("../scratch/cluster_letter_key_geo_lonlat.csv")

# Check joins will not duplicate individuals
stopifnot(!anyDuplicated(pops$ind))
stopifnot(!anyDuplicated(letter_key$Group))

pops.cluster.letter <- pops %>%
  left_join(letter_key, by = "Group")

Q_tibble2 <- Q_tibble %>%
  left_join(pops.cluster.letter, by = "ind")

# Check join did not change row number
stopifnot(nrow(Q_tibble2) == nrow(Q_tibble))

Q_tibble2 <- Q_tibble2 %>%
  filter(!is.na(cluster_letter))

# Desired facet order
desired_order <- c(
  "A", "B", "C", "E", "H", "J", "L", "O", "P", "Q", "R", "S", 
  "T", "U", 
  "G", "I", "K", "M", "N",
  "D", "F"
)

# Preserve individual order from the Q file, grouped by desired letter order
ind_order <- Q_tibble2 %>%
  mutate(cluster_letter = factor(cluster_letter, levels = desired_order)) %>%
  arrange(cluster_letter) %>%
  pull(ind) %>%
  unique()

# Long format
q_long <- Q_tibble2 %>%
  mutate(
    cluster_letter = factor(cluster_letter, levels = desired_order),
    ind = factor(ind, levels = ind_order)
  ) %>%
  pivot_longer(
    cols = all_of(paste0("ESU", 1:K_target)),
    names_to = "popGroup",
    values_to = "prob"
  ) %>%
  mutate(
    popGroup = factor(popGroup, levels = paste0("ESU", 1:K_target))
  )

# Diagnostic: check letter-to-dominant-ESU association
q_long %>%
  group_by(ind, cluster_letter, popGroup) %>%
  summarize(prob = first(prob), .groups = "drop") %>%
  group_by(ind, cluster_letter) %>%
  slice_max(prob, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  count(cluster_letter, popGroup)

# Plot
K_plot <- ggplot(q_long, aes(x = ind, y = prob, fill = popGroup)) +
  geom_col(color = "gray", linewidth = 0.1) +
  facet_grid(~cluster_letter, switch = "x", scales = "free", space = "free") +
  theme_minimal() +
  labs(x = "", y = "Ancestry", title = "") +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_discrete(expand = expansion(add = 1)) +
  scale_fill_manual(values = component_colors, guide = "none") +
  theme(
    panel.spacing.x = unit(0.001, "lines"),
    axis.text.x = element_blank(),
    strip.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5, size = 8),
    panel.grid = element_blank()
  )

K_plot

ggsave("../figures/admix_plots/BANS_admix_overall_alpha0_G4_letters_ordered.png")

pdf("../figures/admix_plots/BANS_admix_overall_alpha0_G4_letters_ordered.pdf", width = 10, height = 2.2)
print(K_plot)
dev.off()
```

Then we need this little admix assignment code for the colored PCA in the next step...
```{r}
admix_group <- q_long %>%
  group_by(ind) %>%
  slice_max(prob, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(
    BGP_ID = ind,
    admix_group = popGroup   # e.g. AU1–AU5
  )

write_tsv(admix_group, "../scratch/admix_output/admix_ESU_group_G4.tsv")
```

### Colored PCA Plots
#### colored by ESU, solid color
```{r}
plink_pca <- read.table("../intermediates/pca/PCA_eigen/BANS.filtered.unrel.ld25-10-0.5.eigenvec", header = FALSE) %>% dplyr::select(-V2)
eigenval <- scan("../intermediates/pca/PCA_eigen/BANS.filtered.unrel.ld25-10-0.5.eigenval")

pve05 <- data.frame(PC = 1:189, pve = eigenval/sum(eigenval)*100)
ggplot(pve05[1:20, ], aes(x = PC, y = pve)) +
  geom_col(width = 0.9) +
  labs(x = "PC", y = "PVE (%)") +
  theme_bw()

# set names
pca05 <- plink_pca
colnames(pca05)[-1] <- paste0("PC", 1:(ncol(pca05) - 1))
names(pca05)[1] <- "ind"
nrow(pca05)
#correct number of samples

cols <- readRDS("../figures/colors/pcadmix_ESU_5_colors_2.rds")

pops <- read_csv("../intermediates/pops.csv", show_col_types = FALSE) %>%
  transmute(BGP_ID, Group)

admix_groups <- fread("../intermediates/admix_output/admix_ESU_group_K5.tsv")

pops.groups <- left_join(pops, admix_groups, by="BGP_ID") %>% rename(ind=BGP_ID)


# Join PCA data
pca.cluster <- left_join(pca05, pops.groups, by="ind")  # left_join PCA first to preserve PC columns


rel.nr <- ggplot(pca.cluster, aes(
    x = PC1, 
    y = PC2, 
    col = admix_group,
    text = paste0(
        "ID: ", ind, "<br>",
        "Group: ", Group, "<br>",
        "PC1: ", round(PC1, 4), "<br>",
        "PC2: ", round(PC2, 4)
    )
)) +
  geom_point(size = 3) +
  scale_color_manual(values = cols,
                     name = "ESU Group") +
  theme_light() +
  xlab(paste0("PC1 (", signif(pve05$pve[1], 3), "%)")) +
  ylab(paste0("PC2 (", signif(pve05$pve[2], 3), "%)"))

rel.nr

ggsave("../figures/pca/BANS_overall_PCA_K5_norel.png", width = 8.5, height = 7)
```

#### colored by ESU with gradient for location
```{r}
# 1) Join letters onto your per-individual table
letters <- read.csv("../intermediates/cluster_letter_key_geo_lonlat.csv", stringsAsFactors = FALSE)

# make sure the join key is Group (some files use Pop)
if ("Pop" %in% names(letters) && !"Group" %in% names(letters)) {
  letters <- letters %>% rename(Group = Pop)
}
stopifnot(all(c("Group", "cluster_letter") %in% names(letters)))

pops.groups.letters <- left_join(pops.groups, letters, by = "Group") %>%
  mutate(cluster_letter = trimws(as.character(cluster_letter)))

ESUs <- paste0("ESU", 1:5)

letter_owner <- pops.groups.letters %>%
  filter(!is.na(cluster_letter), !is.na(admix_group), admix_group %in% ESUs) %>%
  count(cluster_letter, admix_group, name = "n") %>%
  group_by(cluster_letter) %>%
  slice_max(n, n = 1, with_ties = FALSE) %>%   # majority ESU for that letter
  ungroup() %>%
  rename(owner_esu = admix_group)

# (optional) inspect "mixed" letters
# pops.groups.letters %>% count(cluster_letter, admix_group) %>% arrange(cluster_letter, desc(n))

# Add the owner ESU back to each individual (so PCA tooltips can show it)
pops.groups.letters <- pops.groups.letters %>%
  left_join(letter_owner %>% select(cluster_letter, owner_esu), by = "cluster_letter")

esu_base_cols <- cols  # named vector: ESU1..ESU5 -> hex

letter_cols_df <- letter_owner %>%
  arrange(owner_esu, cluster_letter) %>%       # stable ordering
  group_by(owner_esu) %>%
  group_modify(~{
    base <- esu_base_cols[[.y$owner_esu]]
    L <- .x$cluster_letter
    n <- length(L)
    shades <- colorRampPalette(c("white", base))(n + 1)[-1]
    tibble(cluster_letter = L, col = shades)
  }) %>%
  ungroup()

letter_cols <- setNames(letter_cols_df$col, letter_cols_df$cluster_letter)

# sanity: must be unique names
stopifnot(!any(duplicated(names(letter_cols))))

# quick preview
barplot(rep(1, length(letter_cols)),
        col = letter_cols,
        border = NA,
        names.arg = names(letter_cols),
        las = 2,
        cex.names = 0.3)

saveRDS(letter_cols, file = "../figures/colors/pcadmix_ESU_5_colors_2.rds")


plink_pca <- read.table("../intermediates/pca/PCA_eigen/BANS.filtered.unrel.ld25-10-0.5.eigenvec",
                        header = FALSE) %>% dplyr::select(-V2)
eigenval <- scan("../intermediates/pca/PCA_eigen/BANS.filtered.unrel.ld25-10-0.5.eigenval")

pve05 <- data.frame(PC = 1:189, pve = eigenval/sum(eigenval)*100)

pca05 <- plink_pca
colnames(pca05)[-1] <- paste0("PC", 1:(ncol(pca05) - 1))
names(pca05)[1] <- "ind"

# make sure pops.groups.letters has an 'ind' column
pops_for_join <- pops.groups.letters
if ("BGP_ID" %in% names(pops_for_join) && !"ind" %in% names(pops_for_join)) {
  pops_for_join <- pops_for_join %>% rename(ind = BGP_ID)
}
if ("sampleID" %in% names(pops_for_join) && !"ind" %in% names(pops_for_join)) {
  pops_for_join <- pops_for_join %>% rename(ind = sampleID)
}
stopifnot("ind" %in% names(pops_for_join))

pca.cluster <- left_join(pca05, pops_for_join, by = "ind") %>%
  mutate(cluster_letter = trimws(as.character(cluster_letter)))

# lock palette mapping to letters we actually have
letter_cols <- letter_cols[order(names(letter_cols))]

rel.nr <- ggplot(pca.cluster, aes(
  x = PC1, y = PC2, col = cluster_letter,
  text = paste0(
    "ID: ", ind, "<br>",
    "Group: ", Group, "<br>",
    "Individual ESU: ", admix_group, "<br>",
    "Letter owner ESU: ", owner_esu, "<br>",
    "PC1: ", round(PC1, 4), "<br>",
    "PC2: ", round(PC2, 4)
  )
)) +
  geom_point(size = 3) +
  scale_color_manual(
    values = letter_cols,
    limits = names(letter_cols),
    drop = FALSE,
    name = "Sampling location\nColored by ESU"
  ) +
  theme_light() +
  xlab(paste0("PC1 (", signif(pve05$pve[1], 3), "%)")) +
  ylab(paste0("PC2 (", signif(pve05$pve[2], 3), "%)"))

rel.nr
ggsave("../figures/pca/BANS_ESU_K5_PCA_norel_locationcoloring.png", width = 8.5, height = 7)
```

# FUNCTIONS
(Note, the following functions were generated using ChatGPT based on the above written out code that I produced, interpretation of results is my own)
## Color Pallets
The trick here is to get the color right for the spatial ESU rather than having the numbers be the same, cause the numbers aren't consistent across space with different K values.
```{r}
# G4
cols_G3 <-  c(
  "ESU1" = "#FF5A5F",
  "ESU2" = "#187795",
  "ESU3" = "#FFBF46")
barplot(
  rep(1, length(cols_G3)),
  col = cols_G3,
  border = NA,
  axes = FALSE,
  space = 0,
  names.arg = names(cols_G3),
  las = 2
)
saveRDS(cols_G3, "../figures/colors/CU_grouping_colors_3_G3.rds")

# G4
cols_G4 <-  c(
  "ESU1" = "#FF5A5F",
  "ESU2" = "#187795",
  "ESU3" = "#FFBF46",
  "ESU4" = "#C1839F")
barplot(
  rep(1, length(cols_G4)),
  col = cols_G4,
  border = NA,
  axes = FALSE,
  space = 0,
  names.arg = names(cols_G4),
  las = 2
)
saveRDS(cols_G4, "../figures/colors/CU_grouping_colors_3_G4.rds")

# G5
cols_G5 <-  c(
  "ESU1" = "#FF5A5F",
  "ESU2" = "#FFBF46",
  "ESU3" = "#2589BD",
  "ESU4" = "#C1839F",
  "ESU5" = "#8ACB88")
barplot(
  rep(1, length(cols_G5)),
  col = cols_G5,
  border = NA,
  axes = FALSE,
  space = 0,
  names.arg = names(cols_G5),
  las = 2
)
saveRDS(cols_G5, "../figures/colors/CU_grouping_colors_2_G5.rds")

cols_G5 <-  c(
  "ESU1" = "#187795",
  "ESU2" = "#FF5A5F",
  "ESU3" = "#FFBF46",
  "ESU4" = "#C1839F",
  "ESU5" = "#8ACB88")
barplot(
  rep(1, length(cols_G5)),
  col = cols_G5,
  border = NA,
  axes = FALSE,
  space = 0,
  names.arg = names(cols_G5),
  las = 2
)
saveRDS(cols_G5, "../figures/colors/CU_grouping_colors_3_G5.rds")

# G6
cols_G6 <-  c(
  "ESU1" = "#FF5A5F",
  "ESU2" = "#FFBF46",
  "ESU3" = "#2589BD",
  "ESU4" = "#C1839F",
  "ESU5" = "#8ACB88",
  "ESU6" = "#247B7B")
barplot(
  rep(1, length(cols_G6)),
  col = cols_G6,
  border = NA,
  axes = FALSE,
  space = 0,
  names.arg = names(cols_G6),
  las = 2
)
saveRDS(cols_G6, "../figures/colors/CU_grouping_colors_2_G6.rds")
```

## Genoscapes
```{r}
source("prep_genoscape_inputs.R")
source("build_genoscape_surfaces.R")

############# PICK A K VALUE #############
scheme <- "G3"
K <- 3
# read in data, either raw or forced
Q_tibble <- read.table("../scratch/admix_output/pcadmix_output/BANS.filtered.unrel_pruned0.2.alpha0.e2.run1.admix.3.Q") %>%
setNames(paste0("ESU", 1:K))

scheme <- "G5"
K <- 5
# read in data, either raw or forced
Q_tibble <- read.table("../scratch/admix_output/pcadmix_output/BANS.filtered.unrel_pruned0.2.alpha0.e4.run1.admix.5.Q") %>%
setNames(paste0("ESU", 1:K))

scheme <- "G4"
K <- 4
Q_tibble <- read.table("../scratch/admix_output/pcadmix_output/BANS.filtered.unrel_pruned0.2.alpha0.e3.run1.admix.4.Q") %>%
setNames(paste0("ESU", 1:K))

out_plot_genoscape_png <- paste0("BANS_genoscape_",scheme,"_overall_final.png")
out_plot_genoscape_pdf <- paste0("BANS_genoscape_",scheme,"_overall_final.pdf")
genoscape_path <- "../figures/genoscape_plots/"
genoscape_brick_out <- paste0("../scratch/genoscape_output/BANS_overall_genoscape_brick_pcadmix_",scheme,".tif")
cols <- readRDS(paste0("../figures/colors/CU_grouping_colors_3_",scheme,".rds"))
barplot(rep(1, length(cols)), col = cols, border = NA, axes = FALSE, space = 0, names.arg = names(cols), las = 2)

############# OTHER INPUT FILES #############
pca_df <- read.table("../scratch/PCA/BANS.filtered.unrel.ld25-10-0.2.eigenvec", header = FALSE)

all <- read_csv("../../../raw_data/BANS_all_sample_data_shifted.csv")
############# MAKE Q_TIBBLE #############
prep_out <- prep_genoscape_inputs(
  Q_tibble = Q_tibble,
  pca_df = pca_df,
  all = all
)

Q_tibble <- prep_out$Q_tibble
Q_matrix <- prep_out$Q_matrix
long_lat_matrix <- prep_out$long_lat_matrix
LatLong_tibble <- prep_out$LatLong_tibble
K <- ncol(Q_matrix)
q_cols <- colnames(Q_matrix)

############# GET BASE MAP AND RANGE #############
map <- ne_states(
  country = c("United States of America", "Canada"),
  returnclass = "sf"
)

na_union <- st_union(map)

breeding.sf <- st_read("../../../raw_data/spatial_files/banswa_range_2023.gpkg") %>%
  filter(season == "breeding") %>%
  st_transform(crs(map))

breeding_na <- st_intersection(breeding.sf, na_union)

############# BUILD THE GENOSCAPE #############
surf_out <- build_genoscape_surfaces(
  Q_matrix = Q_matrix,
  long_lat_matrix = long_lat_matrix,
  breeding_na = breeding_na,
  cols = cols,
  K = K,
  brick_names = q_cols,
  out_tif = genoscape_brick_out,
  alpha_threshold = 0,
  krig_val = 30
)

genoscape_brick <- surf_out$genoscape_brick
writeRaster(genoscape_brick, genoscape_brick_out, overwrite=TRUE)
genoscape_projected <- surf_out$genoscape_projected

coords_sf <- st_as_sf(
  LatLong_tibble,
  coords = c("Long", "Lat"),
  crs = 4326   # WGS84 (lon/lat)
)

# then we want to get the sample points as coded by letter, which I established previously
cluster_geo <- readRDS("../scratch/geo_informative_labels_centroids.RDS")

## make sf points from lon/lat
cluster_geo_sf <- st_as_sf(
  cluster_geo,
  coords = c("Y", "X"),
  crs = 4326,      # lon/lat
  remove = FALSE
)

## transform to your map CRS (lamproj must be a valid CRS/proj string)
cluster_geo_sf <- st_transform(cluster_geo_sf, st_crs(lamproj))

genoscape <- ggplot() +
  ggspatial::layer_spatial(genoscape_projected) +
  geom_sf(data = breeding_na, fill = NA, linewidth = 0.1225 , alpha = 0.6) +
  geom_sf(data = countries, fill = NA, linewidth = 0.25) +
  geom_sf(data = coastlines, fill = NA, linewidth = 0.1225) +
  geom_sf(data = states, fill = NA, linewidth = 0.1225) +
  geom_sf(data = ocean, fill = "#bddbe6") +
  geom_sf(data = lakes, fill = "#bddbe6") + #bddbe6
  #geom_spatial_point(data = long_lat_tibble, mapping = aes(x = Long, y = Lat), size=1, alpha = 0.7) + 
  theme(panel.border = element_rect(color = "black", fill = NA, linewidth = 1)) +
  geom_sf(data=cluster_geo_sf,
          size = 0.5)+
  coord_sf(
  crs = st_crs(lamproj),
  xlim = c(bb["xmin"], bb["xmax"]),
  ylim = c(bb["ymin"], bb["ymax"]),
  expand = FALSE)+
  geom_label_repel(
    data = cluster_geo_sf,
    aes(label = Group_label, geometry = geometry),
    stat = "sf_coordinates",
    size = 1.5,            # increase text size (try 2.2–3.0
    fontface = "bold",
    label.size = .1,
    label.padding = unit(0.1, "lines"),
    label.r=0,
    fill = "white",
    alpha = 1,
    point.padding = 0.1,
    segment.color = "black",
    segment.size = 0.25,
    min.segment.length = 0
)
genoscape

############# SAVE THE GENOSCAPE #############
out_plot_genoscape_png <- paste0("BANS_genoscape_",scheme,"_overall_newcol2.png")
ggsave(plot=genoscape, path=genoscape_path, filename=out_plot_genoscape_png, width = 7, height = 4.5)

pdf(paste0(genoscape_path,out_plot_genoscape_pdf))
print(genoscape)
dev.off()
```

## Admixture plots
```{r}
source("make_admix_barplot_and_assignments.R")

######## INPUT #########
K <- 3
out_plot_admix <- "BANS_admix_overall_alpha0_G3_letters_2.png"
admix_plot_path <- "../figures/admix_plots/"

out_group_admix <- "BANS_admix_grouping_G3_2.tsv"
admix_group_path <- "../scratch/admix_output/"

Q_tibble <- read.table("../scratch/admix_output/pcadmix_output/BANS.filtered.unrel_pruned0.2.alpha0.e2.run1.admix.3.Q") %>%
setNames(paste0("ESU", 1:K))
cols <- readRDS("../figures/colors/CU_grouping_colors_3_G3.rds")

K <-  5
out_plot_admix <- "BANS_admix_overall_alpha0_G5_letters_2.png"
admix_plot_path <- "../figures/admix_plots/"

out_group_admix <- "BANS_admix_grouping_G5_2.tsv"
admix_group_path <- "../scratch/admix_output/"

Q_tibble <- read.table("../scratch/admix_output/pcadmix_output/BANS.filtered.unrel_pruned0.2.alpha0.e4.run1.admix.5.Q") %>%
  setNames(paste0("ESU", 1:K))
cols <- readRDS("../figures/colors/CU_grouping_colors_3_G5.rds")

K <-  4
out_plot_admix <- "BANS_admix_overall_alpha0_G4_letters_2.png"
admix_plot_path <- "../figures/admix_plots/"

out_group_admix <- "BANS_admix_grouping_G4_2.tsv"
admix_group_path <- "../scratch/admix_output/"

Q_tibble <- read.table("../scratch/admix_output/pcadmix_output/BANS.filtered.unrel_pruned0.2.alpha0.e3.run1.admix.4.Q") %>%
  setNames(paste0("ESU", 1:K))
cols <- readRDS("../figures/colors/CU_grouping_colors_3_G4.rds")

######## RUNNING FUNCTION ########
pca_df <- read.table("../scratch/PCA/BANS.filtered.unrel.ld25-10-0.2.eigenvec", header = FALSE)

pops <- readr::read_csv("../scratch/pops.csv", show_col_types = FALSE)
letter_key <- readRDS("../scratch/geo_informative_labels_centroids.RDS")

group_label_order <- c("WA", "NT", 
                       "AK", "YT", "SK", "MB_1", "MB_2", "MT", "ND", "WY_1", "WY_2", "AB", "BC_1", "BC_2",
                       "ON", "QC_1", "QC_2", "NS_1", "NS_2",
                       "NB", "PE", "NL")

admix_out <- make_admix_barplot_and_assignments(
  K_target = K,
  cols = cols,
  group_label_order = group_label_order,
  threshold = 0.55,
  Q_tibble = Q_tibble,
  pca_df = pca_df,
  pops = pops,
  geo_key = letter_key,
  out_plot = paste0(admix_plot_path,out_plot_admix),
  out_assignments = paste0(admix_group_path,out_group_admix)
)

K_plot <- admix_out$K_plot
admix_group <- admix_out$admix_group
q_long <- admix_out$q_long

# This plot is meaningless for the GM ones but easiest to just write it out anyway
K_plot
ggsave(path=admix_plot_path, filename=out_plot_admix, plot=K_plot, width = 10, height = 2.2, units = "in", dpi = 300, bg = "white")

write_tsv(admix_group, paste0(admix_group_path,out_group_admix))

############
pdf("../figures/admix_plots/BANS_admix_overall_alpha0_G4_letters.pdf", width = 10, height = 2.2)
print(K_plot)
dev.off()
```

## PCA plots
```{r}
source("plot_pca_admix_solid.R")

out_plot_pca1 <- "../figures/pca_plots/BANS_G5_PCA_norel_ESUcoloring.png"
out_plot_pca2 <- "../figures/pca_plots/BANS_G5_PCA_norel_ESUcoloring.png"
out_letters_cols <- "../figures/colors/CU_grouping_letter_colors_G5.rds"

admix_groups <- data.table::fread("../scratch/admix_output/BANS_admix_grouping_G5_2.tsv")
cols <- readRDS("../figures/colors/CU_grouping_colors_2_G5.rds")

plink_pca <- read.table("../scratch/PCA/BANS.filtered.unrel.ld25-10-0.5.eigenvec", header = FALSE)
pops <- readr::read_csv("../scratch/pops.csv", show_col_types = FALSE)
eigenval <- scan("../scratch/PCA/BANS.filtered.unrel.ld25-10-0.5.eigenval")

solid_out <- plot_pca_admix_solid(
  plink_pca = plink_pca,
  eigenval = eigenval,
  cols = cols,
  pops = pops,
  admix_groups = admix_groups,
  out_plot = out_plot_pca1
)

solid_out$pca_plot

########## GRADIENT #########
source("plot_pca_location_gradient.R")

letters <- readRDS("../scratch/geo_informative_labels_centroids.RDS")
desired_order <- c(
  "AK", "YT", "BC_1", "BC_2", "AB", "SK", "MT", "MB_1", "MB_2", "ND", "WY_1", "WY_2", 
  "WA",
  "ON", "NS_1", "NS_2", "QC_1", "QC_2", 
  "NL", "PE", "NB"
)

grad_out <- plot_pca_location_gradient(
  plink_pca = plink_pca,
  eigenval = eigenval,
  cols = cols,
  pops = pops,
  admix_groups = admix_groups,
  geo_key = letters,
  desired_order=desired_order,
  out_plot = out_plot_pca2, 
  out_letter_cols_rds = out_letters_cols)

grad_out$pca_plot

pdf("../figures/pca_plots/BANS_GM6_PCA_norel_ESUcoloring.pdf", width=6, height=4)
print(grad_out$pca_plot)
dev.off()
```

# GENOSCAPE SHAPE FILES
I want to create shapefiles for each genetic group to get demographic data for, either from ebird population trends or BBS. Matt has function for this in Mignette, so install and load that package if you haven't already.
```{r, message = FALSE, warning = FALSE}
library(mignette)

lamproj <- "+proj=lcc +lat_1=20 +lat_2=60 +lat_0=40 +lon_0=-100 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"

out_polygon <- "../outputs/BANS_CU_polygons_thresh45_G5.shp"
genoscape_brick <- raster::brick("../scratch/genoscape_output/BANS_overall_genoscape_brick_pcadmix_G5.tif")
names(genoscape_brick) <- paste0("ESU", 1:5)

crs(genoscape_brick) <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"
genoscape_brick_spat <- terra::rast(genoscape_brick)
genoscape_brick_proj <- terra::project(genoscape_brick_spat, lamproj, method = "near")

genoscape_polygons <- mignette::scape_to_shape(x = genoscape_brick_proj, 
                                               prob_threshold = 0.4,
                                               d = 100,
                                               f = 100,
                                               s = 3)

polygons <- ggplot() +
  ggspatial::layer_spatial(genoscape_polygons, aes(fill = as.factor(Cluster))) +
  #geom_sf(data = breeding_range, fill = NA, linewidth = 0.1225 , alpha = 0.6) +
  geom_sf(data = countries, fill = NA, linewidth = 0.25) +
  geom_sf(data = coastlines, fill = NA, linewidth = 0.1225) +
  geom_sf(data = states, fill = NA, linewidth = 0.1225) +
  #geom_sf(data = ocean, fill = "#bddbe6") +
  #geom_sf(data = lakes, fill = "#bddbe6") +
  #geom_spatial_point(data = long_lat_tibble, mapping = aes(x = Long, y = Lat), size=2) +
  theme_void() + 
  #theme(panel.border = element_rect(color = "black", fill = NA, linewidth = 1)) +
  coord_sf(
  crs = st_crs(lamproj),
  xlim = c(bb["xmin"], bb["xmax"]),
  ylim = c(bb["ymin"], bb["ymax"]),
  expand = FALSE)  # expand = FALSE means put the axes right at the xlim and ylim

polygons
st_write(genoscape_polygons, out_polygon, delete_dsn = TRUE)
```