---
title: "04.a.spatial_predictors"
output: html_document
---

This is going to wrangling the spatial data I want to include in my GAMM model for predicting decline in Bank Swallow.
# Libraries
```{r, message=FALSE, warning=FALSE}
library(ncdf4)
library(terra)
library(raster)
library(dplyr)
library(purrr)
library(data.table)
library(ebirdst)
library(fields)
library(rnaturalearth)
library(sf)
```

# INTRODUCTION
This is the spatial layer wrangling for the spatial GAM modeling. Inputs are the raw spatial files from TerraClim, eBird, and the CU polygons output from 03.b.grouping_scheme_outputs.Rmd

# PREP
## Climate Data
### example delta between 2012 and 2022
```{r}
ncfile <- "../scratch/spatial_predictors/TerraClimate/TerraClimate_tmax_2012.nc"
r <- rast(ncfile)
r2012 <- r[[6]]

ncfile <- "../scratch/spatial_predictors/TerraClimate/TerraClimate_tmax_2022.nc"
r <- rast(ncfile)
r2022 <- r[[6]]

delta <- r2022 - r2012

na_bbox <- ext(-170, -50, 5, 85)
delta_na <- crop(delta, na_bbox)

writeRaster(delta_na, "../scratch/spatial_predictors/tmean_2022_minus_2012_NA.tif", overwrite = TRUE)

plot(delta_na, main = "Mean annual T (2022) − Mean annual T (2012) (°C)")
```

### Loop for all variables
```{r}
file_paths12 <- list.files(
  path = "../scratch/spatial_predictors/TerraClimate",
  pattern = "2012\\.nc$",
  full.names = TRUE
)
file_paths22 <- list.files(
  path = "scratch/TerraClimate",
  pattern = "2022\\.nc$",
  full.names = TRUE
)

layer_names.ny <- basename(file_paths12) |>
  sub("^TerraClimate_", "", x = _) |>
  sub("_2012\\.nc$", "", x = _) |>
  unique()

climate12 <- lapply(file_paths12, rast)
climate22 <- lapply(file_paths22, rast)
names(climate) <- layer_names

for (i in 1:5) {
  r2012_june <- climate12[[i]][[6]]
  r2022_june <- climate22[[i]][[6]]
  
  delta <- r2022_june - r2012_june
  
  na_bbox <- ext(-170, -50, 5, 85)
  delta_na <- crop(delta, na_bbox)
  
  writeRaster(
    delta_na,
    paste0("../scratch/spatial_predictors/delta_climate/", layer_names.ny[[i]], "_june_delta.tif"),
    overwrite = TRUE
  )
}

test <- rast("../scratch/spatial_predictors/delta_climate/ws_june_delta.tif")
test
```

## eBIRD Trends
```{r}
## BANS decline: eBird trends

set_ebirdst_access_key("srtemkuj2h9h", overwrite = TRUE)
ebirdst_download_trends(species = "banswa")

trends_bans <- load_trends("banswa")

trends_sf <- st_as_sf(
  trends_bans,
  coords = c("longitude", "latitude"),
  crs = 4326
)

plot(trends_sf)

trends_circles <- vectorize_trends(
  trends_bans,
  crs = "+proj=eqearth +lon_0=-96"
)

max_trend <- ceiling(max(abs(trends_circles$abd_trend)))
legend_breaks <- seq(0, 40, by = 10)
legend_breaks[length(legend_breaks)] <- max_trend
legend_breaks <- c(-rev(legend_breaks), legend_breaks) |> unique()
legend_labels <- c("<=-40", -20, 0, 20, ">=40")
legend_colors <- ebirdst_palettes(length(legend_breaks) - 1, type = "trends")

# assign colors to circles
trends_circles <- trends_circles |> 
  mutate(
    color = cut(abd_trend, legend_breaks, labels = legend_colors) |>
      as.character()
  )

countries <- ne_countries(returnclass = "sf", continent = "North America") |> 
  st_geometry() |> 
  st_transform(st_crs(trends_circles))

states <- ne_states(iso_a2 = c("US", "CA", "MX")) |> 
  st_geometry() |> 
  st_transform(st_crs(trends_circles))

# plot
plot(st_geometry(trends_circles), border = NA, col = NA)
plot(countries, col = "#cfcfcf", border = "#888888", add = TRUE)
plot(st_geometry(trends_circles),
     col = trends_circles$color, border = NA,
     axes = FALSE, bty = "n", reset = FALSE, add = TRUE)
lines(vect(countries), col = "#ffffff", lwd = 3)
lines(vect(states), col = "#ffffff", lwd = 1.5, xpd = TRUE)

label_breaks <- seq(0, 1, length.out = length(legend_breaks))
image.plot(zlim = c(0, 1), breaks = label_breaks, col = legend_colors,
           smallplot = c(0.90, 0.93, 0.15, 0.85),
           legend.only = TRUE,
           axis.args = list(at = c(0, 0.25, 0.5, 0.75, 1), 
                            labels = legend_labels,
                            col.axis = "black", fg = NA,
                            cex.axis = 0.7, lwd.ticks = 0,
                            line = -0.75),
           legend.args = list(text = "Abundance Trend [% change]",
                              side = 2, line = 0.25))

trends_raster <- rasterize_trends(trends_bans, layers = "abd_trend")
trends_raster.proj <- project(trends_raster, "EPSG:4326")
writeRaster(trends_raster.proj, "../scratch/spatial_predictors/eBird_trends_BANS_proj4326.tif")
plot(trends_raster.proj)

trends_polys <- terra::as.polygons(trends_raster.proj, dissolve = FALSE)

trends.values <- terra::extract(trends_raster.proj, trends_polys, fun = mean)
write_csv(trends.values, "../scratch/spatial_predictors/BANS_abd_trends_values_per_cell.csv")
```

### Climate Trend per Trend Cell
```{r}
## Delta climate per trend cell

file_paths <- list.files(
  path = "../scratch/spatial_predictors/delta_climate",
  pattern = "\\.tif$",
  full.names = TRUE
)

delta_clim <- lapply(file_paths, rast)

layer_names <- basename(file_paths) |>
  sub("\\.tif$", "", x = _)

for (i in 1:5) {
  delta <- delta_clim[[i]]
  delta_trend <- terra::extract(delta, trends_polys, fun = mean)
  colnames(delta_trend) <- c("ID", layer_names[i])
  write.csv(
    delta_trend,
    paste0("../scratch/spatial_predictors/delta_climate/", layer_names[i], ".csv"),
    row.names = FALSE
  )
}
```

```{r}
delta_files <- list.files(
  path = "../scratch/spatial_predictors/delta_climate",
  pattern = "_june_delta\\.csv$",
  full.names = TRUE
)

delta_list <- lapply(delta_files, read.csv)

delta_list <- lapply(delta_list, \(df) df %>% dplyr::select(-matches("^X$")))

delta_all <- reduce(delta_list, left_join, by = "ID")
write_csv(delta_all, "../scratch/spatial_predictors/delta_all_june_climate.csv")
```
## Data join
```{r}
delta_all <- read.csv("../scratch/spatial_predictors/delta_all_june_climate.csv")
trends.values <- read.csv("../scratch/spatial_predictors/BANS_abd_trends_values_per_cell.csv")

delta_abundance <- delta_all %>%
  left_join(trends.values, by = "ID")

trends.cen <- terra::centroids(trends_polys)
trends.xy <- crds(trends.cen)

sp.data.xy <- cbind(delta_abundance, trends.xy)

write_csv(sp.data.xy, "../scratch/spatial_predictors/delta_climate_BANS_abd_trend_xy.csv")
```
# RUNNING THROUGH

## trend data
```{r}
trends_raster.proj <- rast("../scratch/spatial_predictors/eBird_trends_BANS_proj4326.tif")
trends.df <- as.data.frame(trends_raster.proj, xy = TRUE)

plot(trends_raster.proj)
```
# Genoscape/ESU assignment
(Note, the following function were generated using ChatGPT based on the above written out code that I produced, interpretation of results in my own)
```{r}
## Splitting data up per ESU, this is a shape file with all four clusters
source("rasterize_genoscape_polygons.R")f
genoscape_out <- rasterize_genoscape_polygons(
  genoscape_file = "../../03.delineate_ESUs/outputs/BANS_CU_polygons_thresh45_GM6.shp",
  out_tif = "../scratch/genoscape_stars/genoscape_stars_rast_CU_GM6_threshold45.tif"
)

genoscape_rast <- genoscape_out$genoscape_rast
genoscape_terra <- terra::rast(genoscape_rast)

genoscape_rast <- genoscape_out$genoscape_rast
genoscape_terra <- terra::rast(genoscape_rast)

# inspect values actually stored in the raster
vals <- sort(unique(na.omit(terra::values(genoscape_terra)[, 1])))

# make one binary layer per ESU value
esu_layers <- lapply(vals, function(v) {
  x <- genoscape_terra[[1]]
  x <- ifel(x == v, 1, NA)
  x
})

ESU_stack <- rast(esu_layers)
names(ESU_stack) <- paste0("ESU", vals)

ESU_stack
plot(ESU_stack)

q_by_trend <- terra::extract(
  ESU_stack,
  trends_polys,
  fun = mean,
  na.rm = TRUE
)

q_by_trend[is.nan(as.matrix(q_by_trend))] <- NA

q_cols <- names(q_by_trend)[names(q_by_trend) != "ID"]

ESU_by_trend <- q_by_trend %>%
  rowwise() %>%
  mutate(
    ESU = {
      hit <- q_cols[!is.na(c_across(all_of(q_cols)))]
      if (length(hit) == 0) NA_character_ else hit[1]
    }
  ) %>%
  ungroup()

unique(ESU_by_trend$ESU)
```

## ESU to data join
```{r}
sp.data.xy <- read.csv("../scratch/spatial_predictors/delta_climate_BANS_abd_trend_xy.csv")

sp.data <- left_join(sp.data.xy, ESU_by_trend[c("ID", "ESU")], by = "ID")

sp.data.clean <- sp.data %>% tidyr::drop_na()

sp.data.clean %>%
  write.csv("../scratch/spatial_predictors/spatial_predictor_data_xy_from_ESU_GM6_threshold45.csv", row.names = FALSE)
```