---
title: "06.better_results_figures"
output: html_document
---

# LIBRARIES
```{r, message=FALSE, warning=FALSE}
library(ncdf4)
library(terra)
library(raster)
library(dplyr)
library(purrr)
library(data.table)
library(ebirdst)
library(fields)
library(rnaturalearth)
library(sf)
library(ggplot2)
library(ggrepel)
library(mgcv)
library(gratia)
library(tidyterra)
library(ggnewscale)
library(patchwork)
```

# INTRO
This is just a script from making some specific figures for talks and or final manuscript stuff. I'll update it with specific notes for the figure numbers and such based on where things end up in the manuscript...

# ESU labeling and color updates

```{r}
genoscape_polygons <- st_read("../../03.delineate_ESUs/outputs/BANS_CU_polygons_thresh45_G5.shp")

cols <- readRDS("../../03.delineate_ESUs/figures/colors/CU_grouping_colors_3_G5.rds")

genoscape_reference <- ggplot() +
  ggspatial::layer_spatial(
  genoscape_polygons,
  aes(fill  = factor(Cluster))) +
scale_fill_manual(
  values = cols,
  breaks = names(cols),
  name = "ESU"
)
genoscape_reference

# "Interior West",
#     "Mackenzie Basin",
#     "Pacific Coastal",
#     "Newfoundland",
#     "Alaska–Northwest",
#     "East Coastal"
#         "Western United States",
# "Alaska-East Coastal",
        
CU_name_ref <- tibble(
  ESU = c("ESU1", "ESU2", "ESU3", "ESU4", "ESU5"),
  Name = c(
    "Pacific Coastal",
    "Mackenzie Basin",
    "Newfoundland",
    "East Coastal",
    "Alaska-Interior West")
)

cols_fullnames <- cols
names(cols_fullnames) <- CU_name_ref$Name

saveRDS(cols_fullnames, "../colors/CU_grouping_colors_3_G5_fullnames.rds")

CU_name_ref %>% write_csv("../colors/CU_G5_name_ref_2.csv")
```

# TREND RASTER FIGURES
```{r}
trends_raster <- rast("../../04.predictors_of_decline/scratch/spatial_predictors/eBird_trends_BANS_proj4326.tif")
```

```{r, message=FALSE, warning=FALSE, results='hide'}
# get shapefiles from natural earth
coastlines <- ne_download(scale = 10, type = "coastline", category = "physical", returnclass = "sf")
countries <- ne_download(scale = 10, type = "admin_0_boundary_lines_land", category = "cultural", returnclass = "sf")
states <- ne_download(scale = 10, type = "admin_1_states_provinces_lines", category = "cultural", returnclass = "sf")
ocean <- ne_download(scale = 10, type = "ocean", category = "physical", returnclass = "sf")
lakes <- ne_download(scale = 10, type = "lakes", category = "physical", returnclass = "sf")
```

## nicer base trend map
```{r}
map <- ne_states(country = c("United States of America", "Canada"), 
                 returnclass = 'sf')
na_union <- st_union(map)

#BANS breeding range
breeding.sf <- st_read("../../../raw_data/spatial_files/banswa_range_2023.gpkg") %>% filter(season=="breeding")  %>% st_transform(crs(map))

#clip BANS range to just NA and make a single, clean polygon
breeding_na <- st_intersection(breeding.sf, na_union)
lamproj <- "+proj=lcc +lat_1=20 +lat_2=60 +lat_0=40 +lon_0=-100 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"

bb <- st_bbox(breeding_na) %>% st_transform(lamproj)

trends_projected <- terra::project(trends_raster, lamproj, method = "near")

colors <- colorRampPalette(brewer.pal(11, "RdBu"))(7)
barplot(
  rep(1, length(colors)),
  col = colors,
  border = NA,
  axes = FALSE,
  space = 0,
  names.arg = names(colors),
  las = 2
)

trends_plot <- ggplot() +
  tidyterra::geom_spatraster(data = trends_projected, na.rm = TRUE) +
  scale_fill_gradient2(
  low = "#67001F",
  mid = "white",
  high = "#053061",
  midpoint = 0,
  na.value = NA,
  name = "Trend"
)+
  geom_sf(data = breeding_na, fill = "NA", linewidth = 0.1225 , alpha = 0.6) +
  geom_sf(data = countries, fill = NA, linewidth = 0.25) +
  geom_sf(data = coastlines, fill = NA, linewidth = 0.1225) +
  geom_sf(data = states, fill = NA, linewidth = 0.1225) +
  geom_sf(data = ocean, fill = "#bddbe6") +
  geom_sf(data = lakes, fill = "#bddbe6") +
  theme(panel.border = element_rect(color = "black", fill = NA, linewidth = 1))+
  coord_sf(
  crs = st_crs(lamproj),
  xlim = c(bb["xmin"], bb["xmax"]),
  ylim = c(bb["ymin"], bb["ymax"]),
  expand = FALSE)
trends_plot
ggsave(plot=trends_plot,"../figures_output/BANS_trends_plot_ggplot.png")
```

## genoscape polygons for overlap
```{r}
genoscape_polygons <- st_read("../../03.delineate_ESUs/outputs/BANS_CU_polygons_thresh45_G5.shp")
CU_names_ref <- read.csv("../colors/CU_G5_name_ref_2.csv")

genoscape_polygons$Cluster <- factor(
  genoscape_polygons$Cluster,
  levels = CU_names_ref$ESU,
  labels = CU_names_ref$Name
)

cols <- readRDS("../colors/CU_grouping_colors_3_G5_fullnames.rds")

out_overlap_plot <- paste0("../figures_output/trends/BANS_trends_plot_ESU_G5_overlay_ggplot_colored_2.png")
```

```{r}
map_layers_rasterized <- ggrastr::rasterise(
  {
    list(
      geom_sf(data = breeding_na, fill = NA, linewidth = 0.1225, alpha = 0.6),
      geom_sf(data = countries, fill = NA, linewidth = 0.25),
      geom_sf(data = coastlines, fill = NA, linewidth = 0.1225),
      geom_sf(data = states, fill = NA, linewidth = 0.1225),
      geom_sf(data = ocean, fill = "#bddbe6"),
      geom_sf(data = lakes, fill = "#bddbe6"),
      ggspatial::layer_spatial(
        genoscape_polygons,
        aes(fill = factor(Cluster)),
        alpha = 0.25
      )
    )
  },
  dpi = 300
)

trends_plot_overlaid <- ggplot() +
  tidyterra::geom_spatraster(data = trends_projected, na.rm = TRUE) +
  scale_fill_gradient2(
    low = "#CA4741",
    mid = "white",
    high = "#3784BB",
    midpoint = 0,
    na.value = NA,
    name = "Trend"
  ) +
  ggnewscale::new_scale_fill() +
  map_layers_rasterized +
  scale_fill_manual(
    values = cols,
    breaks = names(cols),
    name = "ESU"
  ) +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
  ) +
  coord_sf(
    crs = st_crs(lamproj),
    xlim = c(bb["xmin"], bb["xmax"]),
    ylim = c(bb["ymin"], bb["ymax"]),
    expand = FALSE
  )
trends_plot_overlaid
ggsave(plot=trends_plot_overlaid, out_overlap_plot)

pdf("../figures_output/BANS_trends_plot_ESU_G5_overlay_ggplot_colored.pdf", height=4, width=8)
print(trends_plot_overlaid)
dev.off()
```

```{r}
map_only <- ggplot() +
  geom_sf(data = ocean, fill = "#bddbe6", color = NA) +
  geom_sf(data = countries, fill = "lightgray", color = NA) +
  geom_sf(data = states, fill = "lightgray", color = NA) +
  geom_sf(data = lakes, fill = "#bddbe6", color = NA) +

  tidyterra::geom_spatraster(data = trends_projected, na.rm = TRUE) +
  scale_fill_gradient2(
    low = "#CA4741",
    mid = "white",
    high = "#3784BB",
    midpoint = 0,
    na.value = NA,
    name = "Trend"
  ) +

  geom_sf(data = breeding_na, fill = NA, linewidth = 0.1225, alpha = 0.6) +
  geom_sf(data = countries, fill = NA, linewidth = 0.25) +
  geom_sf(data = coastlines, fill = NA, linewidth = 0.1225) +
  geom_sf(data = states, fill = NA, linewidth = 0.1225) +

  ggnewscale::new_scale_fill() +
  ggspatial::layer_spatial(
    genoscape_polygons,
    aes(fill = factor(Cluster)),
    alpha = 0.25
  ) +
  scale_fill_manual(
    values = cols,
    breaks = names(cols),
    name = "ESU",
    na.value = NA
  ) +
  coord_sf(
    crs = st_crs(lamproj),
    xlim = c(bb["xmin"], bb["xmax"]),
    ylim = c(bb["ymin"], bb["ymax"]),
    expand = FALSE
  ) +
  theme_void() +
  theme(
    legend.position = "none",
    plot.margin = margin(0, 0, 0, 0),
    panel.background = element_rect(fill = "gray90", color = NA),
    plot.background = element_rect(fill = "gray90", color = NA)
  )

ggsave(
  "../figures_output/trends_map_only.png",
  map_only,
  width = 10,
  height = 6,
  dpi = 600,
  bg = NA
)
```

# MODEL FIGURES
(Note, functions following were based on code I wrote and used ChatGPT to make into full functions, all interpretations are my own)
## partial effect plots
```{r}
CU_names_ref <- read.csv("../colors/CU_G5_name_ref_2.csv")
cols <- readRDS("../colors/CU_grouping_colors_3_G5_fullnames.rds")
K <- length(cols)

data <- read.csv("../../04.predictors_of_decline/scratch/spatial_predictors/spatial_predictor_data_xy_from_ESU_G5_threshold45.csv")

clean.data <- data %>% filter(!is.na(ESU))

clean.data.names <- left_join(clean.data, CU_names_ref, by="ESU")

clim_vars <- c("ppt", "tmax", "soil", "tmin", "ws")

for (var in clim_vars){
  modGS <- readRDS(paste0("../../04.predictors_of_decline/outputs/GAM_outputs/function_out/",scheme,"/modGS_",var,".rds"))
  sm_dat <- smooth_estimates(modGS,
                           smooth = paste0("s(",var,"_june_delta,ESU)")
                           )
  sm_dat.names <- left_join(sm_dat, CU_names_ref, by="ESU")
  
  p <- ggplot(sm_dat.names, aes(x = .data[[paste0(var, "_june_delta")]],
                   y = .estimate,
                   colour = Name)) +
  geom_line() +
  facet_wrap(~ Name, nrow=1) +
  scale_colour_manual(values = cols) +
  theme_classic()+
  theme(
  strip.text = element_blank(),
  strip.background = element_blank(),
  legend.position = "none",
  axis.title.y = element_text(size = 7),
  axis.text.x = element_text(size=8, 
                             angle =90,
                             vjust = 0.5,
                             hjust = 1))+
  geom_hline(yintercept = 0, linetype = "dashed")+
  geom_vline(xintercept = 0, linetype = "dashed")+
  xlab(paste0("Change in ",var))+
  ylab(expression("Partial effect on " ~ Delta ~ "Abundance"))
  
ggsave(
    filename = paste0(
      "../figures_output/partial_effects/",scheme,"/BANS_",
      var, "_GAM_", scheme, "_partialeffect.png"
    ),
    plot = p,
    height = 2,
    width = 6
    )
}

```

```{r}
source("plot_partial_effects_by_scheme.R")

CU_names_ref <- read.csv("../colors/CU_G5_name_ref_2.csv")
cols <- readRDS("../colors/CU_grouping_colors_3_G5_fullnames.rds")

# run for both modGS and modS
plot_partial_effects_by_scheme(
  schemes = "G5",
  model = "modGS",
  cols = cols,
  CU_names_ref = CU_names_ref,
  clim_vars = c("ppt", "tmax", "soil", "tmin", "ws")
)
```

## prediction plots
### modG
```{r}
source("plot_modG_predictions_all_schemes.R")
pred_out <- plot_modG_predictions_all_schemes(
  schemes = "G5",
  focal_vars = c("tmax_june_delta", "ppt_june_delta", "soil_june_delta", "ws_june_delta", "tmin_june_delta"),
  cols = cols,
  CU_names_ref = CU_names_ref,
  plot_dims = list(width = 7, height = 4),
  x_labels = c(
    tmax_june_delta = "Change in Max Temperature",
    ppt_june_delta = "Change in Precipitation",
    soil_june_delta = "Change in Soil Moisture",
    ws_june_delta = "Change in Wind Speed",
    tmin_june_delta = "Change in Min Temperature"
  ),
  include_pop_line = TRUE
)

pred_out$G5$results_by_var$tmax_june_delta$plot
```


```{r}
schemes <- names(pred_out)
schemes <- "G5"
for (scheme in schemes) {

  # grab all plot objects for this scheme
  plot_list <- lapply(pred_out[[scheme]]$results_by_var, function(x) x$plot)

  # combine into one figure and collect a single legend
  p_combined <- wrap_plots(plot_list, ncol = 5, guides = "collect") &
    theme(legend.position = "bottom")

  # save into the corresponding scheme folder
  ggsave(
    filename = file.path(
      "../figures_output/modG_predictions",
      scheme,
      paste0("modG_allvars_predictions_", scheme, ".png")
    ),
    plot = p_combined,
    width = 25,
    height = 8
  )
}


```

### modGS
```{r}
source("plot_modS_modGS_predictions_all_schemes.R")

pred_out_modGS <- plot_modS_modGS_predictions_all_schemes(
  schemes = "G5",
  model = "modGS",
  cols = cols,
  CU_names_ref = CU_names_ref,
  focal_vars = c("tmax_june_delta", "ppt_june_delta", "soil_june_delta", "ws_june_delta", "tmin_june_delta"),
  plot_dims = list(width = 7, height = 4),
  x_labels = c(
    tmax_june_delta = "Change in Max Temperature",
    ppt_june_delta = "Change in Precipitation",
    soil_june_delta = "Change in Soil Moisture",
    ws_june_delta = "Change in Wind Speed",
    tmin_june_delta = "Change in Min Temperature"
  )
)

schemes <- names(pred_out_modGS)
clim_vars <- c("tmax_june_delta", "ppt_june_delta", "soil_june_delta", "ws_june_delta", "tmin_june_delta")

for (scheme in schemes) {

  plot_list <- lapply(clim_vars, function(v) {
    pred_out_modGS[[scheme]]$results_by_var[[v]]$plot
  })

  p_combined <- patchwork::wrap_plots(plot_list, ncol = 5, guides = "collect") &
    ggplot2::theme(legend.position = "bottom")

  ggplot2::ggsave(
    filename = file.path(
      "../figures_output/modGS_predictions",
      scheme,
      paste0("modGS_allvars_predictions_", scheme, ".png")
    ),
    plot = p_combined,
    width = 25,
    height = 8
  )
}

scheme="G5"
  
 plot_list <- lapply(clim_vars, function(v) {
    pred_out_modGS[[scheme]]$results_by_var[[v]]$plot
  })

  p_combined <- patchwork::wrap_plots(plot_list, ncol = 5, guides = "collect") &
    ggplot2::theme(legend.position = "bottom")
  
pdf(
    file = file.path(
      "../figures_output/modGS_predictions",
      scheme,
      paste0("modGS_allvars_predictions_", scheme, ".pdf")
    ),
    width = 25,
    height = 8
  )
print(p_combined)
dev.off()
```
#### per env variables
```{r}
data <- read.csv("../../04.predictors_of_decline/scratch/spatial_predictors/spatial_predictor_data_xy_from_AU_K3_threshold55.csv")

#only data where I had a good ESU assignment
clean.data <- data %>% filter(!is.na(ESU))
nrow(clean.data)

clean.data$ESU <- factor(clean.data$ESU)

CU_names_ref <- data.frame(
  ESU = c("AU1", "AU2", "AU3"),
  Name = c(
    "Western",
    "Eastern",
    "Northern"
  )
)

table(clean.data$ESU)
clean.data$ESU_label <- factor(
  clean.data$ESU,
  levels = CU_names_ref$ESU,
  labels = CU_names_ref$Name
  )
```
##### TMIN
```{r}
modGS.tmin <- readRDS("../../04.predictors_of_decline/outputs/GAM_outputs/function_out/AU_K3/modGS_tmin.rds")
```

```{r}
pred_grid <- expand.grid(
  tmin_june_delta = seq(
    min(clean.data$tmin_june_delta, na.rm = TRUE),
    max(clean.data$tmin_june_delta, na.rm = TRUE),
    length.out = 100
  ),
  ESU = levels(clean.data$ESU)
)

cols <- readRDS("../../06.final_figures/colors/AU_grouping_colors_K3.rds")
#names(cols) <- CU_names_ref$Name

pred_grid$ppt_june_delta  <- median(clean.data$ppt_june_delta,  na.rm = TRUE)
pred_grid$soil_june_delta <- median(clean.data$soil_june_delta, na.rm = TRUE)
pred_grid$tmax_june_delta <- median(clean.data$tmax_june_delta, na.rm = TRUE)
pred_grid$ws_june_delta <- median(clean.data$ws_june_delta, na.rm = TRUE)

p_full <- predict(modGS.tmin, newdata = pred_grid, se.fit = TRUE)
pred_grid$fit_full <- p_full$fit
pred_grid$se_full  <- p_full$se.fit

pred_plot <- ggplot(pred_grid, aes(x = tmin_june_delta, y = fit_full, color = ESU, group = ESU)) +
  geom_line() +
  geom_ribbon(aes(ymin = fit_full - 2*se_full, ymax = fit_full + 2*se_full, fill = ESU),
              alpha = 0.15, colour = NA) +
  theme_classic() +
  guides(fill = "none") +
  ylab("Predicted Abd Trend")+
  xlab(expression("Change in" ~T[min]))+
  facet_wrap(~ESU, scales = "free", ncol = 5) +
  scale_fill_manual(values = cols,
                    breaks = names(cols),
                    name = "ESU")+
  theme(
  strip.text = element_blank(),
  axis.text.x = element_text(size=8, 
                             angle =90,
                             vjust = 0.5,
                             hjust = 1),
  legend.position = "none")+
  scale_color_manual(values = cols)

pred_plot
ggsave("../figures_output/BANS_tminGAM_allAU_K3_predictedtrend.png", height=2, width=6)
pdf("../figures_output/BANS_tminGAM_allAU_K3_predictedtrend.pdf", width = 8, height = 2)
print(pred_plot)
dev.off()
```

##### TMAN
```{r}
modGS.tmax <- readRDS("../../04.predictors_of_decline/outputs/GAM_outputs/function_out/AU_K3/modGS_tmax.rds")
```

```{r}
pred_grid <- expand.grid(
  tmax_june_delta = seq(
    min(clean.data$tmax_june_delta, na.rm = TRUE),
    max(clean.data$tmax_june_delta, na.rm = TRUE),
    length.out = 100
  ),
  ESU = levels(clean.data$ESU)
)

cols <- readRDS("../../06.final_figures/colors/AU_grouping_colors_K3.rds")
#names(cols) <- CU_names_ref$Name

pred_grid$ppt_june_delta  <- median(clean.data$ppt_june_delta,  na.rm = TRUE)
pred_grid$soil_june_delta <- median(clean.data$soil_june_delta, na.rm = TRUE)
pred_grid$tmin_june_delta <- median(clean.data$tmin_june_delta, na.rm = TRUE)
pred_grid$ws_june_delta <- median(clean.data$ws_june_delta, na.rm = TRUE)

p_full <- predict(modGS.tmax, newdata = pred_grid, se.fit = TRUE)
pred_grid$fit_full <- p_full$fit
pred_grid$se_full  <- p_full$se.fit

pred_plot <- ggplot(pred_grid, aes(x = tmax_june_delta, y = fit_full, color = ESU, group = ESU)) +
  geom_line() +
  geom_ribbon(aes(ymin = fit_full - 2*se_full, ymax = fit_full + 2*se_full, fill = ESU),
              alpha = 0.15, colour = NA) +
  theme_classic() +
  guides(fill = "none") +
  ylab("Predicted Abd Trend")+
  xlab(expression("Change in" ~T[max]))+
  facet_wrap(~ESU, scales = "free", ncol = 5) +
  scale_fill_manual(values = cols,
                    breaks = names(cols),
                    name = "ESU")+
  theme(
  strip.text = element_blank(),
  axis.text.x = element_text(size=8, 
                             angle =90,
                             vjust = 0.5,
                             hjust = 1),
  legend.position = "none")+
  scale_color_manual(values = cols)

pred_plot
ggsave("../figures_output/BANS_tmaxGAM_allAU_K3_predictedtrend.png", height=2, width=6)
pdf("../figures_output/BANS_tmaxGAM_allAU_K3_predictedtrend.pdf", width = 8, height = 2)
print(pred_plot)
dev.off()
```

##### PRECIP
```{r}
modGS.ppt <- readRDS("../../04.predictors_of_decline/outputs/GAM_outputs/function_out/G5/modGS_ppt.rds")
```

```{r}
pred_grid <- expand.grid(
  ppt_june_delta = seq(
    min(clean.data$ppt_june_delta, na.rm = TRUE),
    max(clean.data$ppt_june_delta, na.rm = TRUE),
    length.out = 100
  ),
  ESU = levels(clean.data$ESU)
)

cols <- readRDS("../../03.delineate_ESUs/figures/colors/CU_grouping_colors_3_G5.rds")

pred_grid$tmax_june_delta  <- median(clean.data$tmax_june_delta,  na.rm = TRUE)
pred_grid$soil_june_delta <- median(clean.data$soil_june_delta, na.rm = TRUE)
pred_grid$tmin_june_delta <- median(clean.data$tmin_june_delta, na.rm = TRUE)
pred_grid$ws_june_delta <- median(clean.data$ws_june_delta, na.rm = TRUE)

p_full <- predict(modGS.ppt, newdata = pred_grid, se.fit = TRUE)
pred_grid$fit_full <- p_full$fit
pred_grid$se_full  <- p_full$se.fit

ggplot(pred_grid, aes(x = ppt_june_delta, y = fit_full, color = ESU, group = ESU)) +
  geom_line() +
  geom_ribbon(aes(ymin = fit_full - 2*se_full, ymax = fit_full + 2*se_full, fill = ESU),
              alpha = 0.15, colour = NA) +
  theme_classic() +
  guides(fill = "none") +
  ylab("Predicted Abd Trend")+
  xlab(expression("Change in" ~ppt))+
  facet_wrap(~ESU, scales = "free", ncol = 5) +
  scale_fill_manual(values = cols,
                    breaks = names(cols),
                    name = "ESU")+
  theme(
  strip.text = element_blank(),
  axis.text.x = element_text(size=8, 
                             angle =90,
                             vjust = 0.5,
                             hjust = 1),
  legend.position = "none")+
  scale_color_manual(values = cols)

  
ggsave("../figures_output/BANS_pptGAM_allG5_predictedtrend.png", height=2, width=6)
```

### modS
```{r}
source("plot_modS_modGS_predictions_all_schemes.R")

pred_out_modS <- plot_modS_modGS_predictions_all_schemes(
  schemes = "G5",
  model = "modS",
  focal_vars = c("tmax_june_delta", "ppt_june_delta", "soil_june_delta", "ws_june_delta", "tmin_june_delta"),
  plot_dims = list(width = 7, height = 4),
  x_labels = c(
    tmax_june_delta = "Change in Max Temperature",
    ppt_june_delta = "Change in Precipitation",
    soil_june_delta = "Change in Soil Moisture",
    ws_june_delta = "Change in Wind Speed",
    tmin_june_delta = "Change in Min Temperature"
  )
)

schemes <- names(pred_out_modS)
clim_vars <- c("tmax_june_delta", "ppt_june_delta", "soil_june_delta", "ws_june_delta", "tmin_june_delta")

for (scheme in schemes) {
  
  plot_list <- lapply(clim_vars, function(v) {
    pred_out_modS[[scheme]]$results_by_var[[v]]$plot
  })
  
  p_combined <- patchwork::wrap_plots(plot_list, ncol = 5, guides = "collect") &
    ggplot2::theme(legend.position = "bottom")
  
  ggplot2::ggsave(
    filename = file.path(
      "../figures_output/modS_predictions",
      scheme,
      paste0("modS_allvars_predictions_", scheme, ".png")
    ),
    plot = p_combined,
    width = 25,
    height = 8
  )
}
```

# GENOMIC OFFSET FIGURES
## GO PER ESU Plots
```{r}
library(sf)
RDA_proj_offset <- readRDS("RDA_proj_offset_all_loci_table.rds")
ESU6_polygons <- st_read("../05.predictors_of_decline/BANS_ESU_K6_polygons_update_thresh45.shp")
ESU6_polygons$Cluster <- factor(
  ESU6_polygons$Cluster,
  levels = c("ESU1","ESU2","ESU3","ESU4","ESU5","ESU6"),
  labels = c(
    "Interior West",
    "Mackenzie Basin",
    "Pacific Coastal",
    "Newfoundland",
    "Alaska–Northwest",
    "East Coastal"
  )
)

pts <- st_as_sf(
  RDA_proj_offset,
  coords = c("x", "y"),
  crs = 4326,
  remove = FALSE
)

esu <- st_transform(ESU6_polygons, 4326)

# 3) join: assigns each point an ESU id
pts_esu <- st_join(pts, esu[, c("Cluster")], join = st_within, left = FALSE)
saveRDS(pts_esu, "pts_ESU6.rds")

# 4) summarize to ESU-level offset
esu_offset <- pts_esu %>%
  st_drop_geometry() %>%
  group_by(Cluster, Date) %>%
  summarize(
    n_cells = n(),
    mean_offset   = mean(Global_offset, na.rm = TRUE),
    median_offset = median(Global_offset, na.rm = TRUE),
    p05 = quantile(Global_offset, 0.05, na.rm = TRUE),
    p95 = quantile(Global_offset, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

esu_offset
saveRDS(esu_offset, "ESU6_offset_stats.rds")
```

```{r}
library(dplyr)
library(ggplot2)
library(ggdist)

names(cols) <- c(
  "Interior West",
  "Mackenzie Basin",
  "Pacific Coastal",
  "Newfoundland",
  "Alaska–Northwest",
  "East Coastal"
) 
# per-cell offsets (NOT summarized)
cells <- readRDS("pts_ESU6.rds") %>%
  st_drop_geometry() %>%
  mutate(Cluster = factor(Cluster, levels = names(cols)))  # keep your ESU order

# if you want only a specific future period, filter Date here:
# cells <- cells %>% filter(Date == "2068-2088")  # or whatever your Date values are
set.seed(5)  # for reproducibility

cells_subsample_1m <- cells[
  sample.int(nrow(cells), 1e6),
]

raincloud <- ggplot(cells, aes(x = Cluster, y = Global_offset, color = Cluster, fill = Cluster)) +
  # half density "trace" (one-sided)
  stat_halfeye(
    side = "right",
    adjust = 2,          # try 1.2–2.5
    width = 1.3,
    alpha = 0.75,
    .width = 0,            # <-- removes interval/box-looking part
    point_colour = NA,
    position = position_nudge(x = 0.1)  # <-- shifts halfeye right
  ) +
  # boxplot on top
  geom_boxplot(
    width = 0.18,
    outlier.shape = NA,
    alpha = 0.75,
    color = "black",
    position = position_nudge(x = -0.05),
    linewidth=0.4
  ) +
  scale_fill_manual(values = cols) +
  scale_color_manual(values = cols) +
  labs(
    title = "Within-ESU Distribution of Genomic Offset | 2071–2100",
    x = NULL,
    y = "Genomic Offset"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 8, angle = 90, hjust = 1),
    legend.position = "none"
  )

#raincloud
ggsave(plot=raincloud, "results/GO_per_ESU_raincloud.png", width = 6, height = 4.5, dpi = 300)

######### MORE SPACE VERSION
esu_order <- c(
  "Interior West",
  "Pacific Coastal",
  "Newfoundland",
  "Alaska–Northwest",
  "East Coastal",
  "Mackenzie Basin"
)

cells <- cells %>%
  mutate(Cluster = factor(Cluster, levels = esu_order))

# 2) optional: per-ESU "caps" so the spiky ones don't spill
halfeye_widths <- c(
  "Interior West"   = .5,
  "Pacific Coastal" = .5,
  "Newfoundland"    = .5,   # shrink these two
  "Alaska–Northwest"= .5,
  "East Coastal"    = .5,
  "Mackenzie Basin" = .5
)

p <- ggplot(cells, aes(x = Cluster, y = Global_offset, fill = Cluster)) +
  scale_fill_manual(values = cols) +
  scale_x_discrete(expand = expansion(mult = c(0.02, 0.02))) +
  labs(
    title = "Within-ESU Distribution of Genomic Offset | 2071–2100",
    x = NULL,
    y = "Genomic Offset"
  ) +
  theme_classic() +
  theme(
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none"
  )
for (esu in esu_order) {
  d_esu <- filter(cells, Cluster == esu)

  p <- p +
    stat_halfeye(
      data = d_esu,
      side = "right",
      adjust = 2,
      width = unname(halfeye_widths[esu]),
      alpha = 0.75,
      .width = 0,
      point_colour = NA,
      position = position_nudge(x = 0.30)
    ) +
    geom_boxplot(
      data = d_esu,
      width = 0.35,
      outlier.shape = NA,
      alpha = 0.75,
      color = "black",
      linewidth = 0.4,
      position = position_nudge(x = 0.10)
    )
}

p


ggsave(plot=p, "results/GO_per_ESU_raincloud_wide.png", width = 8, height = 2)
```

```{r}
esu_offset <- readRDS("ESU6_offset_stats.rds")
esu_long <- esu_offset %>%
  pivot_longer(
    cols = c(mean_offset, median_offset),
    names_to = "Metric",
    values_to = "Offset"
  )


cols <- readRDS("../02.5.delineate_ESUs/figures/colors/pcadmix_ESU_6_colors_2.rds")
names(cols) <- c(
  "Interior West",
  "Mackenzie Basin",
  "Pacific Coastal",
  "Newfoundland",
  "Alaska–Northwest",
  "East Coastal"
) 

ggplot(esu_offset, aes(x = Cluster, y = mean_offset, color=Cluster)) +
  geom_linerange(aes(ymin = p05, ymax = p95),
                 linewidth = 1.1,
                 color = "gray50") +
  geom_point(size = 3) +
  theme_bw(base_size = 12) +
  scale_color_manual(values=cols)+
  labs(
    title = "Within-ESU Distribution of Genomic Offset | 2071-2100",
    subtitle = "Points = mean, bars = 5th–95th percentile",
    y = "Genomic Offset",
    x = "ESU"
  )+
  theme(axis.text.x = element_text(size=8, 
                             angle =45,
                             hjust = 1),
        legend.position = "none")
ggsave("results/GO_per_ESU_mean_bar.png")
```

```{r}
ggplot(
  pts_esu,
  aes(x = Cluster, y = Global_offset)
) +
  geom_violin(outlier.size = 0.2) +
  facet_wrap(~ Date) +
  theme_bw(base_size = 12) +
  labs(
    title = "Distribution of Genomic Offset by ESU",
    y = "Genomic Offset",
    x = "Evolutionarily Significant Unit"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid = element_blank(),
    legend.position = "none"
  )
ggsave("results/GO_per_ESU_mean_bar.png")
```

# NEW MIGRATION MAP
```{r}
map <- ne_states(
  country = c("United States of America", "Canada", "Mexico", "Colombia"),
  returnclass = "sf"
)

na_union <- st_union(map)

breeding.sf <- st_read("../../../raw_data/spatial_files/banswa_range_2023.gpkg") %>%
  filter(season == "breeding") %>%
  st_transform(st_crs(map))

breeding_na <- st_intersection(breeding.sf, na_union)

lamproj <- "+proj=lcc +lat_1=20 +lat_2=60 +lat_0=40 +lon_0=-100 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"

bb <- st_bbox(breeding_na) %>%
  st_as_sfc() %>%
  st_transform(lamproj) %>%
  st_bbox()

# expand southward to include northern South America
bb["ymin"] <- bb["ymin"] - 3000000
bb["xmin"] <- bb["xmin"] - 500000
bb["xmax"] <- bb["xmax"] + 500000

coastlines <- ne_download(scale = 10, type = "coastline", category = "physical", returnclass = "sf")
countries <- ne_download(scale = 10, type = "admin_0_boundary_lines_land", category = "cultural", returnclass = "sf")
states <- ne_download(scale = 10, type = "admin_1_states_provinces_lines", category = "cultural", returnclass = "sf")
ocean <- ne_download(scale = 10, type = "ocean", category = "physical", returnclass = "sf")
lakes <- ne_download(scale = 10, type = "lakes", category = "physical", returnclass = "sf")

ggplot() +
  geom_sf(data = countries, fill = NA, linewidth = 0.25) +
  geom_sf(data = coastlines, fill = NA, linewidth = 0.1225) +
  geom_sf(data = states, fill = NA, linewidth = 0.1225) +
  geom_sf(data = ocean, fill = "#bddbe6") +
  geom_sf(data = lakes, fill = "#bddbe6")+
  coord_sf(
    crs = st_crs(lamproj),
    xlim = c(bb["xmin"], bb["xmax"]),
    ylim = c(bb["ymin"], bb["ymax"]),
    expand = FALSE) +
  theme(panel.border = element_rect(color = "black", fill = NA, linewidth = 1))
ggsave("../figures_output/plain_base_map.png")

```

# UPDATED ABUNDANCE TREND FIGURES
```{r}
data <- read.csv("../../04.predictors_of_decline/scratch/spatial_predictors/spatial_predictor_data_xy_from_AU_K3_threshold4.csv")

clean.data <- data %>% filter(!is.na(ESU))
nrow(clean.data)

clean.data$ESU <- factor(clean.data$ESU)

table(clean.data$ESU)

cols <- readRDS("../colors/AU_grouping_colors_K3.rds")

CU_names_ref <- data.frame(
  ESU = c("AU1", "AU2", "AU3"),
  Name = c(
    "Western",
    "Eastern",
    "Northern"
  )
)

K <- length(cols)

clean.data <- data %>% filter(!is.na(ESU))

clean.data.names <- left_join(clean.data, CU_names_ref, by="ESU")

abd_trend <- ggplot(clean.data.names, aes(x = Name, y = abd_trend, fill = ESU, color = ESU)) +
  geom_violin(trim = FALSE, alpha = 0.5) +
  geom_boxplot(width = 0.1, outlier.shape = NA, alpha = 0.5) +
  scale_fill_manual(values = cols) +
  scale_color_manual(values = cols) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_classic() +
  theme(legend.position = "none")

ggsave(plot=abd_trend, "../figures_output/updated_abd_trend_AU_K3_violin.png", width = 6, height=3)

pdf("../figures_output/updated_abd_trend_G5_violin.pdf", width=6, height=3)
print(abd_trend)
dev.off()
```

## UPDATED ADMIX PLOT
```{r}
library(data.table)
# which K value do we want to pull?
K_target <- 5

# read in the Q files and pca files
Q_tibble <- read.table("../../03.delineate_ESUs/scratch/pcadmix_output/BANS.filtered.unrel_pruned.alpha0.e4.run1.admix.5.Q") %>%
  setNames(paste0("ESU", 1:5))
pca_df <- read.table("../../03.delineate_ESUs/scratch/PCA/BANS.filtered.unrel.ld25-10-0.5.eigenvec", header = FALSE) %>% dplyr::select(-V2)

colnames(pca_df)[-1] <- paste0("PC", 1:(ncol(pca_df) - 1))
names(pca_df)[1] <- "ind"

Q_tibble$ind <- pca_df$ind

# colors from before, just renamed
component_colors <- readRDS("../../03.delineate_ESUs/figures/colors/CU_grouping_colors_2_G5.rds")

# Cluster membership and letter
pops <- read.csv("../../03.delineate_ESUs/scratch/pops.csv") %>%
  transmute(ind = BGP_ID, Group)

letter_key <- read.csv("../../03.delineate_ESUs/scratch/cluster_letter_key_geo_lonlat.csv")

pops.cluster.letter <- left_join(pops, letter_key, by="Group")

Q_tibble2 <- Q_tibble %>%
  left_join(pops.cluster.letter, by = "ind") %>%
  filter(!is.na(cluster_letter))

# Long format for ggplot
desired_order <- c("A", "B", "C", "E", "H", "J", "L", "O", "P", "Q", "R", "S", 
                   "T", "U", 
                   "G", "I", "K", "M", "N",
                   "D", "F")

ind_order <- Q_tibble2 %>%
  mutate(
    cluster_letter = factor(cluster_letter, levels = desired_order)
  ) %>%
  arrange(cluster_letter, Group, ind) %>%
  pull(ind) %>%
  unique()

q_long <- Q_tibble2 %>%
  mutate(
    cluster_letter = factor(cluster_letter, levels = unique(letter_key$cluster_letter)),
    ind = factor(ind, levels = ind_order)
  ) %>%
  pivot_longer(
    cols = starts_with("ESU"),
    names_to = "popGroup",
    values_to = "prob"
  ) %>%
  mutate(
    popGroup = factor(popGroup, levels = paste0("ESU", 1:K_target))
  )

# Plot
K_plot <- ggplot(q_long, aes(x = ind, y = prob, fill = popGroup)) +
  geom_col(color = "gray", linewidth = 0.1) +
  facet_grid(~cluster_letter, switch = "x", scales = "free", space = "free") +
  theme_minimal() +
  labs(x = "", y = "Ancestry", title = "") +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_discrete(expand = expansion(add = 1)) +
  theme(
    panel.spacing.x = unit(0.001, "lines"),
    axis.text.x = element_blank(),
    strip.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5, size = 8),
    panel.grid = element_blank()
  ) +
  scale_fill_manual(values = component_colors, guide = "none")

K_plot

out_plot <- "../../03.delineate_ESUs/figures/admix_plots/BANS_admix_overall_alpha0_G4_letters_reordered.png"

ggsave(out_plot, K_plot, width = 10, height = 2.2, units = "in", dpi = 300, bg = "white")

pops %>% count(Group)
```

# OVERALL PCA OUTPUT
```{r}
plink_pca <- read.table("../../03.delineate_ESUs/scratch/PCA/BANS.filtered.unrel.ld25-10-0.2.eigenvec", header = FALSE) %>% dplyr::select(-V2)
eigenval <- scan("../../03.delineate_ESUs/scratch/PCA/BANS.filtered.unrel.ld25-10-0.2.eigenval")

pve05 <- data.frame(PC = 1:199, pve = eigenval/sum(eigenval)*100)

# set names
pca05 <- plink_pca
colnames(pca05)[-1] <- paste0("PC", 1:(ncol(pca05) - 1))
names(pca05)[1] <- "ind"
nrow(pca05)
#correct number of samples

bg <- readRDS("../../03.delineate_ESUs/figures/colors/bg_colors_new.rds")

pops <- read.csv("../../03.delineate_ESUs/scratch/pops.csv") %>% filter(BGP_ID %in% pca05$ind)

# Join PCA data
pca.cluster <- left_join(pca05, pops %>% rename(ind=BGP_ID), by="ind")  # left_join PCA first to preserve PC columns

samp_unrel  <-  read.table("../../03.delineate_ESUs/scratch/BANS_unrel_sample_list_short_2.txt", header = F)
pca.cluster.norel <- pca.cluster %>% filter(ind %in% samp_unrel$V1)
nrow(pca.cluster.norel)

geo_key <- readRDS("../../03.delineate_ESUs/scratch/geo_informative_labels.RDS")
region_order <- c(
  "Cluster_1", "Cluster_5", "Cluster_17", "Cluster_4", "Cluster_8", "Cluster_13", "Cluster_2", "Cluster_21",
  "Cluster_3", "Cluster_7", "Cluster_18", "Cluster_19", "Cluster_22", "Cluster_20",
  "Cluster_12", "Cluster_15", "Cluster_14", "Cluster_10", "Cluster_9", "Cluster_11", "Cluster_6", "Cluster_16"
)

pca.cluster.norel <- pca.cluster.norel %>%
  left_join(geo_key, by = "Group") %>%
  mutate(
    Group = factor(Group, levels = region_order)
  )

# make one label per Group, in the same order
group_labels <- pca.cluster.norel %>%
  distinct(Group, Group_label) %>%
  arrange(Group)

label_vec <- setNames(group_labels$Group_label, group_labels$Group)

# make sure colors are in the same order
bg_use <- bg[region_order]

overall.pca <- ggplot(
  pca.cluster.norel,
  aes(
    x = PC1,
    y = PC2,
    color = Group,
    text = paste0(
      "ID: ", ind, "<br>",
      "Location: ", Group_label, "<br>",
      "PC1: ", round(PC1, 4), "<br>",
      "PC2: ", round(PC2, 4)
    )
  )
) +
  geom_point(size = 3) +
  scale_color_manual(
    values = bg_use,
    breaks = region_order,
    labels = label_vec[region_order],
    name = "Location"
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = gray(0.8)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = gray(0.8)) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(hjust = 0.5),
    legend.position = "right"
  )+
  xlab(paste0("PC1 (", signif(pve05$pve[1], 3), "%)")) +
  ylab(paste0("PC2 (", signif(pve05$pve[2], 3), "%)"))

overall.pca

pdf("../figures_output/BANS_overall_ld0.2_PCA.pdf", width = 8, height = 6)
print(overall.pca)
dev.off()
```

```{r}
plink_pca <- read.table("../../07.adaptive_analysis/scratch/PCA_eigen/BANS.env_only.outliers_z3_list.eigenvec", header = FALSE) %>% dplyr::select(-V2)
eigenval <- scan("../../07.adaptive_analysis/scratch/PCA_eigen/BANS.env_only.outliers_z3_list.eigenval")

pve05 <- data.frame(PC = 1:199, pve = eigenval/sum(eigenval)*100)
ggplot(pve05[1:20, ], aes(x = PC, y = pve)) +
  geom_col(width = 0.9) +
  labs(x = "PC", y = "PVE (%)") +
  theme_bw()

# set names
pca05 <- plink_pca
colnames(pca05)[-1] <- paste0("PC", 1:(ncol(pca05) - 1))
names(pca05)[1] <- "ind"
nrow(pca05)
#correct number of samples

bg <- readRDS("../../03.delineate_ESUs/figures/colors/bg_colors_new.rds")

pops <- read.csv("../../03.delineate_ESUs/scratch/pops.csv") %>% filter(BGP_ID %in% pca05$ind)

# Join PCA data
pca.cluster <- left_join(pca05, pops %>% rename(ind=BGP_ID), by="ind")  # left_join PCA first to preserve PC columns

samp_unrel  <-  read.table("../../03.delineate_ESUs/scratch/BANS_unrel_sample_list_short_2.txt", header = F)
pca.cluster.norel <- pca.cluster %>% filter(ind %in% samp_unrel$V1)
nrow(pca.cluster.norel)

geo_key <- readRDS("../../03.delineate_ESUs/scratch/geo_informative_labels.RDS")
region_order <- c(
  "Cluster_1", "Cluster_5", "Cluster_17", "Cluster_4", "Cluster_8", "Cluster_13", "Cluster_2", "Cluster_21",
  "Cluster_3", "Cluster_7", "Cluster_18", "Cluster_19", "Cluster_22", "Cluster_20",
  "Cluster_12", "Cluster_15", "Cluster_14", "Cluster_10", "Cluster_9", "Cluster_11", "Cluster_6", "Cluster_16"
)

pca.cluster.norel <- pca.cluster.norel %>%
  left_join(geo_key, by = "Group") %>%
  mutate(
    Group = factor(Group, levels = region_order)
  )

# make one label per Group, in the same order
group_labels <- pca.cluster.norel %>%
  distinct(Group, Group_label) %>%
  arrange(Group)

label_vec <- setNames(group_labels$Group_label, group_labels$Group)

# make sure colors are in the same order
bg_use <- bg[region_order]

adaptive.pca <- ggplot(
  pca.cluster.norel,
  aes(
    x = PC1,
    y = PC2,
    color = Group,
    text = paste0(
      "ID: ", ind, "<br>",
      "Location: ", Group_label, "<br>",
      "PC1: ", round(PC1, 4), "<br>",
      "PC2: ", round(PC2, 4)
    )
  )
) +
  geom_point(size = 3) +
  scale_color_manual(
    values = bg_use,
    breaks = region_order,
    labels = label_vec[region_order],
    name = "Location"
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = gray(0.8)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = gray(0.8)) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(hjust = 0.5),
    legend.position = "right"
  )+
  xlab(paste0("PC1 (", signif(pve05$pve[1], 3), "%)")) +
  ylab(paste0("PC2 (", signif(pve05$pve[2], 3), "%)"))

adaptive.pca

pdf("../figures_output/BANS_adaptive_z3_PCA.pdf", width = 8, height = 6)
print(adaptive.pca)
dev.off()
```


