---
title: "06.ESU_figures"
output: html_document
---
# LIBRARIES
```{r, message=FALSE, warning=FALSE}
library(terra)
library(raster)
library(data.table)
library(ebirdst)
library(rnaturalearth)
library(sf)
library(ggplot2)
library(ggrepel)
library(tidyterra)
library(ggnewscale)
```

# OVERALL LOCI PLOT
(Note, the following functions were generated using ChatGPT based on the above written out code that I produced, interpretation of results is my own)
## Color Pallets
The trick here is to get the color right for the spatial ESU rather than having the numbers be the same, cause the numbers aren't consistent across space with different K values.
```{r}
cols_ESU <- c(
  "ESU1" = "#187795",
  "ESU2" = "#FFBF46",
  "ESU3" = "#C1839F"
)
saveRDS(cols_ESU, "../colors/ESU_grouping_colors_K3.rds")

cols_G4 <-  c(
  "ESU1" = "#FFBF46",
  "ESU2" = "#FF5A5F",
  "ESU3" = "#8ACB88",
  "ESU4" = "#C1839F")
barplot(
  rep(1, length(cols_G4)),
  col = cols_G4,
  border = NA,
  axes = FALSE,
  space = 0,
  names.arg = names(cols_G4),
  las = 2
)
saveRDS(cols_G4, "../colors/CU_grouping_colors_K4.rds")

cols_G5 <-  c(
  "ESU1" = "#FF5A5F",
  "ESU2" = "#C1839F",
  "ESU3" = "#FFBF46",
  "ESU4" = "#8ACB88",
  "ESU5" = "#187795")
barplot(
  rep(1, length(cols_G5)),
  col = cols_G5,
  border = NA,
  axes = FALSE,
  space = 0,
  names.arg = names(cols_G5),
  las = 2
)
saveRDS(cols_G5, "../colors/CU_grouping_colors_K5.rds")
```

## Genoscapes
```{r}
source("../../03.delineate_ESUs/code/prep_genoscape_inputs.R")
source("../../03.delineate_ESUs/code/build_genoscape_surfaces.R")

############# PICK A K VALUE #############
K <- 4
Q_tibble <- read.table("../../03.delineate_ESUs/scratch/admix_output/pcadmix_output/BANS.filtered.unrel_pruned0.2.alpha0.e3.run1.admix.4.Q") %>%
setNames(paste0("ESU", 1:K))

out_plot_genoscape_png <- paste0("BANS_genoscape_overall_K4_final.png")
out_plot_genoscape_pdf <- paste0("BANS_genoscape_overall_K4_final.pdf")
genoscape_path <- "../figures_output/genoscape_outputs/"
cols <- readRDS("../colors/CU_grouping_colors_K4.rds")

K <- 3
Q_tibble <- read.table("../../03.delineate_ESUs/scratch/admix_output/pcadmix_output/BANS.filtered.unrel_pruned0.2.alpha0.e2.run1.admix.3.Q") %>%
setNames(paste0("ESU", 1:K))

out_plot_genoscape_png <- paste0("BANS_genoscape_overall_K3_final.png")
out_plot_genoscape_pdf <- paste0("BANS_genoscape_overall_K3_final.pdf")
genoscape_path <- "../figures_output/genoscape_outputs/"
cols <- readRDS("../colors/ESU_grouping_colors_K3.rds")
############# OTHER INPUT FILES #############
pca_df <- read.table("../../03.delineate_ESUs/scratch/PCA/BANS.filtered.unrel.ld25-10-0.2.eigenvec", header = FALSE)

all <- read_csv("../../../raw_data/BANS_all_sample_data_shifted.csv")
############# MAKE Q_TIBBLE #############
prep_out <- prep_genoscape_inputs(
  Q_tibble = Q_tibble,
  pca_df = pca_df,
  all = all
)

Q_tibble <- prep_out$Q_tibble
Q_matrix <- prep_out$Q_matrix
long_lat_matrix <- prep_out$long_lat_matrix
LatLong_tibble <- prep_out$LatLong_tibble
K <- ncol(Q_matrix)
q_cols <- colnames(Q_matrix)

############# GET BASE MAP AND RANGE #############
map <- ne_states(
  country = c("United States of America", "Canada"),
  returnclass = "sf"
)

na_union <- st_union(map)

breeding.sf <- st_read("../../../raw_data/spatial_files/banswa_range_2023.gpkg") %>%
  filter(season == "breeding") %>%
  st_transform(crs(map))

breeding_na <- st_intersection(breeding.sf, na_union)

############# BUILD THE GENOSCAPE #############
surf_out <- build_genoscape_surfaces(
  Q_matrix = Q_matrix,
  long_lat_matrix = long_lat_matrix,
  breeding_na = breeding_na,
  cols = cols,
  K = K,
  brick_names = q_cols,
  alpha_threshold = 0,
  krig_val=30
)

genoscape_projected <- surf_out$genoscape_projected

coords_sf <- st_as_sf(
  LatLong_tibble,
  coords = c("Long", "Lat"),
  crs = 4326   # WGS84 (lon/lat)
)

# then we want to get the sample points as coded by letter, which I established previously
cluster_geo <- readRDS("../../03.delineate_ESUs/scratch/geo_informative_labels_centroids.RDS")

## make sf points from lon/lat
cluster_geo_sf <- st_as_sf(
  cluster_geo,
  coords = c("Y", "X"),
  crs = 4326,      # lon/lat
  remove = FALSE
)

## transform to your map CRS (lamproj must be a valid CRS/proj string)
cluster_geo_sf <- st_transform(cluster_geo_sf, st_crs(lamproj))

genoscape <- ggplot() +
  ggspatial::layer_spatial(genoscape_projected) +
  geom_sf(data = breeding_na, fill = NA, linewidth = 0.1225 , alpha = 0.6) +
  geom_sf(data = countries, fill = NA, linewidth = 0.25) +
  geom_sf(data = coastlines, fill = NA, linewidth = 0.1225) +
  geom_sf(data = states, fill = NA, linewidth = 0.1225) +
  geom_sf(data = ocean, fill = "#bddbe6") +
  geom_sf(data = lakes, fill = "#bddbe6") + #bddbe6
  #geom_spatial_point(data = long_lat_tibble, mapping = aes(x = Long, y = Lat), size=1, alpha = 0.7) + 
  theme(panel.border = element_rect(color = "black", fill = NA, linewidth = 1)) +
  geom_sf(data=cluster_geo_sf,
          size = 0.5)+
  coord_sf(
  crs = st_crs(lamproj),
  xlim = c(bb["xmin"], bb["xmax"]),
  ylim = c(bb["ymin"], bb["ymax"]),
  expand = FALSE)+
  geom_label_repel(
    data = cluster_geo_sf,
    aes(label = Group_label, geometry = geometry),
    stat = "sf_coordinates",
    size = 1.5,            # increase text size (try 2.2–3.0
    fontface = "bold",
    label.size = .1,
    label.padding = unit(0.1, "lines"),
    label.r=0,
    fill = "white",
    alpha = 1,
    point.padding = 0.1,
    segment.color = "black",
    segment.size = 0.25,
    min.segment.length = 0
)

############# SAVE THE GENOSCAPE #############
ggsave(plot=genoscape, path=genoscape_path, filename=out_plot_genoscape_png, width = 7, height = 4.5)

pdf(paste0(genoscape_path,out_plot_genoscape_pdf))
print(genoscape)
dev.off()
```

## Admixture plots
```{r}
source("../../03.delineate_ESUs/code/make_admix_barplot_and_assignments.R")


admix_plot_path <- "../figures_output/admix_plots/"
admix_group_path <- "../scratch/admix_output/"
pca_df <- read.table("../../03.delineate_ESUs/scratch/PCA/BANS.filtered.unrel.ld25-10-0.2.eigenvec", header = FALSE)

pops <- readr::read_csv("../../03.delineate_ESUs/scratch/pops.csv", show_col_types = FALSE)
letter_key <- readRDS("../../03.delineate_ESUs/scratch/geo_informative_labels_centroids.RDS")

######## INPUT #########
K <- 4
Q_tibble <- read.table("../../03.delineate_ESUs/scratch/admix_output/pcadmix_output/BANS.filtered.unrel_pruned0.2.alpha0.e3.run1.admix.4.Q") %>%
setNames(paste0("ESU", 1:K))
out_plot_admix <- "BANS_admix_overall_K4.png"
out_plot_admix_pdf <- "BANS_admix_overall_K4.pdf"
out_group_admix <- "BANS_admix_grouping_K4.tsv"
cols <- readRDS("../colors/CU_grouping_colors_K4.rds")
group_label_order <- c("WA", "AK", "NT", "YT", "BC_1", "AB", "BC_2", "SK", "MT", "MB_1", "WY_1", "WY_2", "ND", "MB_2", "ON", "NS_1", "QC_2", "QC_1", "NB", "NS_2", "PE", "NL")
length(group_label_order)==22

K <- 3
Q_tibble <- read.table("../../03.delineate_ESUs/scratch/admix_output/pcadmix_output/BANS.filtered.unrel_pruned0.2.alpha0.e2.run1.admix.3.Q") %>%
setNames(paste0("ESU", 1:K))
out_plot_admix <- "BANS_admix_overall_K3.png"
out_plot_admix_pdf <- "BANS_admix_overall_K3.pdf"
out_group_admix <- "BANS_admix_grouping_K3.tsv"
cols <- readRDS("../colors/ESU_grouping_colors_K3.rds")
group_label_order <- group_label_order <- c("WA", "NT", "AK", "YT", "SK", "MB_1", "MB_2", "MT", "ND", "WY_1", "WY_2", "AB", "BC_1", "BC_2", "ON", "QC_1", "QC_2", "NS_1", "NS_2", "NB", "PE", "NL")
length(group_label_order)==22

######## RUNNING FUNCTION ########
admix_out <- make_admix_barplot_and_assignments(
  K_target = K,
  cols = cols,
  group_label_order = group_label_order,
  threshold = 0.55,
  Q_tibble = Q_tibble,
  pca_df = pca_df,
  pops = pops,
  geo_key = letter_key,
  out_plot = paste0(admix_plot_path,out_plot_admix),
  out_assignments = paste0(admix_group_path,out_group_admix)
)

K_plot <- admix_out$K_plot

K_plot
ggsave(path=admix_plot_path, filename=out_plot_admix, plot=K_plot, width = 10, height = 2.2, units = "in", dpi = 300, bg = "white")
############
pdf(paste0(admix_plot_path, out_plot_admix_pdf), width = 10, height = 2.2)
print(K_plot)
dev.off()
```

# ADAPTIVE
## Color Pallets
The trick here is to get the color right for the spatial ESU rather than having the numbers be the same, cause the numbers aren't consistent across space with different K values.
```{r}
cols_AU <- c(
  "AU1" = "#2589BD",
  "AU2" = "#FF5A5F",
  "AU3" = "#8ACB88"
)
saveRDS(cols_AU, "../colors/AU_grouping_colors_K3.rds")

cols_AU <- c(
  "AU1" = "#187795",
  "AU2" = "#FFBF46",
  "AU3" = "#C1839F"
)

cols_G4 <-  c(
  "AU1" = "#187795",
  "AU2" = "#FF5A5F",
  "AU3" = "#8ACB88",
  "AU4" = "#C1839F")
barplot(
  rep(1, length(cols_G4)),
  col = cols_G4,
  border = NA,
  axes = FALSE,
  space = 0,
  names.arg = names(cols_G4),
  las = 2
)
saveRDS(cols_G4, "../colors/AU_grouping_colors_K4.rds")

cols_G4 <-  c(
  "AU1" = "#FF5A5F",
  "AU2" = "#187795",
  "AU3" = "#FFBF46",
  "AU4" = "#C1839F")
barplot(
  rep(1, length(cols_G4)),
  col = cols_G4,
  border = NA,
  axes = FALSE,
  space = 0,
  names.arg = names(cols_G4),
  las = 2
)
saveRDS(cols_G4, "../colors/AU_grouping_colors_K4_2.rds")

cols_G5 <-  c(
  "AU1" = "#FFBF46",
  "AU2" = "#FF5A5F",
  "AU3" = "#C1839F",
  "AU4" = "#187795",
  "AU5" = "#8ACB88")
barplot(
  rep(1, length(cols_G5)),
  col = cols_G5,
  border = NA,
  axes = FALSE,
  space = 0,
  names.arg = names(cols_G5),
  las = 2
)
saveRDS(cols_G5, "../colors/AU_grouping_colors_K5.rds")
```

## Genoscapes
```{r}
source("../../03.delineate_ESUs/code/prep_genoscape_inputs.R")
source("../../03.delineate_ESUs/code/build_genoscape_surfaces.R")

############# PICK A K VALUE #############
K <- 3
Q_tibble <- read.table("../../07.adaptive_analysis/scratch/pcadmix_output/BANS.env_only.outliers_list_z3.alpha0.e2.run1.admix.3.Q") %>% setNames(paste0("AU", 1:K))

out_plot_genoscape_png <- paste0("BANS_genoscape_adaptive_K3_final.png")
out_plot_genoscape_pdf <- paste0("BANS_genoscape_adaptive_K3_final.pdf")
genoscape_path <- "../figures_output/genoscape_outputs/"
cols <- readRDS("../colors/AU_grouping_colors_K3.rds")

K <- 4
Q_tibble <- read.table("../../07.adaptive_analysis/scratch/pcadmix_output/BANS.env_only.outliers_list_z3.alpha0.e3.run1.admix.4.Q") %>% setNames(paste0("AU", 1:K))

out_plot_genoscape_png <- paste0("BANS_genoscape_adaptive_K4_final_2.png")
out_plot_genoscape_pdf <- paste0("BANS_genoscape_adaptive_K4_final_2.pdf")
genoscape_path <- "../figures_output/genoscape_outputs/"
cols <- readRDS("../colors/AU_grouping_colors_K4_2.rds")

K <- 5
Q_tibble <- read.table("../../07.adaptive_analysis/scratch/pcadmix_output/BANS.env_only.outliers_list_z3.alpha0.e4.run1.admix.5.Q") %>% setNames(paste0("AU", 1:K))

out_plot_genoscape_png <- paste0("BANS_genoscape_adaptive_K5_final.png")
out_plot_genoscape_pdf <- paste0("BANS_genoscape_adaptive_K5_final.pdf")
genoscape_path <- "../figures_output/genoscape_outputs/"
cols <- readRDS("../colors/AU_grouping_colors_K5.rds")

############# OTHER INPUT FILES #############
pca_df <- read.table("../../07.adaptive_analysis/scratch/PCA_eigen/BANS.env_only.outliers_z3_list.eigenvec", header = FALSE)

all <- read_csv("../../../raw_data/BANS_all_sample_data_shifted.csv")
############# MAKE Q_TIBBLE #############
prep_out <- prep_genoscape_inputs(
  Q_tibble = Q_tibble,
  pca_df = pca_df,
  all = all
)

Q_tibble <- prep_out$Q_tibble
Q_matrix <- prep_out$Q_matrix
long_lat_matrix <- prep_out$long_lat_matrix
LatLong_tibble <- prep_out$LatLong_tibble
K <- ncol(Q_matrix)
q_cols <- colnames(Q_matrix)

############# GET BASE MAP AND RANGE #############
map <- ne_states(
  country = c("United States of America", "Canada"),
  returnclass = "sf"
)

na_union <- st_union(map)

breeding.sf <- st_read("../../../raw_data/spatial_files/banswa_range_2023.gpkg") %>%
  filter(season == "breeding") %>%
  st_transform(crs(map))

breeding_na <- st_intersection(breeding.sf, na_union)

############# BUILD THE GENOSCAPE #############
surf_out <- build_genoscape_surfaces(
  Q_matrix = Q_matrix,
  long_lat_matrix = long_lat_matrix,
  breeding_na = breeding_na,
  cols = cols,
  K = K,
  brick_names = q_cols,
  alpha_threshold = 0,
  krig_val = 15
)

genoscape_projected <- surf_out$genoscape_projected

coords_sf <- st_as_sf(
  LatLong_tibble,
  coords = c("Long", "Lat"),
  crs = 4326   # WGS84 (lon/lat)
)

# then we want to get the sample points as coded by letter, which I established previously
cluster_geo <- readRDS("../../03.delineate_ESUs/scratch/geo_informative_labels_centroids.RDS")

## make sf points from lon/lat
cluster_geo_sf <- st_as_sf(
  cluster_geo,
  coords = c("Y", "X"),
  crs = 4326,      # lon/lat
  remove = FALSE
)

## transform to your map CRS (lamproj must be a valid CRS/proj string)
cluster_geo_sf <- st_transform(cluster_geo_sf, st_crs(lamproj))

genoscape <- ggplot() +
  ggspatial::layer_spatial(genoscape_projected) +
  geom_sf(data = breeding_na, fill = NA, linewidth = 0.1225 , alpha = 0.6) +
  geom_sf(data = countries, fill = NA, linewidth = 0.25) +
  geom_sf(data = coastlines, fill = NA, linewidth = 0.1225) +
  geom_sf(data = states, fill = NA, linewidth = 0.1225) +
  geom_sf(data = ocean, fill = "#bddbe6") +
  geom_sf(data = lakes, fill = "#bddbe6") + #bddbe6
  #geom_spatial_point(data = long_lat_tibble, mapping = aes(x = Long, y = Lat), size=1, alpha = 0.7) + 
  theme(panel.border = element_rect(color = "black", fill = NA, linewidth = 1)) +
  geom_sf(data=cluster_geo_sf,
          size = 0.5)+
  coord_sf(
  crs = st_crs(lamproj),
  xlim = c(bb["xmin"], bb["xmax"]),
  ylim = c(bb["ymin"], bb["ymax"]),
  expand = FALSE)+
  geom_label_repel(
    data = cluster_geo_sf,
    aes(label = Group_label, geometry = geometry),
    stat = "sf_coordinates",
    size = 1.5,            # increase text size (try 2.2–3.0
    fontface = "bold",
    label.size = .1,
    label.padding = unit(0.1, "lines"),
    label.r=0,
    fill = "white",
    alpha = 1,
    point.padding = 0.1,
    segment.color = "black",
    segment.size = 0.25,
    min.segment.length = 0
)

############# SAVE THE GENOSCAPE #############
ggsave(plot=genoscape, path=genoscape_path, filename=out_plot_genoscape_png, width = 7, height = 4.5)

pdf(paste0(genoscape_path,out_plot_genoscape_pdf))
print(genoscape)
dev.off()
```

## Admixture plots
```{r}
source("../../07.adaptive_analysis/code/make_admix_barplot_and_assignments_adaptive.R")

admix_plot_path <- "../figures_output/admix_plots/"
admix_group_path <- "../scratch/admix_output/"
pca_df <- read.table("../../03.delineate_ESUs/scratch/PCA/BANS.filtered.unrel.ld25-10-0.2.eigenvec", header = FALSE)

pops <- readr::read_csv("../../03.delineate_ESUs/scratch/pops.csv", show_col_types = FALSE)
letter_key <- readRDS("../../03.delineate_ESUs/scratch/geo_informative_labels_centroids.RDS")

######## INPUT #########
K <- 3
Q_tibble <- read.table("../../07.adaptive_analysis/scratch/pcadmix_output/BANS.env_only.outliers_list_z3.alpha0.e2.run1.admix.3.Q") %>% setNames(paste0("AU", 1:K))
out_plot_admix <- "BANS_admix_adaptive_K3.png"
out_plot_admix_pdf <- "BANS_admix_adaptive_K3.pdf"
out_group_admix <- "BANS_admix_adaptive_grouping_K3.tsv"
cols <- readRDS("../colors/AU_grouping_colors_K3.rds")
group_label_order <- c("AK", "YT", "NT", "BC_1", "AB", "SK", "BC_2", "MB_1", "WA", "MT", "WY_1", "WY_2", "ND", "MB_2", "ON", "QC_2", "NS_1", "NB", "NS_2", "QC_1", "PE", "NL")
length(group_label_order)==22

K <- 4
Q_tibble <- read.table("../../07.adaptive_analysis/scratch/pcadmix_output/BANS.env_only.outliers_list_z3.alpha0.e3.run1.admix.4.Q") %>% setNames(paste0("AU", 1:K))
out_plot_admix <- "BANS_admix_adaptive_K4.png"
out_plot_admix_pdf <- "BANS_admix_adaptive_K4.pdf"
out_group_admix <- "BANS_admix_adaptive_grouping_K4.tsv"
cols <- readRDS("../colors/AU_grouping_colors_K4.rds")
group_label_order <- c("WA", "AK", "NT", "YT", "BC_1", "AB", "BC_2", "SK", "MT", "MB_1", "WY_1", "WY_2", "ND", "MB_2", "ON", "NS_1", "QC_2", "QC_1", "NB", "NS_2", "PE", "NL")
length(group_label_order)==22

K <- 5
Q_tibble <- read.table("../../07.adaptive_analysis/scratch/pcadmix_output/BANS.env_only.outliers_list_z3.alpha0.e4.run1.admix.5.Q") %>% setNames(paste0("AU", 1:K))
out_plot_admix <- "BANS_admix_adaptive_K5.png"
out_plot_admix_pdf <- "BANS_admix_adaptive_K5.pdf"
out_group_admix <- "BANS_admix_adaptive_grouping_K5.tsv"
cols <- readRDS("../colors/AU_grouping_colors_K5.rds")
group_label_order <- c("WA", "NT", "YT", "BC_1", "AB", "BC_2", "AK", "SK", "MT", "MB_1", "WY_1", "WY_2", "ND", "MB_2", "ON", "NS_1", "QC_2", "QC_1", "NB", "NS_2", "PE", "NL")
length(group_label_order)==22

######## RUNNING FUNCTION ########
admix_out <- make_admix_barplot_and_assignments(
  K_target = K,
  cols = cols,
  group_label_order = group_label_order,
  threshold = 0.4,
  Q_tibble = Q_tibble,
  pca_df = pca_df,
  pops = pops,
  geo_key = letter_key,
  out_plot = paste0(admix_plot_path,out_plot_admix),
  out_assignments = paste0(admix_group_path,out_group_admix)
)

K_plot <- admix_out$K_plot

K_plot
ggsave(path=admix_plot_path, filename=out_plot_admix, plot=K_plot, width = 10, height = 2.2, units = "in", dpi = 300, bg = "white")
############
pdf(paste0(admix_plot_path, out_plot_admix_pdf), width = 10, height = 2.2)
print(K_plot)
dev.off()
```
## PCA plots
```{r}
source("../../03.delineate_ESUs/code/plot_pca_admix_solid.R")

out_plot_pca1 <- "../figures_output/pca_plots/BANS_K5_PCA_ESUcoloring.png"
base_cols_rds <- "../colors/CU_grouping_colors_K5.rds"

cols <- readRDS(base_cols_rds)
cols["No Assignment"] <- "#BDBDBD"

admix_groups <- data.table::fread("../scratch/admix_output/BANS_admix_grouping_K5.tsv") %>% 
  mutate(
    admix_group = tidyr::replace_na(admix_group, "No Assignment")
  )
cols <- readRDS(base_cols_rds)

cols["No Assignment"] <- "#BDBDBD"

pca_df <- read.table("../../03.delineate_ESUs/scratch/PCA/BANS.filtered.unrel.ld25-10-0.2.eigenvec", header = FALSE)

pops <- readr::read_csv("../../03.delineate_ESUs/scratch/pops.csv", show_col_types = FALSE)
letter_key <- readRDS("../../03.delineate_ESUs/scratch/geo_informative_labels_centroids.RDS")

eigenval <- scan("../../03.delineate_ESUs/scratch/PCA/BANS.filtered.unrel.ld25-10-0.2.eigenval")

solid_out <- plot_pca_admix_solid(
  plink_pca = pca_df,
  eigenval = eigenval,
  cols = cols,
  pops = pops,
  admix_groups = admix_groups,
  out_plot = out_plot_pca1
)

library(plotly)
ggplotly(solid_out$pca_plot)
```

```{r}
########## Colors ##########
geo_key <- readRDS("../../03.delineate_ESUs/scratch/geo_informative_labels.RDS")
base_cols <- readRDS("../../06.final_figures/colors/AU_grouping_colors_K3.rds")

out_group_label_cols_rds <- "../figures/pcadmix_AU_3_group_label_colors.rds"
out_plot <- "../figures/pca_plots/BANS_K3_PCA_AU_coloring.pdf"

admix_groups <- data.table::fread("../../07.adaptive_analysis/scratch/admix_output/BANS_admix_grouping_adaptive_G3_2.tsv")

pca_df <- pca_df <- read.table("../../07.adaptive_analysis/scratch/PCA_eigen/BANS.env_only.outliers_z3_list.eigenvec", header = FALSE) %>% select(-V1)

colnames(pca_df)[-1] <- paste0("PC", seq_len(ncol(pca_df) - 1))
names(pca_df)[1] <- "BGP_ID"

eigenval <- scan("../../07.adaptive_analysis/scratch/PCA_eigen/BANS.env_only.outliers_z3_list.eigenval")

pve <- tibble(
  PC = seq_along(eigenval),
  pve = eigenval / sum(eigenval) * 100
)

pops.groups.labels <- pops %>%
  transmute(BGP_ID, Group) %>%
  left_join(geo_key, by = "Group") %>%
  left_join(admix_groups, by = "BGP_ID") %>%
  mutate(
    Group_label = trimws(as.character(Group_label)),
    admix_group = replace_na(admix_group, "No Assignment")
  )


base_cols["No Assignment"] <- "#BDBDBD"

label_owner <- pops.groups.labels %>%
  filter(!is.na(Group_label)) %>%
  count(Group_label, admix_group, name = "n") %>%
  group_by(Group_label) %>%
  slice_max(n, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  rename(owner_au = admix_group) %>%
  mutate(owner_au = replace_na(owner_au, "No Assignment"))

missing_cols <- setdiff(unique(label_owner$owner_au), names(base_cols))

group_label_cols_df <- label_owner %>%
  arrange(owner_au, Group_label) %>%
  group_by(owner_au) %>%
  group_modify(~{
    base <- base_cols[[as.character(.y$owner_au)]]
    labels <- .x$Group_label
    
    tibble(
      Group_label = labels,
      col = colorRampPalette(c("white", base))(length(labels) + 1)[-1]
    )
  }) %>%
  ungroup()

group_label_cols <- setNames(
  group_label_cols_df$col,
  group_label_cols_df$Group_label
)

saveRDS(group_label_cols, out_group_label_cols_rds)

# Optional color check
barplot(
  rep(1, length(group_label_cols)),
  col = group_label_cols,
  border = NA,
  names.arg = names(group_label_cols),
  las = 2,
  cex.names = 0.3
)

pca.cluster <- pca_df %>%
  left_join(
    pops.groups.labels %>%
      left_join(
        label_owner %>% select(Group_label, owner_au),
        by = "Group_label"
      ),
    by = "BGP_ID"
  ) %>%
  mutate(
    Group_label = trimws(as.character(Group_label)),
    admix_group = replace_na(admix_group, "No Assignment"),
    owner_au = replace_na(owner_au, "No Assignment")
  )

if (!is.null(desired_order)) {
  desired_order <- as.character(desired_order)
  
  missing_from_palette <- setdiff(desired_order, names(group_label_cols))
  
  if (length(missing_from_palette) > 0) {
    warning(
      "These labels are in desired_order but not in group_label_cols: ",
      paste(missing_from_palette, collapse = ", ")
    )
  }
  
  legend_order <- c(
    desired_order[desired_order %in% names(group_label_cols)],
    sort(setdiff(names(group_label_cols), desired_order))
  )
} else {
  legend_order <- sort(names(group_label_cols))
}

group_label_cols <- group_label_cols[legend_order]

pca.cluster <- pca.cluster %>%
  mutate(Group_label = factor(Group_label, levels = legend_order))

p <- ggplot(
  pca.cluster,
  aes(
    x = PC1,
    y = PC2,
    color = Group_label,
    shape=owner_au
  )
) +
  scale_shape_manual(
    values = c(
      "AU1" = 16,   # filled circle
      "AU2" = 17,   # filled triangle
      "AU3" = 15,    # filled square
      "No Assignment" = 18
    ),
    name = "Assigned AU"
  ) +
  geom_point(size = 3) +
  scale_color_manual(
    values = group_label_cols,
    limits = legend_order,
    drop = FALSE,
    name = "Sampling location\nColored by AU"
  ) +
  theme_classic() +
  xlab(paste0("PC1 (", signif(pve$pve[1], 3), "%)")) +
  ylab(paste0("PC2 (", signif(pve$pve[2], 3), "%)"))

p

pdf("../figures_output/pca_plots/BANS_K3_PCA_AU_coloring.pdf", width=7, height=5)
print(p)
dev.off()
```

# TREND PLOTS
```{r}
trends_raster <- rast("../../04.predictors_of_decline/scratch/spatial_predictors/eBird_trends_BANS_proj4326.tif")
```

```{r, message=FALSE, warning=FALSE, results='hide'}
# get shapefiles from natural earth
coastlines <- ne_download(scale = 10, type = "coastline", category = "physical", returnclass = "sf")
countries <- ne_download(scale = 10, type = "admin_0_boundary_lines_land", category = "cultural", returnclass = "sf")
states <- ne_download(scale = 10, type = "admin_1_states_provinces_lines", category = "cultural", returnclass = "sf")
ocean <- ne_download(scale = 10, type = "ocean", category = "physical", returnclass = "sf")
lakes <- ne_download(scale = 10, type = "lakes", category = "physical", returnclass = "sf")
```

## nicer base trend map
```{r}
map <- ne_states(country = c("United States of America", "Canada"), 
                 returnclass = 'sf')
na_union <- st_union(map)

#BANS breeding range
breeding.sf <- st_read("../../../raw_data/spatial_files/banswa_range_2023.gpkg") %>% filter(season=="breeding")  %>% st_transform(crs(map))

#clip BANS range to just NA and make a single, clean polygon
breeding_na <- st_intersection(breeding.sf, na_union)
lamproj <- "+proj=lcc +lat_1=20 +lat_2=60 +lat_0=40 +lon_0=-100 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"

bb <- st_bbox(breeding_na) %>% st_transform(lamproj)

trends_projected <- terra::project(trends_raster, lamproj, method = "near")

colors <- colorRampPalette(brewer.pal(11, "RdBu"))(7)
barplot(
  rep(1, length(colors)),
  col = colors,
  border = NA,
  axes = FALSE,
  space = 0,
  names.arg = names(colors),
  las = 2
)

 # low = "#67001F",
 #  mid = "white",
 #  high = "#053061",

trends_plot <- ggplot() +
  tidyterra::geom_spatraster(data = trends_projected, na.rm = TRUE) +
  scale_fill_gradient2(
  low = "#3F3F3F",
  mid = "#F5F3ED",
  high = "#E0AA28",
  midpoint = 0,
  oob = scales::squish,
  na.value=NA,
  name = "Trend"
)+
  geom_sf(data = breeding_na, fill = "NA", linewidth = 0.1225 , alpha = 0.6) +
  geom_sf(data = countries, fill = NA, linewidth = 0.25) +
  geom_sf(data = coastlines, fill = NA, linewidth = 0.1225) +
  geom_sf(data = states, fill = NA, linewidth = 0.1225) +
  geom_sf(data = ocean, fill = "#bddbe6") +
  geom_sf(data = lakes, fill = "#bddbe6") +
  theme(panel.border = element_rect(color = "black", fill = NA, linewidth = 1))+
  coord_sf(
  crs = st_crs(lamproj),
  xlim = c(bb["xmin"], bb["xmax"]),
  ylim = c(bb["ymin"], bb["ymax"]),
  expand = FALSE)
trends_plot
ggsave(plot=trends_plot,"../figures_output/BANS_trends_plot_ggplot_2.png", height=4, width=8)
```

## genoscape polygons for overlap
```{r}
genoscape_polygons <- st_read("../../07.adaptive_analysis/outputs/BANS_AU_polygons_thresh.55_K3.shp")

CU_names_ref <- data.frame(
  ESU = c("AU1", "AU2", "AU3"),
  Name = c(
    "Western",
    "Eastern",
    "Northern"
  )
)

genoscape_polygons$Cluster <- factor(
  genoscape_polygons$Cluster,
  levels = CU_names_ref$ESU,
  labels = CU_names_ref$Name
)

cols <- readRDS("../colors/AU_grouping_colors_K3.rds")
names(cols) <- CU_names_ref$Name

out_overlap_plot <- paste0("../figures_output/trends/BANS_trends_plot_AU_K3_overlay_ggplot_colored_3.png")
```

```{r}
trends_plot_overlaid <- ggplot() +
  tidyterra::geom_spatraster(data = trends_projected, na.rm = TRUE) +
  scale_fill_gradient2(
  low = "#3F3F3F",
  mid = "#F5F3ED",
  high = "#E0AA28",
  midpoint = 0,
  oob = scales::squish,
  na.value=NA,
  name = "Trend"
)+
  geom_sf(data = countries, fill = NA, linewidth = 0.25)+
  geom_sf(data = coastlines, fill = NA, linewidth = 0.1225)+
  geom_sf(data = states, fill = NA, linewidth = 0.1225, alpha=.5)+
  geom_sf(data = ocean, fill = "#bddbe6")+
  geom_sf(data = lakes, fill = "#bddbe6")+
  ggnewscale::new_scale_fill() +
  ggspatial::layer_spatial(
        genoscape_polygons,
        aes(color = factor(Cluster)),
        fill = NA,
        linewidth=.5
      )+
  scale_color_manual(
    values = cols,
    breaks = names(cols),
    name = "AU"
  ) +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
  ) +
  coord_sf(
    crs = st_crs(lamproj),
    xlim = c(bb["xmin"], bb["xmax"]),
    ylim = c(bb["ymin"], bb["ymax"]),
    expand = FALSE
  )
trends_plot_overlaid
```

```{r}
ggsave(plot=trends_plot_overlaid, "../figures_output/trends/BANS_trends_plot_AU_K3_overlay_ggplot_colored_3.png", height=4, width=8)

pdf("../figures_output/trends/BANS_trends_plot_AU_K3_overlay_ggplot_colored_2.pdf", height=4, width=8)
print(trends_plot_overlaid)
dev.off()
```

```{r}
map_only <- ggplot() +
  geom_sf(data = ocean, fill = "#bddbe6", color = NA) +
  geom_sf(data = countries, fill = "lightgray", color = NA) +
  geom_sf(data = states, fill = "lightgray", color = NA) +
  geom_sf(data = lakes, fill = "#bddbe6", color = NA) +

  tidyterra::geom_spatraster(data = trends_projected, na.rm = TRUE) +
  scale_fill_gradient2(
    low = "#CA4741",
    mid = "white",
    high = "#3784BB",
    midpoint = 0,
    na.value = NA,
    name = "Trend"
  ) +

  geom_sf(data = breeding_na, fill = NA, linewidth = 0.1225, alpha = 0.6) +
  geom_sf(data = countries, fill = NA, linewidth = 0.25) +
  geom_sf(data = coastlines, fill = NA, linewidth = 0.1225) +
  geom_sf(data = states, fill = NA, linewidth = 0.1225) +

  ggnewscale::new_scale_fill() +
  ggspatial::layer_spatial(
    genoscape_polygons,
    aes(fill = factor(Cluster)),
    alpha = 0.25
  ) +
  scale_fill_manual(
    values = cols,
    breaks = names(cols),
    name = "ESU",
    na.value = NA
  ) +
  coord_sf(
    crs = st_crs(lamproj),
    xlim = c(bb["xmin"], bb["xmax"]),
    ylim = c(bb["ymin"], bb["ymax"]),
    expand = FALSE
  ) +
  theme_void() +
  theme(
    legend.position = "none",
    plot.margin = margin(0, 0, 0, 0),
    panel.background = element_rect(fill = "gray90", color = NA),
    plot.background = element_rect(fill = "gray90", color = NA)
  )

ggsave(
  "../figures_output/trends_map_only.png",
  map_only,
  width = 10,
  height = 6,
  dpi = 600,
  bg = NA
)
```

## average trend, violin plot
```{r}
data <- read.csv("../../04.predictors_of_decline/scratch/spatial_predictors/spatial_predictor_data_xy_from_AU_K3_threshold55.csv")

clean.data <- data %>% filter(!is.na(ESU))
nrow(clean.data)

clean.data$ESU <- factor(clean.data$ESU)

table(clean.data$ESU)

cols <- readRDS("../colors/AU_grouping_colors_K3.rds")

CU_names_ref <- data.frame(
  ESU = c("AU1", "AU2", "AU3"),
  Name = c(
    "Western",
    "Eastern",
    "Northern"
  )
)

K <- length(cols)

clean.data <- data %>% filter(!is.na(ESU))

clean.data.names <- left_join(clean.data, CU_names_ref, by="ESU")

data.summary <- clean.data.names %>%
  group_by(Name) %>%
  summarise(
    across(
      where(is.numeric),
      list(
        n = ~sum(!is.na(.)),
        mean = ~mean(., na.rm = TRUE),
        sd = ~sd(., na.rm = TRUE),
        min = ~min(., na.rm = TRUE),
        median = ~median(., na.rm = TRUE),
        max = ~max(., na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    )
  )

abd_trend <- ggplot(clean.data.names, aes(x = Name, y = abd_trend, fill = ESU, color = ESU)) +
  geom_violin(trim = FALSE, alpha = 0.5) +
  geom_boxplot(width = 0.1, outlier.shape = NA, alpha = 0.5) +
  scale_fill_manual(values = cols) +
  scale_color_manual(values = cols) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_classic() +
  theme(legend.position = "none")

abd_trend
ggsave(plot=abd_trend, "../figures_output/trends/updated_abd_trend_AU_K3_violin.png", width = 6, height=3)

pdf("../figures_output/trends/updated_abd_trend_G3_violin_wide.pdf", width = 8, height = 3)
print(abd_trend)
dev.off()
```
```{r}
Eastern <- clean.data %>% filter(ESU=="AU2") %>% select(abd_trend)
```