---
title: "01b.vcf_eval"
author: "Erica Robertson"
date: "2025-11-14"
output: html_document
---
```{r}
knitr::opts_knit$set(root.dir="~/Desktop/BANS_Pub/")
```
#libraries
```{r}
library(tidyverse)
library(data.table)
```
# stats
## depth
```{bash}
rsync -avzP ericacnr@colostate.edu@login.rc.colorado.edu:/scratch/alpine/ericacnr@colostate.edu/BANS/01.fastq_processing/mega-non-model-wgs-snakeflow/results/bqsr-round-0/DS_control/sample_info.tsv /Users/ericarobertson/Desktop/BANS_Pub/analysis/01.fastq_processing/data_qc/
```

```{r}
depth <- read_table("../data_qc/sample_info.tsv")
plates <- read_table("../snakemake/BANS.units.tsv")
meta.all.depth <- left_join(plates, depth, by="sample")
ggplot(meta.all, aes(x = sample, y = ave_depth, color = library)) +
  geom_point(size = 3) +
  xlab("Individual") +
  ylab("Average Depth") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
ggsave("../figures/qc/BANS_all_ind_depth.png")
```

## missingness
Let's just get some basic stats...
```{bash}
vcf="/scratch/alpine/ericacnr@colostate.edu/BANS/01.fastq_processing/mega-non-model-wgs-snakeflow/results/bqsr-round-0/vcf/filtered_HC/BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.recode.vcf"
vcftools --gzvcf $vcf --missing-indv
```

```{bash}
bcftools stats $vcf > BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.0.8miss.recode.stats
```
SN      0       number of samples:      206
SN      0       number of records:      5219558
SN      0       number of no-ALTs:      0
SN      0       number of SNPs: 5219558
SN      0       number of MNPs: 0
SN      0       number of indels:       0
SN      0       number of others:       0
SN      0       number of multiallelic sites:   0
SN      0       number of multiallelic SNP sites:       0

So, this all looks okay.

```{bash}
rsync -avzP ericacnr@colostate.edu@login.rc.colorado.edu://scratch/alpine/ericacnr@colostate.edu/BANS/01.fastq_processing/mega-non-model-wgs-snakeflow/vcf_eval/out.imiss_filtered_ind /Users/ericarobertson/Desktop/BANS_Pub/analysis/01.fastq_processing/data_qc
```

```{r}
miss <- read_table("../data_qc/out.imiss_filtered_ind") %>% rename(sample=INDV)
plates <- read_table("../snakemake/BANS.units.tsv")
meta.all <- left_join(plates, miss, by="sample")
ggplot(meta.all, aes(x = sample, y = F_MISS, color = library)) +
  geom_point(size = 3) +
  xlab("Individual") +
  ylab("Proportion Missing (F_MISS)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
ggsave("../figures/qc/BANS_all_ind_missingness_2.png")

ggplot(meta.all, aes(x = library, y = F_MISS, color = library)) +
  geom_boxplot(size = 3) +
  xlab("Individual") +
  ylab("Proportion Missing (F_MISS)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

Well, the plate that I extracted has much lower missingness, and much less variable missingess. But I did the individual level missingness filtering right and we're not seeing a huge spread or any outliers.

# Sample Type Check
```{r}
tissue_samples <- read.csv("../../../raw_data/BANS_Master_sheet1.csv") %>% filter(SampleType=="Tissue") %>% dplyr::select(BGP_ID)
plates <- read_table("../snakemake/BANS.units.tsv")
meta.all <- left_join(plates, miss, by="sample")

meta.all <- meta.all %>%
  mutate(
    tissue_type = if_else(sample %in% tissue_samples$BGP_ID, "Tissue", "Blood")
  )

ggplot(meta.all, aes(x = sample, y = F_MISS, color = tissue_type)) +
  geom_point(size = 3) +
  xlab("Individual") +
  ylab("Proportion Missing (F_MISS)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
ggsave("../figures/qc/BANS_all_ind_missingness_tissue_2.png")
```

```{r}
tissue_samples <- read.csv("../../../raw_data/BANS_Master_sheet1.csv") %>% filter(SampleType=="Tissue") %>% dplyr::select(BGP_ID)
plates <- read_table("../snakemake/BANS.units.tsv")
meta.all <- left_join(plates, depth, by="sample")

meta.all <- meta.all %>%
  mutate(
    tissue_type = if_else(sample %in% tissue_samples$BGP_ID, "Tissue", "Blood")
  )

ggplot(meta.all, aes(x = sample, y = ave_depth, color = tissue_type)) +
  geom_point(size = 3) +
  xlab("Individual") +
  ylab("Average Depth") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
ggsave("../figures/qc/BANS_all_ind_depth_tissue_2.png")
```
# PCA for plate effects
## All Individuals (after missingness filtering)
#### pruning
```{bash}
#GIVING SNP IDS
vcf="/scratch/alpine/ericacnr@colostate.edu/BANS/01.fastq_processing/mega-non-model-wgs-snakeflow/results/bqsr-round-0/vcf/filtered_HC/BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.ind03miss.0.8miss.recode.vcf"

conda activate GWAS2
plink --vcf $vcf --aec --recode --out BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.0.8miss.recode

cat BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.0.8miss.recode.map | awk -F"\t" '{split($1, a, "__"); print $1"\t"a[1]"."$4"\t"$3"\t"$4}' > BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.0.8miss.recode_loc.map

plink --ped BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.0.8miss.recode.ped \
--map BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.0.8miss.recode_loc.map \
--make-bed --aec \
--out BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.0.8miss.recode

bfile="BANS.all.ds6x.pass-maf-0.05.bialSNP.filtered.0.8miss.recode"

plink --bfile $bfile --aec \
--indep-pairwise 25 10 0.5 --out PC1_allindv_noimpute
#Pruning complete.  555915 of 5219558 variants removed.

plink --bfile $bfile --aec --extract PC1_allindv_noimpute.prune.in \
  --make-bed --out pruned_allindv
#4663643 variants and 206 people pass filters and QC.

plink --bfile pruned_allindv --aec --allow-no-sex --pca 205 \
  --out PC1_allindv_noimpute.ld25-10-0.5
#Total genotyping rate is 0.887949.
#4663643 variants and 206 people pass filters and QC.
```

```{bash}
rsync -avzP ericacnr@colostate.edu@login.rc.colorado.edu:/scratch/alpine/ericacnr@colostate.edu/BANS/01.fastq_processing/mega-non-model-wgs-snakeflow/vcf_eval/PC1_allindv_noimpute.ld25-10-0.5*eigen* /Users/ericarobertson/Desktop/BANS_Pub/analysis/01.fastq_processing/data_qc
```
#### Time to plot the PCA.
####grouping pops
The actual group pops is missing indv that I filtered for missingness, so this is a quick temporary one...

```{r}
library(rnaturalearth)
library(geosphere)

meta <- read.csv("../../../raw_data/BANS_all_sample_data_shifted.csv")
coords <- meta %>% dplyr::select(Long, Lat)
dist_matrix <- distm(as.matrix(coords), fun = distHaversine)

libraries <- read.delim("../snakemake/BANS.units.tsv") %>% rename(BGP_ID=sample)

lib.meta <- left_join(meta, libraries %>% dplyr::select(library, BGP_ID), by="BGP_ID")

# clustering
hclust_groups <- hclust(as.dist(dist_matrix), method = "complete") # complete method here avoids chaining, ensuring all samples in a cluster are within the specified distance of each other. If you want some flexibility you could try "average"

# cut tree at 111.111 km to assign spatial clusters
meta$GroupID <- cutree(hclust_groups, h = 111111)

# create population labels and filter for minimum cluster size
meta.sub <- meta %>%
  mutate(Pop = paste0("Cluster_", GroupID)) %>%
  group_by(Pop) %>%
  filter(n() > 3) %>%  # keep clusters with ≥4 individuals
  ungroup()

pops <- data.frame(BGP_ID = meta.sub$BGP_ID, Group = meta.sub$Pop)

library(rnaturalearth)
library(raster)
library(terra)
map <- ne_states(country = c("United States of America", "Canada", "Mexico"), 
                 returnclass = 'sf')

# create colors for clusters
n_clusters <- length(unique(meta.sub$Pop))
cluster_colors <- rainbow(n_clusters)
names(cluster_colors) <- unique(meta.sub$Pop)

meta.b <- meta.sub %>%
  group_by(Pop) %>%
  mutate(cluster_size = n()) %>%
  ungroup()

# plot it
p <- ggplot() +
  geom_sf(data = map, fill = "white", color = "black", size = 0.1) +
  geom_point(data = meta.b, aes(x = Long, y = Lat, color = Pop),
             size = 1.5, position = position_jitter(width = 0.1, height = 0.1)) +
  geom_text(data = meta.b,
            aes(x = Long, y = Lat, label = Pop),
            size = 2, vjust = -0.6) +
  scale_colour_manual(values = cluster_colors, name = "Spatial Clusters", guide = "none") +
  coord_sf(xlim = c(-145, -55), ylim = c(15, 80)) + 
  theme_minimal() +
  labs(title = "All clusters with ≥4 individuals",
       x = "Longitude",
       y = "Latitude")
p
ggsave("../figures/maps/cluster_plot_randomcolors.png")
```

```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
plink_pca <- read.table("../data_qc/PC1_allindv_noimpute.ld25-10-0.5.eigenvec", header = FALSE) %>% dplyr::select(-V2)
eigenval <- scan("../data_qc/PC1_allindv_noimpute.ld25-10-0.5.eigenval")

pve05 <- data.frame(PC = 1:205, pve = eigenval/sum(eigenval)*100)

# set names
pca05 <- plink_pca
colnames(pca05)[-1] <- paste0("PC", 1:(ncol(pca05) - 1))
names(pca05)[1] <- "ind"
nrow(pca05)
#correct number of samples

# Join PCA data
pca.cluster <- left_join(pca05, pops %>% rename(ind=BGP_ID), by="ind")  # left_join PCA first to preserve PC columns

# Factor for cluster order
region_order <- c(
  "Cluster_1", "Cluster_5", "Cluster_16", "Cluster_4", "Cluster_9", "Cluster_12", "Cluster_2", "Cluster_20",
  "Cluster_3", "Cluster_8", "Cluster_17", "Cluster_21", "Cluster_18", "Cluster_19",
  "Cluster_6", "Cluster_14", "Cluster_11", "Cluster_10", "Cluster_13", "Cluster_15", "Cluster_7"
)
pca.cluster$Cluster <- factor(pca.cluster$Group, levels = region_order)

# Define clusters by region
West_clusters    <- c("Cluster_1", "Cluster_5", "Cluster_16", "Cluster_4", "Cluster_8", "Cluster_12", "Cluster_3")
Central_clusters <- c("Cluster_2", "Cluster_20", "Cluster_17", "Cluster_7", "Cluster_18", "Cluster_19", "Cluster_21")
East_clusters    <- c("Cluster_11", "Cluster_14", "Cluster_13", "Cluster_15", "Cluster_6", "Cluster_9", "Cluster_10")

# Assign colors by cluster
library(RColorBrewer)
W.cols <- setNames(brewer.pal(n = length(West_clusters), "Reds"), West_clusters)
C.cols <- setNames(brewer.pal(n = length(Central_clusters), "Blues"), Central_clusters)
E.cols <- setNames(brewer.pal(n = length(East_clusters), "Greens"), East_clusters)
bg <- c(W.cols, C.cols, E.cols)

# Plot
b05 <- ggplot(pca.cluster, aes(PC1, PC2, col = Cluster)) +
  geom_point(size = 3) +
  scale_color_manual(values = bg) +
  theme_light() +
  xlab(paste0("PC1 (", signif(pve05$pve[1], 3), "%)")) +
  ylab(paste0("PC2 (", signif(pve05$pve[2], 3), "%)"))

b05
ggsave("../figures/pca/BANS_initialeval_prunedloci_noimputation_bestindv_PCA_2.png", width=8)
```

```{r}
p <- ggplot() +
  geom_sf(data = map, fill = "white", color = "black", size = 0.1) +
  geom_point(data = meta.b, aes(x = Long, y = Lat, color = Pop),
             size = 1.5, position = position_jitter(width = 0.1, height = 0.1)) +
  geom_text(data = meta.b,
            aes(x = Long, y = Lat, label = Pop),
            size = 2, vjust = -0.6) +
  scale_colour_manual(values = bg, name = "Spatial Clusters", guide = "none") +
  coord_sf(xlim = c(-145, -55), ylim = c(15, 80)) + 
  theme_minimal() +
  labs(title = "All clusters with ≥4 individuals",
       x = "Longitude",
       y = "Latitude")
p
ggsave("../figures/maps/cluster_bestindv_plot_2.png")
```

### no related indv
```{r}
library(plotly)

rel <- ggplot(pca.cluster, aes(
    x = PC1, 
    y = PC2, 
    col = Group,
    text = paste0(
        "ID: ", ind, "<br>",
        "Group: ", Group, "<br>",
        "PC1: ", round(PC1, 4), "<br>",
        "PC2: ", round(PC2, 4)
    )
)) +
  geom_point(size = 3) +
  scale_color_manual(values = bg) +
  theme_light() +
  xlab(paste0("PC1 (", signif(pve05$pve[1], 3), "%)")) +
  ylab(paste0("PC2 (", signif(pve05$pve[2], 3), "%)"))

rel
ggplotly(rel)
```

