---
title: "05.2.RDA_env_selection"
author: "Erica Robertson"
date: "2026-03-23"
output: html_document
---

# LIBRARIES
```{r, message=F}
library(vegan)
library(adegenet)
library(tidyverse)
require(readr)
library(data.table)
# if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# BiocManager::install("LEA")
library(LEA)
library(openxlsx)
library(sf)
library(geosphere)
```

# INTRODUCTION
The follow script outlines to steps for environmental variable selection using RDA forward selection. The goal of this process is to take all the potential variables I've collected and pick those that most strongly predict allele frequencies. This is computationally intensive, so the actual running of stuff will be on the cluster. I've got that outlined in 05.1.RDA_env_vars_cluster.R. In that script I also subsample to 1mil SNPs: this is to save on computation time and effort.

The input files for the bash script are:
1. subsampled_noheader.raw
2. data/Bans_env_pop.txt

Some notes on the forward selection process: "Forward selection starts from a “null” model where the response is explained only by an intercept. Variables are then added to the model one by one to try to reach the amount of variance explained by a “full” model (i.e., model including all the explanatory variables), while limiting the amount of redundancy among included variables." (Capblanq 2018)

Some things of note about the selection parameters.
direction="forward" : starts with no predictors, add them one at a time (alternatives are backward or both). forwards is preffered to avoid over-fitting from the get go.
R2permutations=1000 : for each variable, the functions adds it to the model, runs a permutation test (here, 1000 permutations) and estimates a p-value for the increase in explained variance. more permutations = more stable p-values, 1000 is standard.
Pin=0.01 : entry threshold (p-value cutoff). a variable is only added in the p_perm < 0.01. This is a fairly strict cutoff, 0.05 is more standard but can help reduce false positives.
R2scope = T : the activates the double-stopping criterion, a variable is added only if BOTH are true, 1) is p-value is < pin, 2) the cumulative adjusted R2 of the model does not exceed the adjusted R2 of the full model. This prevents over-fitting and protects against adding weak but "significant" variables later in the process.

# ENVIRONMENTAL VARIABLES
So this is getting together the climate data for each population and filtering to pops (clusters) that have more than 4 samples.
```{r}
pops <- read.csv("../03.delineate_ESUs/scratch/pops.csv")
env <- fread("../scratch/BANS.CHELSA.txt")

#renaming the soil content columns so it's consistent and also filtering by sample size
clim <- pops %>%
  left_join(env, by = "BGP_ID") %>%
  add_count(Group, name = "n_group") %>%
  filter(n_group >= 4) %>%
  select(-n_group)
```

```{r}
## reorder so that env and allele frequencies per pop are reported in the same order
plink_pca <- read.table("../../03.delineate_ESUs/scratch/PCA/BANS.filtered.unrel.ld25-10-0.5.eigenvec", header = FALSE) %>% dplyr::select(V1) %>% rename(BGP_ID=V1)

clim_ordered <- plink_pca %>% left_join(clim)

write.table(clim_ordered, file = "../scratch/Bans_envonly_pop.txt", row.names=F, quote=F, sep="\t")
```

```{bash}
rsync -avzP /Users/ericarobertson/Desktop/BANS_Pub/analysis/05.genomic_offset/scratch/Bans_envonly_pop.txt ericacnr@colostate.edu@login.rc.colorado.edu:/scratch/alpine/ericacnr@colostate.edu/BANS/03.GEA/data
```

# GENOTYPE FILE
Bash script to make the input .raw file:
```{bash}
vcf="/scratch/alpine/ericacnr@colostate.edu/BANS/02.5.neutral_pop_struc/BANS.filtered.unrel.vcf.gz"

conda activate GWAS2
plink --vcf $vcf --aec --recode --out BANS.unrel_ind.imputed4.1

cat BANS.unrel_ind.imputed4.1.map | awk -F"\t" '{split($1, a, "__"); print $1"\t"a[1]"."$4"\t"$3"\t"$4}' > BANS.unrel_ind.imputed4.1_loc.map

plink --ped BANS.unrel_ind.imputed4.1.ped \
  --map BANS.unrel_ind.imputed4.1_loc.map \
  --make-bed --aec \
  --out BANS.unrel_ind.imputed4.1

######## SUBSAMPLED SNPS ##########
conda activate GWAS2
N=1000000
bfile="BANS.unrel_ind.imputed4.1"

cut -f2 ${bfile}.bim | shuf -n $N > snps_1mil.txt
plink --bfile $bfile --extract snps_1mil.txt -aec --make-bed --out subsampled_pruned_unrelindv.1mil

plink --bfile subsampled_pruned_unrelindv.1mil -aec --recode A --out subsampled_pruned_unrelindv.1mil

#recut header
cut -d$'\t' -f7- subsampled_pruned_unrelindv.1mil.raw > subsampled_noheader.raw
```

# ACTUAL SBATCH SCRIPT
```{bash}
#!/bin/bash
#SBATCH --job-name=RDA
#SBATCH --output=RDA.%j.out
#SBATCH --error=RDA.%j.err
#SBATCH -t 8:00:00
#SBATCH -p amilan
#SBATCH --qos=normal
#SBATCH --nodes=1
#SBATCH --ntasks-per-node 24
#SBATCH --mem=90G
#SBATCH --mail-type=ALL
#SBATCH  --mail-user=ericacnr@colostate.edu

source ~/.bashrc

############ Load and Run RDA.R script ############
conda activate R
Run your R script
Rscript RDA_env_vars_cluster.R
```

# CLUSTER RESULTS
RDA_output/BANS.varsel_results.RDS

```{bash}
rsync -avzP ericacnr@colostate.edu@login.rc.colorado.edu:/scratch/alpine/ericacnr@colostate.edu/BANS/03.GEA/results/RDA_output/BANS.varsel_results1mil_narrow.RDS /Users/ericarobertson/Desktop/BANS_adaptive_units/analysis/03.GEA/results/RDA_output/
```

```{r}
# Get names of selected predictors
mod <- readRDS("../scratch/RDA_output/BANS.varsel_envonly_results1mil.RDS")
mod$anova
vif.cca(mod)

top_vars <- attr(terms(mod), "term.labels")

top_vars
env <- fread("../scratch/Bans_envonly_pop.txt")
env %>% select(Group, BGP_ID, Long, Lat, bio14, bio09, bio06) %>% write.csv("../scratch/RDA_output/BANS_topRDA_envonly_vars1mil_narrow.csv", row.names = F)
```
Results:
                                      R2.adj Df    AIC      F Pr(>F)   
+ Temperate.or.Subpolar.Shrubland 0.0019295  1 2513.7 1.3867  0.002 **
+ bio14                           0.0033711  1 2514.4 1.2878  0.002 **
+ bio09                           0.0043522  1 2515.2 1.1951  0.002 **
+ Wetland                         0.0049450  1 2516.1 1.1173  0.002 **
+ bio06                           0.0054900  1 2516.9 1.1074  0.002 **
+ clay                            0.0059692  1 2517.8 1.0940  0.002 **
<All variables>                   0.0102967                           

```{r}
env_pop <- read.csv("../scratch/RDA_output/BANS_topRDA_vars1mil_narrow.csv")
cor_mat <- cor(env_pop %>% dplyr::select(-Group, -BGP_ID, -Long, -Lat), use = "pairwise.complete.obs", method = "pearson")

cor_df <- as.data.frame(cor_mat) %>%
  rownames_to_column("Var1") %>%
  pivot_longer(-Var1, names_to = "Var2", values_to = "cor")

ggplot(cor_df, aes(Var1, Var2, fill = cor)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B",
    midpoint = 0, limits = c(-1, 1)
  ) +
  coord_fixed() +
  theme_void() +
  theme(
    axis.text.x = element_text(angle = 90, size = 6),
    axis.text.y = element_text(size = 6)
  )
```
So we have some variables come up that are highly correlated. Let's check those out and exclude some.

bio14 Precipitation of the Driest Month
bio09 Mean Daily Near-Surface Air Temperature of the Wettest Quarter
bio06 Mean Daily Minimum Near-Surface Air Temperature of the Coldest Month

So, I'm removing all the "coldest" ones because we're interested in stuff happening in during the breeding season, which would not be the coldest part of the year. To me this means that we want the more biologically appropriate variable that it came up as correlated with. So that leaves Mean Monthly Precipitation of the Driest Quarter (bio17) and Mean Daily Near-Surface Air Temperature of the Wettest Quarter (bio08), which makes some sense biologically.

So that leaves these are the top variables:
Temperate or Subpolar Shrubland land cover
bio17 (Mean Monthly Precipitation of the Driest Quarter)
Wetland
clay
bio08 (Mean Daily Near-Surface Air Temperature of the Wettest Quarter)
Water

```{r}
top_vars_uncor <- c("Temperate.or.Subpolar.Shrubland", "bio06", "Wetland", "clay", "bio14", "bio09")

env %>% select(Pop, BGP_ID, Long, Lat, all_of(top_vars_uncor)) %>% write.csv("BANS_topRDA_vars1mil_uncor.csv", row.names = F)

env_pop <- read.csv("BANS_topRDA_vars1mil_uncor.csv")
cor_mat <- cor(env_pop %>% dplyr::select(-Pop, -BGP_ID, -Long, -Lat), use = "pairwise.complete.obs", method = "pearson")

cor_df <- as.data.frame(cor_mat) %>%
  rownames_to_column("Var1") %>%
  pivot_longer(-Var1, names_to = "Var2", values_to = "cor")

ggplot(cor_df, aes(Var1, Var2, fill = cor)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B",
    midpoint = 0, limits = c(-1, 1)
  ) +
  coord_fixed() +
  theme_void() +
  theme(
    axis.text.x = element_text(angle = 90, size = 6),
    axis.text.y = element_text(size = 6)
  )
```
That looks a lot better! Let's go ahead with these variables.