---
title: "05.2.RDA_env_selection"
author: "Erica Robertson"
date: "2026-03-23"
output: html_document
---

# LIBRARIES
```{r, message=F}
library(vegan)
library(adegenet)
library(tidyverse)
require(readr)
library(data.table)
# if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# BiocManager::install("LEA")
library(LEA)
library(openxlsx)
library(sf)
library(geosphere)
```

# INTRODUCTION
The follow script outlines to steps for environmental variable selection using RDA forward selection. The goal of this process is to take all the potential variables I've collected and pick those that most strongly predict allele frequencies. This is computationally intensive, so the actual running of stuff will be on the cluster. I've got that outlined in 05.1.RDA_env_vars_cluster.R. In that script I also subsample to 1mil SNPs: this is to save on computation time and effort.

The input files for the bash script are:
1. subsampled_noheader.raw
2. data/Bans_env_pop.txt

Some notes on the forward selection process: "Forward selection starts from a “null” model where the response is explained only by an intercept. Variables are then added to the model one by one to try to reach the amount of variance explained by a “full” model (i.e., model including all the explanatory variables), while limiting the amount of redundancy among included variables." (Capblanq 2018)

Some things of note about the selection parameters.
direction="forward" : starts with no predictors, add them one at a time (alternatives are backward or both). forwards is preffered to avoid over-fitting from the get go.
R2permutations=1000 : for each variable, the functions adds it to the model, runs a permutation test (here, 1000 permutations) and estimates a p-value for the increase in explained variance. more permutations = more stable p-values, 1000 is standard.
Pin=0.01 : entry threshold (p-value cutoff). a variable is only added in the p_perm < 0.01. This is a fairly strict cutoff, 0.05 is more standard but can help reduce false positives.
R2scope = T : the activates the double-stopping criterion, a variable is added only if BOTH are true, 1) is p-value is < pin, 2) the cumulative adjusted R2 of the model does not exceed the adjusted R2 of the full model. This prevents over-fitting and protects against adding weak but "significant" variables later in the process.

# ENVIRONMENTAL VARIABLES
So this is getting together the climate data for each population and filtering to pops (clusters) that have more than 4 samples.
```{r}
pops <- read.csv("../03.delineate_ESUs/scratch/pops.csv")
env <- fread("../scratch/BANS.CHELSA.txt")

#renaming the soil content columns so it's consistent and also filtering by sample size
clim <- pops %>%
  left_join(env, by = "BGP_ID") %>%
  add_count(Group, name = "n_group") %>%
  filter(n_group >= 4) %>%
  select(-n_group)
```

```{r}
## reorder so that env and allele frequencies per pop are reported in the same order
plink_pca <- read.table("../../03.delineate_ESUs/scratch/PCA/BANS.filtered.unrel.ld25-10-0.5.eigenvec", header = FALSE) %>% dplyr::select(V1) %>% rename(BGP_ID=V1)

clim_ordered <- plink_pca %>% left_join(clim)

write.table(clim_ordered, file = "../scratch/Bans_envonly_pop.txt", row.names=F, quote=F, sep="\t")
```

```{bash}
rsync -avzP /Users/ericarobertson/Desktop/BANS_Pub/analysis/05.genomic_offset/scratch/Bans_envonly_pop.txt ericacnr@colostate.edu@login.rc.colorado.edu:/scratch/alpine/ericacnr@colostate.edu/BANS/03.GEA/data
```

# GENOTYPE FILE
Bash script to make the input .raw file:
```{bash}
vcf="/scratch/alpine/ericacnr@colostate.edu/BANS/02.5.neutral_pop_struc/BANS.filtered.unrel.vcf.gz"

conda activate GWAS2
plink --vcf $vcf --aec --recode --out BANS.unrel_ind.imputed4.1

cat BANS.unrel_ind.imputed4.1.map | awk -F"\t" '{split($1, a, "__"); print $1"\t"a[1]"."$4"\t"$3"\t"$4}' > BANS.unrel_ind.imputed4.1_loc.map

plink --ped BANS.unrel_ind.imputed4.1.ped \
  --map BANS.unrel_ind.imputed4.1_loc.map \
  --make-bed --aec \
  --out BANS.unrel_ind.imputed4.1

######## SUBSAMPLED SNPS ##########
N=1000000
bfile="BANS.unrel_ind.imputed4.1"

cut -f2 ${bfile}.bim | shuf -n $N > snps_1mil.txt
plink --bfile $bfile --extract snps_1mil.txt -aec --make-bed --out subsampled_unrelindv.1mil

plink --bfile subsampled_unrelindv.1mil -aec --recode A --out subsampled_unrelindv.1mil

#recut header
cut -d$'\t' -f7- subsampled_unrelindv.1mil.raw > subsampled_noheader.raw
```

# PRUNING TO JUST RELEVANT VARIABLES
So, on initial runs of this workflow my VIF was very inflated suggesting high collinearity between predictors. This is to be expected as lots of environmental variables are highly correlated. So, the goal is to take the CHELSA variables and prune them a bit using what seems biologically relevant and what isn't already super correlated and then run those variables into the forward selection. The filter by VIF later. I want to try and catch at least one temp variable, one precip variable, a moisture variable and a seasonaility variable...

General Temperature:
| Variable | Name                    | Interpretation                               |
| -------- | ----------------------- | -------------------------------------------- |
| BIO1     | Annual Mean Temperature | Overall average temperature across year      |
| BIO2     | Mean Diurnal Range      | Avg (daily max − min); day-night variability |
| BIO3     | Isothermality           | Day-night variation relative to annual range |
| BIO4     | Temperature Seasonality | How variable temperature is across year      |

Temperature Extremes:
| Variable | Name                             | Interpretation                      |
| -------- | -------------------------------- | ----------------------------------- |
| BIO5     | Max Temperature of Warmest Month | Heat extremes                       |
| BIO6     | Min Temperature of Coldest Month | Cold extremes                       |
| BIO7     | Temperature Annual Range         | BIO5 − BIO6 (overall thermal range) |

Seasonal Temperature:
| Variable | Name                      | Interpretation             |
| -------- | ------------------------- | -------------------------- |
| BIO8     | Mean Temp Wettest Quarter | Temp during wet season     |
| BIO9     | Mean Temp Driest Quarter  | Temp during dry season     |
| BIO10    | Mean Temp Warmest Quarter | Growing season temperature |
| BIO11    | Mean Temp Coldest Quarter | Winter severity            |

General Precip:
| Variable | Name                        | Interpretation               |
| -------- | --------------------------- | ---------------------------- |
| BIO12    | Annual Precipitation        | Total yearly precipitation   |
| BIO13    | Precipitation Wettest Month | Extreme wet conditions       |
| BIO14    | Precipitation Driest Month  | Extreme dry conditions       |
| BIO15    | Precipitation Seasonality   | Variability in precipitation |

Seasonal Precip:
| Variable | Name                          | Interpretation               |
| -------- | ----------------------------- | ---------------------------- |
| BIO16    | Precipitation Wettest Quarter | Wet season precipitation     |
| BIO17    | Precipitation Driest Quarter  | Dry season precipitation     |
| BIO18    | Precipitation Warmest Quarter | Growing season precipitation |
| BIO19    | Precipitation Coldest Quarter | Winter precipitation         |

So let's try
Temperature

BIO1 — mean climate
BIO5 — heat extremes
BIO6 — cold extremes
BIO10 — breeding season temperature

Precipitation

BIO12 — total moisture
BIO18 — breeding season precipitation

Seasonality

BIO4 — temperature variability
BIO15 — precipitation variability

# ACTUAL SBATCH SCRIPT
```{bash, label="run_var_sel.sbatch"}
#!/bin/bash
#SBATCH --job-name=varsel1mil
#SBATCH --output=varsel1mil.%j.out
#SBATCH --error=varsel1mil.%j.err
#SBATCH -t 8:00:00
#SBATCH -p amilan
#SBATCH --qos=normal
#SBATCH --nodes=1
#SBATCH --ntasks-per-node 24
#SBATCH --mem=90G
#SBATCH --mail-type=ALL
#SBATCH  --mail-user=ericacnr@colostate.edu

source ~/.bashrc

############ Load and Run RDA.R script ############
conda activate R
Run your R script
Rscript RDA_env_vars_cluster.R
```

# CLUSTER RESULTS
## Run with All Environmental Variables (Bio01-19)
RDA_output/BANS.varsel_results.RDS

RDA Model results:
$r.squared
[1] 0.08888219

$adj.r.squared
[1] 0.00965455

Variable Selection:
                  R2.adj Df    AIC      F Pr(>F)
+ bio17         0.0020712  1 2513.7 1.4151  0.002 **
+ bio06         0.0032368  1 2514.4 1.2327  0.002 **
+ bio09         0.0038653  1 2515.3 1.1249  0.002 **
+ bio19         0.0043596  1 2516.2 1.0978  0.002 **
+ bio08         0.0051870  1 2517.0 1.1630  0.002 **
+ bio14         0.0062037  1 2517.7 1.1995  0.002 **
+ bio01         0.0065352  1 2518.6 1.0647  0.002 **
+ bio05         0.0070751  1 2519.5 1.1050  0.002 **
+ bio04         0.0074375  1 2520.3 1.0701  0.002 **
+ bio02         0.0080437  1 2521.2 1.1167  0.002 **
+ bio18         0.0085708  1 2522.0 1.1010  0.002 **
+ bio03         0.0088972  1 2522.9 1.0622  0.006 **
<All variables> 0.0096545
---
[1] "VIF"
     bio17      bio06      bio09      bio19      bio08      bio14      bio01
1664.02485  198.96653   49.73612  433.99826   12.64477  981.48488  140.74337
     bio05      bio04      bio02      bio18      bio03
 129.44836  192.83612  230.74923   22.36195  264.74469
 
## Run with the Pruned Variable Set

Notes on references for using the VIF:
1. VIF > 10 indicates a serious collinearity problem https://quantifyinghealth.com/vif-threshold/ 
2. To address multicollinearity, we retained layers with a Spearman correlation coefficient (r) <0.7 and Variance Inflation Factor (VIF) <10. https://www.sciencedirect.com/science/article/pii/S2351989425000186?utm_source=chatgpt.com
3. By applying a Variance Inflation Factor (VIF) criterion of <10... https://advanced.onlinelibrary.wiley.com/doi/10.1002/advs.202506492
4. Variance inflation factor (VIF) analysis was then usedto exclude highly correlated variables using a VIF threshold of 10(Dyer et al., 2010). https://doi.org/10.1111/mec.15637

```{bash}
rsync -avzP ericacnr@colostate.edu@login.rc.colorado.edu:/scratch/alpine/ericacnr@colostate.edu/BANS/03.GEA/results/RDA_output/BANS.varsel_envonly_results1mil.RDS /Users/ericarobertson/Desktop/BANS_Pub/analysis/05.genomic_offset/scratch/RDA_output/
```

```{r}
# Get names of selected predictors
mod <- readRDS("../scratch/RDA_output/BANS.varsel_envonly_results1mil.RDS")

top_vars <- attr(terms(mod), "term.labels")

top_vars

env <- fread("../scratch/Bans_envonly_pop.txt")
env %>% select(Group, BGP_ID, Long, Lat, top_vars) %>% write.csv("../scratch/RDA_output/BANS_topRDA_envonly_vars1mil_narrow.csv", row.names = F)
```

```{r}
env_pop <- read.csv("../scratch/RDA_output/BANS_topRDA_envonly_vars1mil_narrow.csv")
cor_mat <- cor(env_pop %>% dplyr::select(-Group, -BGP_ID, -Long, -Lat), use = "pairwise.complete.obs", method = "pearson")

cor_df <- as.data.frame(cor_mat) %>%
  rownames_to_column("Var1") %>%
  pivot_longer(-Var1, names_to = "Var2", values_to = "cor")

ggplot(cor_df, aes(Var1, Var2, fill = cor)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B",
    midpoint = 0, limits = c(-1, 1)
  ) +
  coord_fixed() +
  theme_void() +
  theme(
    axis.text.x = element_text(angle = 90, size = 6),
    axis.text.y = element_text(size = 6)
  )
```

```{r}
top_vars_uncor <- c()

env %>% select(Pop, BGP_ID, Long, Lat, all_of(top_vars_uncor)) %>% write.csv("BANS_envonly_vars1mil_uncorvars.csv", row.names = F)

env_pop <- read.csv("BANS_envonly_vars1mil_uncorvars.csv")
cor_mat <- cor(env_pop %>% dplyr::select(-Pop, -BGP_ID, -Long, -Lat), use = "pairwise.complete.obs", method = "pearson")

cor_df <- as.data.frame(cor_mat) %>%
  rownames_to_column("Var1") %>%
  pivot_longer(-Var1, names_to = "Var2", values_to = "cor")

ggplot(cor_df, aes(Var1, Var2, fill = cor)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B",
    midpoint = 0, limits = c(-1, 1)
  ) +
  coord_fixed() +
  theme_void() +
  theme(
    axis.text.x = element_text(angle = 90, size = 6),
    axis.text.y = element_text(size = 6)
  )
```