---
title: "05.3.RDA_offset"
author: "Erica Robertson"
date: "2026-03-23"
output: html_document
---

So, I'm following the Capblancq and Forester paper for using the RDA offset. Some specific considerations are:

1. I'm using the same variables as I did for the RDA to identify adaptive loci, but excluding those that aren't environmental (so soil stuff) because I don't have future projections
2. using CHELSA future projection data for this

I'm running this all on the cluster, with a little bit of prep to get the inpute files sorted. The basic process is as follows:
*Step 1: Set up RDA*
1. Get MEMS that account for PCA
2. Read in PLINK data (using all available snps)
3. Get environmental data
4. Run RDA on environmental variables
Output: results/rda_justenv.rds
*Step 2: Set up for the genomic offset*
1. Read in the contemporary and future environmental data (stacked on local computer and moved over)
2. Project and remove NA values so they're compatible with other data sets
3. Read in the breeding range masking files (constructed on local computer and moved over)
4. Read in coords for individuals and get their contemporary environmental values
5. Center and scale these values so they are comparable
*Step 3: Run genomic offset*
1. source the genomic_offset.R script
2. run the genomic offset
3. convert results to dataframe and save that
Output: results/RDA_proj_offset_envonly_table.rds

# IMPORTANT CONSIDERATIONS
So, according the the Lind Lotterhos paper, using just the adaptive loci is basically the same and maybe a little worse than using all of the loci for genomic offset.

# INPUT FILES
## Getting environmental data ready
So, this is getting the contemporary (1981-2010) data ready for the important variables (bio06, bio09, and bio14). These get read in, stacked, projected.
```{r}
library(raster)
# 1981-2010
cont.paths <- list.files("../scratch/spatial_files/ClimateNA_1981-2010/", recursive = TRUE, full.names = TRUE)

cont.stack <- raster::stack(cont.paths)
names(cont.stack) <- c("bio06", "bio09", "bio14")

target_crs <- CRS("+proj=lcc +lat_1=48 +lat_2=33 +lon_0=-100 +datum=WGS84")
cont.stack_lcc <- raster::projectRaster(cont.stack, crs = target_crs, method = "bilinear")

# 2041-2070
futr.paths_41 <- list.files("../scratch/spatial_files/ClimateNA_2041-2070/", recursive = TRUE, full.names = T)
futr.stack_41 <- lapply(futr.paths_41, raster)
names(futr.stack_41) <- c("bio06", "bio09", "bio14")
plot(futr.stack_41$bio06)

futr.stack_41 <- raster::stack(futr.stack_41) %>% 
  raster::projectRaster(
    crs = "+proj=lcc +lat_1=48 +lat_2=33 +lon_0=-100 +datum=WGS84",
    method = "bilinear")
saveRDS(cont.stack, "../scratch/spatial_files/cont_stack.rds")
saveRDS(futr.stack_41, "../scratch/spatial_files/2041_futr_stack.rds")

# 2071-2100
futr.paths_71 <- list.files("../scratch/spatial_files/ClimateNA_2071-2100/", recursive = TRUE, full.names = T)
futr.stack_71 <- lapply(futr.paths_71, raster)
names(futr.stack_71) <- c("bio06", "bio09", "bio14")
plot(futr.stack_71$bio06)

futr.stack_71 <- raster::stack(futr.stack_71) %>% 
  raster::projectRaster(
    crs = "+proj=lcc +lat_1=48 +lat_2=33 +lon_0=-100 +datum=WGS84",
    method = "bilinear")
saveRDS(futr.stack_71, "../scratch/spatial_files/2071_futr_stack.rds")
```

I will also be using a breeding range mask for the genomic offset so it doesn't over project into areas that aren't relevant. Right now the breeding range is global and also a sf object. We need it to be just North America and also a raster file so it can work with the genomic offset function.
```{r}
# Breeding Range
map <- ne_states(country = c("United States of America", "Canada"), returnclass = 'sf')
na_union <- st_union(map)
breeding.sf <- st_read("../../../raw_data/banswa_range_2023.gpkg") %>% filter(season=="breeding") %>% st_transform(crs(map))
breeding_na <- st_intersection(breeding.sf, na_union)
breeding_na <- st_make_valid(breeding_na)
range_sf <- st_union(breeding_na) |> st_make_valid() |> st_transform(4326)
range_sp <- as(range_sf, "Spatial")
plot(range_sp)

template <- cont.stack[[1]]

# rasterize: 1 inside breeding range, NA outside
range_raster <- rasterize(
  range_sp,
  template,
  field = 1,
  background = NA
)

plot(range_raster)
saveRDS(range_sp, "../scratch/spatial_files/BANS_breeding_raster.rds")
```

## file checklist
---------- Sourced code
./src/genomic_offset.R
---------- RDS objects
BANS.mems.popstr_unrel.rds (MEMs / spatial covariates)
env2_narrow.rds (environmental table; must have the predictor columns you reference)
cont_stack.rds (current climate raster stack; must be Raster* if using your NA remover)
futr_stack.rds (future climate raster stack; same class/vars as cont_stack)
BANS_breeding_NA.rds
---------- Genotype / PLINK-ish inputs
BANS.unrel_ind.imputed4.1.raw (PLINK raw genotype file for adegenet::read.PLINK)
BANS.unrel_ind.imputed4.1.fam (used for ID order; note: not under data/ in your code)
---------- Sample lists / metadata
sample_unrel_short.txt (unrelated individuals list)
BANS_all_sample_data_shifted.csv (contains BGP_ID, Long, Lat)

# GENOMIC OFFSET
results/RDA_outliers_justenv.rds
```{bash}
rsync -avzP ericacnr@colostate.edu@login.rc.colorado.edu:/scratch/alpine/ericacnr@colostate.edu/BANS/06.genomic_offset/results/RDA_proj_offset_all_loci_table.rds /Users/ericarobertson/Desktop/BANS_adaptive_units/analysis/06.genomic_offset/ 
```

# VISUALIZING RESULTS
---------- Outputs you write
results/rda_justenvironment_narrow.rds
results/RDA_outliers_justenv.rds
results/RDA_proj_offset_table.rds

So, we can plot the adaptive loci from this environment only RDA.
## Plot loci and environmental vectors in RDA space (biplot)

Goal: visualize how loci and predictors align on RDA axes.
```{r, message=FALSE}
library(ggplot2)
library(vegan)

RDA_outliers <- readRDS("RDA_outliers_justenv.rds")

TAB_loci <- as.data.frame(scores(RDA_outliers, choices=c(1:2), display="species", scaling="none"))
TAB_var <- as.data.frame(scores(RDA_outliers, choices=c(1:2), display="bp"))

p <- ggplot() +
  geom_hline(yintercept=0, linetype="dashed", color = gray(.80), linewidth=0.6) +
  geom_vline(xintercept=0, linetype="dashed", color = gray(.80), linewidth=0.6) +
  geom_point(data = TAB_loci, aes(x=RDA1*3, y=RDA2*3), colour = "#EB8055FF", size = 2, alpha = 0.8) + #"#F9A242FF"
  geom_segment(data = TAB_var, aes(xend=RDA1, yend=RDA2, x=0, y=0), colour="black", linewidth=0.15, linetype=1, arrow=arrow(length = unit(0.02, "npc"))) +
  geom_text(data = TAB_var, aes(x=1.1*RDA1, y=1.1*RDA2, label = row.names(TAB_var)), size = 2.5, family = "Times") +
  xlab("RDA 1 (XX%)") + ylab("RDA 2 (XX%)") +
  facet_wrap(~"Adaptively not really enriched RDA space") +
  guides(color=guide_legend(title="Locus type")) +
  theme_bw(base_size = 11, base_family = "Times") +
  theme(panel.grid = element_blank(), plot.background = element_blank(), panel.background = element_blank(), strip.text = element_text(size=11))

ggsave("results/RDA_environmental_only_outlier_biplot.png")
```

## Genomic Offset

```{r}
library(RColorBrewer)
RDA_proj_offset <- readRDS("RDA_proj_offset_all_loci_table.rds")

## Projecting genomic offset on a map
colors <- c(colorRampPalette(brewer.pal(11, "Spectral")[6:5])(2), colorRampPalette(brewer.pal(11, "Spectral")[4:3])(2), colorRampPalette(brewer.pal(11, "Spectral")[2:1])(3))

coastlines <- ne_download(scale = 10, type = "coastline", category = "physical", returnclass = "sf")
countries <- ne_download(scale = 10, type = "admin_0_boundary_lines_land", category = "cultural", returnclass = "sf")
states <- ne_download(scale = 10, type = "admin_1_states_provinces_lines", category = "cultural", returnclass = "sf")
ocean <- ne_download(scale = 10, type = "ocean", category = "physical", returnclass = "sf")
lakes <- ne_download(scale = 10, type = "lakes", category = "physical", returnclass = "sf")

RDA_GO <- ggplot() +
  geom_sf(data = ocean, fill = "#bddbe6", color = NA) +
  geom_sf(data = lakes, fill = "#bddbe6", color = NA) +
  geom_raster(
    data = RDA_proj_offset,
    aes(x = x, y = y, fill = Global_offset)
  ) +
  geom_sf(data = countries, fill = NA, linewidth = 0.25) +
  geom_sf(data = coastlines, fill = NA, linewidth = 0.1225) +
  geom_sf(data = states, fill = NA, linewidth = 0.1225) +
  geom_sf(data = breeding_na, fill = NA, linewidth = 0.1225, alpha = 0.6) +
  scale_fill_gradientn(
    colours = rev(brewer.pal(11, "Spectral")),
    name = "Genomic offset"
  ) +
  coord_sf(xlim = range(RDA_proj_offset$x), ylim = range(RDA_proj_offset$y), expand = FALSE) +
  theme_bw(base_size = 11, base_family = "Times") +
  theme(
  panel.grid = element_blank(),
  strip.text = element_text(size = 11),
  legend.position = c(0.1, 0.18),   # x, y inside panel (0–1 scale)
  legend.justification = c("left", "bottom"),
  legend.background = element_rect(fill = "white", color = "black"),
  legend.key.height = unit(0.35, "cm"),
  legend.key.width  = unit(0.35, "cm")
)
  labs(title = "Bank Swallow breeding range | ClimateNA 2071–2100",
       subtitle = "GFDL-ESM4 SSP370",
       fill = "Genomic offset")

RDA_GO

ggsave(plot = RDA_GO, "results/full_RDA_genomic_offset_all_loci_2100_internallegend.png", width = 7, height=4.5)
```

# GO PER ESU Plots
```{r}
library(sf)
RDA_proj_offset <- readRDS("RDA_proj_offset_all_loci_table.rds")
ESU6_polygons <- st_read("../05.predictors_of_decline/BANS_ESU_K6_polygons_update_thresh45.shp")
ESU6_polygons$Cluster <- factor(
  ESU6_polygons$Cluster,
  levels = c("ESU1","ESU2","ESU3","ESU4","ESU5","ESU6"),
  labels = c(
    "Interior West",
    "Mackenzie Basin",
    "Pacific Coastal",
    "Newfoundland",
    "Alaska–Northwest",
    "East Coastal"
  )
)

pts <- st_as_sf(
  RDA_proj_offset,
  coords = c("x", "y"),
  crs = 4326,
  remove = FALSE
)

esu <- st_transform(ESU6_polygons, 4326)

# 3) join: assigns each point an ESU id
pts_esu <- st_join(pts, esu[, c("Cluster")], join = st_within, left = FALSE)
saveRDS(pts_esu, "pts_ESU6.rds")

# 4) summarize to ESU-level offset
esu_offset <- pts_esu %>%
  st_drop_geometry() %>%
  group_by(Cluster, Date) %>%
  summarize(
    n_cells = n(),
    mean_offset   = mean(Global_offset, na.rm = TRUE),
    median_offset = median(Global_offset, na.rm = TRUE),
    p05 = quantile(Global_offset, 0.05, na.rm = TRUE),
    p95 = quantile(Global_offset, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

esu_offset
saveRDS(esu_offset, "ESU6_offset_stats.rds")
```

```{r}
library(dplyr)
library(ggplot2)
library(ggdist)

names(cols) <- c(
  "Interior West",
  "Mackenzie Basin",
  "Pacific Coastal",
  "Newfoundland",
  "Alaska–Northwest",
  "East Coastal"
) 
# per-cell offsets (NOT summarized)
cells <- readRDS("pts_ESU6.rds") %>%
  st_drop_geometry() %>%
  mutate(Cluster = factor(Cluster, levels = names(cols)))  # keep your ESU order

# if you want only a specific future period, filter Date here:
# cells <- cells %>% filter(Date == "2068-2088")  # or whatever your Date values are
set.seed(5)  # for reproducibility

cells_subsample_1m <- cells[
  sample.int(nrow(cells), 1e6),
]

raincloud <- ggplot(cells, aes(x = Cluster, y = Global_offset, color = Cluster, fill = Cluster)) +
  # half density "trace" (one-sided)
  stat_halfeye(
    side = "right",
    adjust = 2,          # try 1.2–2.5
    width = 1.3,
    alpha = 0.75,
    .width = 0,            # <-- removes interval/box-looking part
    point_colour = NA,
    position = position_nudge(x = 0.1)  # <-- shifts halfeye right
  ) +
  # boxplot on top
  geom_boxplot(
    width = 0.18,
    outlier.shape = NA,
    alpha = 0.75,
    color = "black",
    position = position_nudge(x = -0.05),
    linewidth=0.4
  ) +
  scale_fill_manual(values = cols) +
  scale_color_manual(values = cols) +
  labs(
    title = "Within-ESU Distribution of Genomic Offset | 2071–2100",
    x = NULL,
    y = "Genomic Offset"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 8, angle = 90, hjust = 1),
    legend.position = "none"
  )

#raincloud
ggsave(plot=raincloud, "results/GO_per_ESU_raincloud.png", width = 6, height = 4.5, dpi = 300)

######### MORE SPACE VERSION
esu_order <- c(
  "Interior West",
  "Pacific Coastal",
  "Newfoundland",
  "Alaska–Northwest",
  "East Coastal",
  "Mackenzie Basin"
)

cells <- cells %>%
  mutate(Cluster = factor(Cluster, levels = esu_order))

# 2) optional: per-ESU "caps" so the spiky ones don't spill
halfeye_widths <- c(
  "Interior West"   = .5,
  "Pacific Coastal" = .5,
  "Newfoundland"    = .5,   # shrink these two
  "Alaska–Northwest"= .5,
  "East Coastal"    = .5,
  "Mackenzie Basin" = .5
)

p <- ggplot(cells, aes(x = Cluster, y = Global_offset, fill = Cluster)) +
  scale_fill_manual(values = cols) +
  scale_x_discrete(expand = expansion(mult = c(0.02, 0.02))) +
  labs(
    title = "Within-ESU Distribution of Genomic Offset | 2071–2100",
    x = NULL,
    y = "Genomic Offset"
  ) +
  theme_classic() +
  theme(
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none"
  )
for (esu in esu_order) {
  d_esu <- filter(cells, Cluster == esu)

  p <- p +
    stat_halfeye(
      data = d_esu,
      side = "right",
      adjust = 2,
      width = unname(halfeye_widths[esu]),
      alpha = 0.75,
      .width = 0,
      point_colour = NA,
      position = position_nudge(x = 0.30)
    ) +
    geom_boxplot(
      data = d_esu,
      width = 0.35,
      outlier.shape = NA,
      alpha = 0.75,
      color = "black",
      linewidth = 0.4,
      position = position_nudge(x = 0.10)
    )
}

p


ggsave(plot=p, "results/GO_per_ESU_raincloud_wide.png", width = 8, height = 2)
```

```{r}
esu_offset <- readRDS("ESU6_offset_stats.rds")
esu_long <- esu_offset %>%
  pivot_longer(
    cols = c(mean_offset, median_offset),
    names_to = "Metric",
    values_to = "Offset"
  )


cols <- readRDS("../02.5.delineate_ESUs/figures/colors/pcadmix_ESU_6_colors_2.rds")
names(cols) <- c(
  "Interior West",
  "Mackenzie Basin",
  "Pacific Coastal",
  "Newfoundland",
  "Alaska–Northwest",
  "East Coastal"
) 

ggplot(esu_offset, aes(x = Cluster, y = mean_offset, color=Cluster)) +
  geom_linerange(aes(ymin = p05, ymax = p95),
                 linewidth = 1.1,
                 color = "gray50") +
  geom_point(size = 3) +
  theme_bw(base_size = 12) +
  scale_color_manual(values=cols)+
  labs(
    title = "Within-ESU Distribution of Genomic Offset | 2071-2100",
    subtitle = "Points = mean, bars = 5th–95th percentile",
    y = "Genomic Offset",
    x = "ESU"
  )+
  theme(axis.text.x = element_text(size=8, 
                             angle =45,
                             hjust = 1),
        legend.position = "none")
ggsave("results/GO_per_ESU_mean_bar.png")
```

```{r}
ggplot(
  pts_esu,
  aes(x = Cluster, y = Global_offset)
) +
  geom_violin(outlier.size = 0.2) +
  facet_wrap(~ Date) +
  theme_bw(base_size = 12) +
  labs(
    title = "Distribution of Genomic Offset by ESU",
    y = "Genomic Offset",
    x = "Evolutionarily Significant Unit"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid = element_blank(),
    legend.position = "none"
  )
ggsave("results/GO_per_ESU_mean_bar.png")
```
# GO with ESU outlines
```{r}
library(RColorBrewer)
RDA_proj_offset <- readRDS("RDA_proj_offset_all_loci_table.rds")

## Projecting genomic offset on a map
colors <- c(colorRampPalette(brewer.pal(11, "Spectral")[6:5])(2), colorRampPalette(brewer.pal(11, "Spectral")[4:3])(2), colorRampPalette(brewer.pal(11, "Spectral")[2:1])(3))

coastlines <- ne_download(scale = 10, type = "coastline", category = "physical", returnclass = "sf")
countries <- ne_download(scale = 10, type = "admin_0_boundary_lines_land", category = "cultural", returnclass = "sf")
states <- ne_download(scale = 10, type = "admin_1_states_provinces_lines", category = "cultural", returnclass = "sf")
ocean <- ne_download(scale = 10, type = "ocean", category = "physical", returnclass = "sf")
lakes <- ne_download(scale = 10, type = "lakes", category = "physical", returnclass = "sf")

genoscape_polygons <- st_read("../BANS_ESU_K6_polygons_update_thresh45.shp")
bb <- readRDS("../05.predictors_of_decline/figures/bounding_box_breeedingNA.rds")

library(ggnewscale)
cols <- readRDS("../02.5.delineate_ESUs/figures/colors/pcadmix_ESU_6_colors_2.rds")

RDA_GO <- ggplot() +
  geom_sf(data = ocean, fill = "#bddbe6", color = NA) +
  geom_sf(data = lakes, fill = "#bddbe6", color = NA) +
  geom_raster(
    data = RDA_proj_offset,
    aes(x = x, y = y, fill = Global_offset)) +
  scale_fill_gradientn(
    colours = rev(brewer.pal(11, "Spectral")),
    name = "Genomic offset") +
  geom_sf(data = countries, fill = NA, linewidth = 0.25) +
  geom_sf(data = coastlines, fill = NA, linewidth = 0.1225) +
  geom_sf(data = states, fill = NA, linewidth = 0.1225) +
  ggnewscale::new_scale_fill() +  # reset fill scale for the polygons (even if we only use color)
  ggspatial::layer_spatial(
    genoscape_polygons,
    aes(color = factor(Cluster), fill = NA)) +
  scale_color_manual(values = cols, breaks = names(cols), name = "ESU") +
  #scale_fill_manual(values = cols, breaks = names(cols), name = "ESU")+
  guides(fill = "none")+
  coord_sf(xlim = range(RDA_proj_offset$x), ylim = range(RDA_proj_offset$y), expand = FALSE) +
  labs(
    title = "Bank Swallow breeding range | ClimateNA 2071–2100",
    subtitle = "GFDL-ESM4 SSP370"
  ) +
  theme_bw(base_size = 11, base_family = "Times") +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(size = 11)
  )

ggsave(plot = RDA_GO, "results/full_RDA_genomic_offset_all_loci_2100_ESUoverlaid.png", width = 7, height=4.5)
```

