---
title: "07.2.adaptive_structure"
output: html_document
---

# Libraries
```{r}
library(tidyverse)
library(ggthemes)
library(patchwork)
library(cowplot)
library(ggpubr)
library(sf)
library(raster)
library(terra)
library(rnaturalearth)
library(tess3r)
library(genoscapeRtools)
library(mignette)
library(ggplot2)
library(ggrepel)
library(pophelper)
library(data.table)
library(stringr)
library(gridExtra)
library(grid)
```

# RDA BIPLOTS
So, we've run an RDA and identified candidate SNPs. 
```{r}
cand <- read.table("../scratch/outliers/BANS.RDAresults.env_only.outliers_z3.txt", header=T)
nrow(cand)
head(cand)
```

# PULLING CANDIDATE SNPS
```{r}
infile <- "../scratch/outliers/BANS.RDAresults.env_only.outliers_z3.txt"

outfile <- "../scratch/outliers/BANS.RDAresults.env_only.outliers_z3_list.txt"

cand <- read.table(infile, header = TRUE)
head(cand)

cand_snps <- cand$snp

snp_list <- gsub("_[ACGT]$", "", cand_snps)

snp_list <- unique(snp_list)
head(snp_list)

writeLines(snp_list, outfile)
```

```{r}
library(readr)

models <- c(
  "env_mem4",
  "env_only",
  "env_pc1",
  "env_mems",
  "env_pc12"
)

for (model in models) {

  infile <- paste0(
    "../03.GEA/results/RDA_output/outliers/BANS.RDAresults.",
    model,
    ".outliers_qvalue.txt"
  )

  outfile <- paste0(
    "../scratch/outliers/BANS.",
    model,
    ".outliers_list_qvalue.txt"
  )

  cand <- read.table(infile, header = TRUE)

  # Extract SNP names
  cand_snps <- cand$Loci

  # Remove allele suffix
  snp_list <- gsub("_[ACGT]$", "", cand_snps)

  # Keep unique loci
  snp_list <- unique(snp_list)
  head(snp_list)

  writeLines(snp_list, outfile)

  cat("Wrote:", outfile, "\n")
}
```

[1] "BANS.env_mem4.outliers_list_qvalue
[2] "BANS.env_mems.outliers_list_qvalue
[3] "BANS.env_only.outliers_list_qvalue
[4] "BANS.env_pc1.outliers_list_qvalue
[5] "BANS.env_pc12.outliers_list_qvalue

# PCA PLOTS
```{bash}
conda activate GWAS2
vcf="/scratch/alpine/ericacnr@colostate.edu/BANS/02.5.neutral_pop_struc/BANS.filtered.unrel.vcf.gz"

plink --vcf $vcf --aec --recode --out BANS.filtered.unrel
#5219142 variants and 200 people pass filters and QC.

cat BANS.filtered.unrel.map | awk -F"\t" '{split($1, a, "__"); print $1"\t"a[1]"."$4"\t"$3"\t"$4}' > BANS.filtered.unrel_loc.map

plink --ped BANS.filtered.unrel.ped \
--map BANS.filtered.unrel_loc.map \
--make-bed --aec \
--out BANS.filtered.unrel

bfile="BANS.filtered.unrel"

plink --bfile $bfile --extract BANS.env_mems.outliers_list_qvalue.txt --make-bed --aec --out BANS.env_mems.outliers_list_qvalue
plink --bfile $bfile --extract BANS.env_only.outliers_list_qvalue.txt --make-bed --aec --out BANS.env_only.outliers_list_qvalue
plink --bfile $bfile --extract BANS.RDAresults.env_only.outliers_z3_list.txt --make-bed --aec --out BANS.env_only.outliers_z3_list

plink --bfile BANS.env_mems.outliers_list_qvalue \
--make-bed --aec \
--out BANS.env_mems.outliers_list_qvalue
plink --bfile BANS.env_only.outliers_list_qvalue \
--make-bed --aec \
--out BANS.env_only.outliers_list_qvalue
plink --bfile BANS.env_only.outliers_z3_list \
--make-bed --aec \
--out BANS.env_only.outliers_z3_list
```

```{bash}
plink --bfile BANS.env_mems.outliers_list_qvalue --aec --pca 199 --out BANS.env_mems.outliers_list_qvalue
plink --bfile BANS.env_only.outliers_list_qvalue --aec --pca 199 --out BANS.env_only.outliers_list_qvalue
plink --bfile BANS.env_only.outliers_z3_list --aec --pca 199 --out BANS.env_only.outliers_z3_list
```

```{bash}
rsync -avzP ericacnr@colostate.edu@login.rc.colorado.edu:/scratch/alpine/ericacnr@colostate.edu/BANS/04.delineate_AUs/BANS.env*.eigen* /Users/ericarobertson/Desktop/BANS_Pub/analysis/07.adaptive_analysis/scratch/PCA_eigen/
```

## Plotting env_only
```{r}
library(readr)
plink_pca <- read.table("../scratch/PCA_eigen/BANS.env_only.outliers_list_qvalue.eigenvec", header = FALSE) %>% dplyr::select(-V2)
eigenval <- scan("../scratch/PCA_eigen/BANS.env_only.outliers_list_qvalue.eigenval")

pve05 <- data.frame(PC = 1:45, pve = eigenval/sum(eigenval)*100)
ggplot(pve05[1:20, ], aes(x = PC, y = pve)) +
  geom_col(width = 0.9) +
  labs(x = "PC", y = "PVE (%)") +
  theme_bw()

# set names
pca05 <- plink_pca
colnames(pca05)[-1] <- paste0("PC", 1:(ncol(pca05) - 1))
names(pca05)[1] <- "ind"
nrow(pca05)
#correct number of samples

bg <- readRDS("../../03.delineate_ESUs/figures/colors/bg_colors_new.rds")

pops <- read.csv("../../03.delineate_ESUs/scratch/pops.csv") %>% filter(BGP_ID %in% pca05$ind)

# Join PCA data
pca.cluster <- left_join(pca05, pops %>% rename(ind=BGP_ID), by="ind")  # left_join PCA first to preserve PC columns

samp_unrel  <-  read.table("../../03.delineate_ESUs/scratch/BANS_unrel_sample_list_short_2.txt", header = F)
pca.cluster.norel <- pca.cluster %>% filter(ind %in% samp_unrel$V1)
nrow(pca.cluster.norel)

geo_key <- readRDS("../../03.delineate_ESUs/scratch/geo_informative_labels.RDS")
region_order <- c(
  "Cluster_1", "Cluster_5", "Cluster_17", "Cluster_4", "Cluster_8", "Cluster_13", "Cluster_2", "Cluster_21",
  "Cluster_3", "Cluster_7", "Cluster_18", "Cluster_19", "Cluster_22", "Cluster_20",
  "Cluster_12", "Cluster_15", "Cluster_14", "Cluster_10", "Cluster_9", "Cluster_11", "Cluster_6", "Cluster_16"
)

pca.cluster.norel <- pca.cluster.norel %>%
  left_join(geo_key, by = "Group") %>%
  mutate(
    Group = factor(Group, levels = region_order)
  )

# make one label per Group, in the same order
group_labels <- pca.cluster.norel %>%
  distinct(Group, Group_label) %>%
  arrange(Group)

label_vec <- setNames(group_labels$Group_label, group_labels$Group)

# make sure colors are in the same order
bg_use <- bg[region_order]

adaptive.pca <- ggplot(
  pca.cluster.norel,
  aes(
    x = PC1,
    y = PC2,
    color = Group,
    text = paste0(
      "ID: ", ind, "<br>",
      "Location: ", Group_label, "<br>",
      "PC1: ", round(PC1, 4), "<br>",
      "PC2: ", round(PC2, 4)
    )
  )
) +
  geom_point(size = 3) +
  scale_color_manual(
    values = bg_use,
    breaks = region_order,
    labels = label_vec[region_order],
    name = "Location"
  ) +
  theme_light() +
  xlab(paste0("PC1 (", signif(pve05$pve[1], 3), "%)")) +
  ylab(paste0("PC2 (", signif(pve05$pve[2], 3), "%)"))

adaptive.pca

ggsave("../figures/PCA/BANS_AU_env_only_PCA.png", plot=adaptive.pca, width = 8.5, height = 7)
```

## Plotting env_mems
```{r}
library(readr)
plink_pca <- read.table("../scratch/PCA_eigen/BANS.env_mems.outliers_list_qvalue.eigenvec", header = FALSE) %>% dplyr::select(-V2)
eigenval <- scan("../scratch/PCA_eigen/BANS.env_mems.outliers_list_qvalue.eigenval")

pve05 <- data.frame(PC = 1:14, pve = eigenval/sum(eigenval)*100)
ggplot(pve05[1:20, ], aes(x = PC, y = pve)) +
  geom_col(width = 0.9) +
  labs(x = "PC", y = "PVE (%)") +
  theme_bw()

# set names
pca05 <- plink_pca
colnames(pca05)[-1] <- paste0("PC", 1:(ncol(pca05) - 1))
names(pca05)[1] <- "ind"
nrow(pca05)
#correct number of samples

bg <- readRDS("../../03.delineate_ESUs/figures/colors/bg_colors_new.rds")

pops <- read.csv("../../03.delineate_ESUs/scratch/pops.csv") %>% filter(BGP_ID %in% pca05$ind)

# Join PCA data
pca.cluster <- left_join(pca05, pops %>% rename(ind=BGP_ID), by="ind")  # left_join PCA first to preserve PC columns

samp_unrel  <-  read.table("../../03.delineate_ESUs/scratch/BANS_unrel_sample_list_short_2.txt", header = F)
pca.cluster.norel <- pca.cluster %>% filter(ind %in% samp_unrel$V1)
nrow(pca.cluster.norel)

geo_key <- readRDS("../../03.delineate_ESUs/scratch/geo_informative_labels.RDS")
region_order <- c(
  "Cluster_1", "Cluster_5", "Cluster_17", "Cluster_4", "Cluster_8", "Cluster_13", "Cluster_2", "Cluster_21",
  "Cluster_3", "Cluster_7", "Cluster_18", "Cluster_19", "Cluster_22", "Cluster_20",
  "Cluster_12", "Cluster_15", "Cluster_14", "Cluster_10", "Cluster_9", "Cluster_11", "Cluster_6", "Cluster_16"
)

pca.cluster.norel <- pca.cluster.norel %>%
  left_join(geo_key, by = "Group") %>%
  mutate(
    Group = factor(Group, levels = region_order)
  )

# make one label per Group, in the same order
group_labels <- pca.cluster.norel %>%
  distinct(Group, Group_label) %>%
  arrange(Group)

label_vec <- setNames(group_labels$Group_label, group_labels$Group)

# make sure colors are in the same order
bg_use <- bg[region_order]

adaptive.pca <- ggplot(
  pca.cluster.norel,
  aes(
    x = PC1,
    y = PC2,
    color = Group,
    text = paste0(
      "ID: ", ind, "<br>",
      "Location: ", Group_label, "<br>",
      "PC1: ", round(PC1, 4), "<br>",
      "PC2: ", round(PC2, 4)
    )
  )
) +
  geom_point(size = 3) +
  scale_color_manual(
    values = bg_use,
    breaks = region_order,
    labels = label_vec[region_order],
    name = "Location"
  ) +
  theme_light() +
  xlab(paste0("PC1 (", signif(pve05$pve[1], 3), "%)")) +
  ylab(paste0("PC2 (", signif(pve05$pve[2], 3), "%)"))

adaptive.pca

ggsave("../figures/PCA/BANS_AU_env_mems_PCA.png", plot=adaptive.pca, width = 8.5, height = 7)
```
## Plotting env_only
```{r}
library(readr)
plink_pca <- read.table("../scratch/PCA_eigen/BANS.env_only.outliers_z3_list.eigenvec", header = FALSE) %>% dplyr::select(-V2)
eigenval <- scan("../scratch/PCA_eigen/BANS.env_only.outliers_z3_list.eigenval")

pve05 <- data.frame(PC = 1:199, pve = eigenval/sum(eigenval)*100)
ggplot(pve05[1:20, ], aes(x = PC, y = pve)) +
  geom_col(width = 0.9) +
  labs(x = "PC", y = "PVE (%)") +
  theme_bw()

# set names
pca05 <- plink_pca
colnames(pca05)[-1] <- paste0("PC", 1:(ncol(pca05) - 1))
names(pca05)[1] <- "ind"
nrow(pca05)
#correct number of samples

bg <- readRDS("../../03.delineate_ESUs/figures/colors/bg_colors_new.rds")

pops <- read.csv("../../03.delineate_ESUs/scratch/pops.csv") %>% filter(BGP_ID %in% pca05$ind)

# Join PCA data
pca.cluster <- left_join(pca05, pops %>% rename(ind=BGP_ID), by="ind")  # left_join PCA first to preserve PC columns

samp_unrel  <-  read.table("../../03.delineate_ESUs/scratch/BANS_unrel_sample_list_short_2.txt", header = F)
pca.cluster.norel <- pca.cluster %>% filter(ind %in% samp_unrel$V1)
nrow(pca.cluster.norel)

geo_key <- readRDS("../../03.delineate_ESUs/scratch/geo_informative_labels.RDS")
region_order <- c(
  "Cluster_1", "Cluster_5", "Cluster_17", "Cluster_4", "Cluster_8", "Cluster_13", "Cluster_2", "Cluster_21",
  "Cluster_3", "Cluster_7", "Cluster_18", "Cluster_19", "Cluster_22", "Cluster_20",
  "Cluster_12", "Cluster_15", "Cluster_14", "Cluster_10", "Cluster_9", "Cluster_11", "Cluster_6", "Cluster_16"
)

pca.cluster.norel <- pca.cluster.norel %>%
  left_join(geo_key, by = "Group") %>%
  mutate(
    Group = factor(Group, levels = region_order)
  )

# make one label per Group, in the same order
group_labels <- pca.cluster.norel %>%
  distinct(Group, Group_label) %>%
  arrange(Group)

label_vec <- setNames(group_labels$Group_label, group_labels$Group)

# make sure colors are in the same order
bg_use <- bg[region_order]

adaptive.pca <- ggplot(
  pca.cluster.norel,
  aes(
    x = PC1,
    y = PC2,
    color = Group,
    text = paste0(
      "ID: ", ind, "<br>",
      "Location: ", Group_label, "<br>",
      "PC1: ", round(PC1, 4), "<br>",
      "PC2: ", round(PC2, 4)
    )
  )
) +
  geom_point(size = 3) +
  scale_color_manual(
    values = bg_use,
    breaks = region_order,
    labels = label_vec[region_order],
    name = "Location"
  ) +
  theme_light() +
  xlab(paste0("PC1 (", signif(pve05$pve[1], 3), "%)")) +
  ylab(paste0("PC2 (", signif(pve05$pve[2], 3), "%)"))

adaptive.pca

ggsave("../figures/PCA/BANS.env_only.outliers_z3_list.png", plot=adaptive.pca, width = 8.5, height = 7)
```

# ADMIXTURE
```{bash}
pcangsd_path="/projects/ericacnr@colostate.edu/pcangsd"
bfile="BANS.env_only.outliers_z3_list"
export PYTHONPATH="/projects/ericacnr@colostate.edu/pcangsd:$PYTHONPATH"

for eig in {1..5}
do
  for r in {1..5}
         do
          $CONDA_PREFIX/bin/python $pcangsd_path/pcangsd/main.py -p $bfile -t 4 --admix --admix-seed $RANDOM -e $eig --admix-alpha 0 -o pcadmix_output/BANS.env_only.outliers_list_z3.alpha0.e"$eig".run"$r"
        done
done
```

```{bash}
rsync -avzP ericacnr@colostate.edu@login.rc.colorado.edu:/scratch/alpine/ericacnr@colostate.edu/BANS/04.delineate_AUs/pcadmix_output/BANS.*z3*.Q /Users/ericarobertson/Desktop/BANS_Pub/analysis/07.adaptive_analysis/scratch/pcadmix_output/
```

# FUNCTIONS FOR PLOTTING
(Note, the following functions were generated using ChatGPT based on the above written out code that I produced, interpretation of results is my own)
## Color Pallets
The trick here is to get the color right for the spatial ESU rather than having the numbers be the same, cause the numbers aren't consistent across space with different K values.
```{r}
# G3
cols_G3 <-  c(
  "AU1" = "#B8E152",
  "AU2" = "#EF7A58",
  "AU3" = "#44649F")
barplot(
  rep(1, length(cols_G3)),
  col = cols_G3,
  border = NA,
  axes = FALSE,
  space = 0,
  names.arg = names(cols_G3),
  las = 2
)
saveRDS(cols_G3, "../figures/AU_grouping_colors_3_G3.rds")

# G4
cols_G4 <-  c(
  "AU1" = "#FF5A5F",
  "AU2" = "#187795",
  "AU3" = "#FFBF46",
  "AU4" = "#C1839F")
barplot(
  rep(1, length(cols_G4)),
  col = cols_G4,
  border = NA,
  axes = FALSE,
  space = 0,
  names.arg = names(cols_G4),
  las = 2
)
saveRDS(cols_G4, "../figures/AU_grouping_colors_3_G4.rds")

#G5
cols_G5 <-  c(
  "AU1" = "#187795",
  "AU2" = "#FF5A5F",
  "AU3" = "#FFBF46",
  "AU4" = "#C1839F",
  "AU5" = "#8ACB88")
barplot(
  rep(1, length(cols_G5)),
  col = cols_G5,
  border = NA,
  axes = FALSE,
  space = 0,
  names.arg = names(cols_G5),
  las = 2
)
saveRDS(cols_G5, "../figures/AU_grouping_colors_3_G5.rds")

# G6
cols_G6 <-  c(
  "AU1" = "#187795",
  "AU2" = "#FF5A5F",
  "AU3" = "#FFBF46",
  "AU4" = "#C1839F",
  "AU5" = "#8ACB88",
  "AU6" = "#2589BD")
barplot(
  rep(1, length(cols_G6)),
  col = cols_G6,
  border = NA,
  axes = FALSE,
  space = 0,
  names.arg = names(cols_G6),
  las = 2
)
saveRDS(cols_G6, "../figures/AU_grouping_colors_3_G6.rds")
```

## Genoscapes
```{r}
source("../../03.delineate_ESUs/code/prep_genoscape_inputs.R")
source("../../03.delineate_ESUs/code/build_genoscape_surfaces.R")

############# PICK A K VALUE #############
scheme <- "G4"
K <- 4
# read in data, either raw or forced
Q_tibble <- read.table("../scratch/pcadmix_output/BANS.env_only.outliers_list_z3.alpha0.e3.run1.admix.4.Q") %>%
setNames(paste0("AU", 1:K))

scheme <- "G5"
K <- 5
Q_tibble <- read.table("../scratch/pcadmix_output/BANS.env_only.outliers_list_z3.alpha0.e4.run1.admix.5.Q") %>%
setNames(paste0("AU", 1:K))

scheme <- "G3"
K <- 3
Q_tibble <- read.table("../scratch/pcadmix_output/BANS.env_only.outliers_list_z3.alpha0.e2.run1.admix.3.Q") %>%
setNames(paste0("AU", 1:K))

out_plot_genoscape_png <- paste0("BANS_genoscape_",scheme,"_adaptive_final.png")
out_plot_genoscape_pdf <- paste0("BANS_genoscape_",scheme,"_adaptive_final.pdf")
genoscape_path <- "../figures/genoscape/"
cols <- readRDS(paste0("../figures/AU_grouping_colors_3_G4.rds"))
cols <- cols[1:3]
barplot(rep(1, length(cols)), col = cols, border = NA, axes = FALSE, space = 0, names.arg = names(cols), las = 2)

############# OTHER INPUT FILES #############
pca_df <- read.table("../scratch/PCA_eigen/BANS.env_only.outliers_z3_list.eigenvec", header = FALSE)

all <- read_csv("../../../raw_data/BANS_all_sample_data_shifted.csv")
############# MAKE Q_TIBBLE #############
prep_out <- prep_genoscape_inputs(
  Q_tibble = Q_tibble,
  pca_df = pca_df,
  all = all
)

Q_tibble <- prep_out$Q_tibble
Q_matrix <- prep_out$Q_matrix
long_lat_matrix <- prep_out$long_lat_matrix
LatLong_tibble <- prep_out$LatLong_tibble
K <- ncol(Q_matrix)
q_cols <- colnames(Q_matrix)

############# GET BASE MAP AND RANGE #############
map <- ne_states(
  country = c("United States of America", "Canada"),
  returnclass = "sf"
)

na_union <- st_union(map)

breeding.sf <- st_read("../../../raw_data/spatial_files/banswa_range_2023.gpkg") %>%
  filter(season == "breeding") %>%
  st_transform(crs(map))

breeding_na <- st_intersection(breeding.sf, na_union)

############# BUILD THE GENOSCAPE #############
surf_out <- build_genoscape_surfaces(
  Q_matrix = Q_matrix,
  long_lat_matrix = long_lat_matrix,
  breeding_na = breeding_na,
  cols = cols,
  K = K,
  brick_names = q_cols,
  out_tif = NULL,
  alpha_threshold = 0,
  krig_val = 15
)

genoscape_brick <- surf_out$genoscape_brick
genoscape_projected <- surf_out$genoscape_projected

coords_sf <- st_as_sf(
  LatLong_tibble,
  coords = c("Long", "Lat"),
  crs = 4326   # WGS84 (lon/lat)
)

# then we want to get the sample points as coded by letter, which I established previously
cluster_geo <- readRDS("../scratch/geo_informative_labels_centroids.RDS")

## make sf points from lon/lat
cluster_geo_sf <- st_as_sf(
  cluster_geo,
  coords = c("Y", "X"),
  crs = 4326,      # lon/lat
  remove = FALSE
)

## transform to your map CRS (lamproj must be a valid CRS/proj string)
cluster_geo_sf <- st_transform(cluster_geo_sf, st_crs(lamproj))

genoscape <- ggplot() +
  ggspatial::layer_spatial(genoscape_projected) +
  geom_sf(data = breeding_na, fill = NA, linewidth = 0.1225 , alpha = 0.6) +
  geom_sf(data = countries, fill = NA, linewidth = 0.25) +
  geom_sf(data = coastlines, fill = NA, linewidth = 0.1225) +
  geom_sf(data = states, fill = NA, linewidth = 0.1225) +
  geom_sf(data = ocean, fill = "#bddbe6") +
  geom_sf(data = lakes, fill = "#bddbe6") + #bddbe6
  #geom_spatial_point(data = long_lat_tibble, mapping = aes(x = Long, y = Lat), size=1, alpha = 0.7) + 
  theme(panel.border = element_rect(color = "black", fill = NA, linewidth = 1)) +
  geom_sf(data=cluster_geo_sf,
          size = 0.5)+
  coord_sf(
  crs = st_crs(lamproj),
  xlim = c(bb["xmin"], bb["xmax"]),
  ylim = c(bb["ymin"], bb["ymax"]),
  expand = FALSE)+
  geom_label_repel(
    data = cluster_geo_sf,
    aes(label = Group_label, geometry = geometry),
    stat = "sf_coordinates",
    size = 1.5,            # increase text size (try 2.2–3.0
    fontface = "bold",
    label.size = .1,
    label.padding = unit(0.1, "lines"),
    label.r=0,
    fill = "white",
    alpha = 1,
    point.padding = 0.1,
    segment.color = "black",
    segment.size = 0.25,
    min.segment.length = 0
)

############# SAVE THE GENOSCAPE #############
ggsave(plot=genoscape, path=genoscape_path, filename=out_plot_genoscape_png, width = 7, height = 4.5)

pdf(paste0(genoscape_path,out_plot_genoscape_pdf))
print(genoscape)
dev.off()
```

## Admixture plots
```{r}
source("make_admix_barplot_and_assignments_adaptive.R")

######## INPUT #########
K <-  5
out_plot_admix <- "BANS_admix_overall_alpha0_adaptive_G5_letters_2.png"
admix_plot_path <- "../figures/admix_plots/"

out_group_admix <- "BANS_admix_grouping_adaptive_G5_2.tsv"
admix_group_path <- "../scratch/admix_output/"

Q_tibble <- read.table("../scratch/pcadmix_output/BANS.env_only.outliers_list_z3.alpha0.e4.run1.admix.5.Q") %>%
  setNames(paste0("AU", 1:K))
cols <- readRDS("../figures/AU_grouping_colors_3_G5.rds")

K <-  4
out_plot_admix <- "BANS_admix_overall_alpha0_adaptiveG4_letters_2.png"
admix_plot_path <- "../figures/admix_plots/"

out_group_admix <- "BANS_admix_grouping_adaptive_G4_2.tsv"
admix_group_path <- "../scratch/admix_output/"

Q_tibble <- read.table("../scratch/pcadmix_output/BANS.env_only.outliers_list_z3.alpha0.e3.run1.admix.4.Q") %>%
  setNames(paste0("AU", 1:K))
cols <- readRDS("../figures/AU_grouping_colors_3_G4.rds")

K <-  3
out_plot_admix <- "BANS_admix_overall_alpha0_adaptive_G3_letters_2.png"
admix_plot_path <- "../figures/admix_plots/"

out_group_admix <- "BANS_admix_grouping_adaptive_G3_2.tsv"
admix_group_path <- "../scratch/admix_output/"

Q_tibble <- read.table("../scratch/pcadmix_output/BANS.env_only.outliers_list_z3.alpha0.e2.run1.admix.3.Q") %>%
  setNames(paste0("AU", 1:K))
cols <- readRDS("../figures/AU_grouping_colors_3_G6.rds")
cols <- cols[1:3]

######## RUNNING FUNCTION ########
pca_df <- read.table("../scratch/PCA_eigen/BANS.env_only.outliers_z3_list.eigenvec", header = FALSE)

pops <- readr::read_csv("../../03.delineate_ESUs/scratch/pops.csv", show_col_types = FALSE)
letter_key <- readRDS("../../03.delineate_ESUs/scratch/geo_informative_labels_centroids.RDS")

group_label_order <- c("WA", "NT", 
                       "AK", "YT", "SK", "MB_1", "MB_2", "MT", "ND", "WY_1", "WY_2", "AB", "BC_1", "BC_2",
                       "ON", "QC_1", "QC_2", "NS_1", "NS_2",
                       "NB", "PE", "NL")

admix_out <- make_admix_barplot_and_assignments(
  K_target = K,
  cols = cols,
  group_label_order = group_label_order,
  threshold = 0.45,
  Q_tibble = Q_tibble,
  pca_df = pca_df,
  pops = pops,
  geo_key = letter_key,
  out_plot = paste0(admix_plot_path,out_plot_admix),
  out_assignments = paste0(admix_group_path,out_group_admix)
)

K_plot <- admix_out$K_plot
admix_group <- admix_out$admix_group
q_long <- admix_out$q_long

# This plot is meaningless for the GM ones but easiest to just write it out anyway
K_plot
ggsave(path=admix_plot_path, filename=out_plot_admix, plot=K_plot, width = 10, height = 2.2, units = "in", dpi = 300, bg = "white")

write_tsv(admix_group, paste0(admix_group_path,out_group_admix))

############
pdf("../figures/admix_plots/BANS_admix_overall_alpha0_G4_letters.pdf", width = 10, height = 2.2)
print(K_plot)
dev.off()
```

## PCA plots
```{r}
source("../../03.delineate_ESUs/code/plot_pca_admix_solid.R")

out_plot_pca1 <- "../figures/pca_plots/BANS_G4_PCA_AU_coloring.png"
out_plot_pca2 <- "../figures/pca_plots/BANS_G4_PCA_AU_coloring.png"
base_cols <- "../../06.final_figures/colors/AU_grouping_colors_K4_2.rds"

admix_groups <- data.table::fread("../scratch/admix_output/BANS_admix_grouping_adaptive_G4_2.tsv")
cols <- readRDS(base_cols)

plink_pca <- read.table("../scratch/PCA_eigen/BANS.env_only.outliers_z3_list.eigenvec", header = FALSE)
pops <- readr::read_csv("../../03.delineate_ESUs/scratch/pops.csv", show_col_types = FALSE)
eigenval <- scan("../scratch/PCA_eigen/BANS.env_only.outliers_z3_list.eigenval")

solid_out <- plot_pca_admix_solid(
  plink_pca = plink_pca,
  eigenval = eigenval,
  cols = cols,
  pops = pops,
  admix_groups = admix_groups,
  out_plot = out_plot_pca1,
)

solid_out$pca_plot
```

```{r}
geo_key <- readRDS("../../03.delineate_ESUs/scratch/geo_informative_labels.RDS")
base_cols <- readRDS("../../06.final_figures/colors/AU_grouping_colors_K4_2.rds")
base_cols <- base_cols[1:3]

out_group_label_cols_rds <- "../figures/pcadmix_AU_3_group_label_colors.rds"
out_plot <- "../figures/pca_plots/BANS_K3_PCA_AU_coloring.pdf"

desired_order <- NULL

admix_groups <- data.table::fread("../scratch/admix_output/BANS_admix_grouping_adaptive_G3_2.tsv")

pca_df <- plink_pca %>%
  select(-V2)

colnames(pca_df)[-1] <- paste0("PC", seq_len(ncol(pca_df) - 1))
names(pca_df)[1] <- "ind"

pve <- tibble(
  PC = seq_along(eigenval),
  pve = eigenval / sum(eigenval) * 100
)

# -----------------------------
# Metadata
# -----------------------------
pops.groups.labels <- pops %>%
  transmute(BGP_ID, Group) %>%
  left_join(geo_key, by = "Group") %>%
  left_join(admix_groups, by = "BGP_ID") %>%
  mutate(
    Group_label = trimws(as.character(Group_label)),
    admix_group = replace_na(admix_group, "No Assignment")
  )


base_cols["No Assignment"] <- "#BDBDBD"

label_owner <- pops.groups.labels %>%
  filter(!is.na(Group_label)) %>%
  count(Group_label, admix_group, name = "n") %>%
  group_by(Group_label) %>%
  slice_max(n, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  rename(owner_au = admix_group) %>%
  mutate(owner_au = replace_na(owner_au, "No Assignment"))

missing_cols <- setdiff(unique(label_owner$owner_au), names(base_cols))

group_label_cols_df <- label_owner %>%
  arrange(owner_au, Group_label) %>%
  group_by(owner_au) %>%
  group_modify(~{
    base <- base_cols[[as.character(.y$owner_au)]]
    labels <- .x$Group_label
    
    tibble(
      Group_label = labels,
      col = colorRampPalette(c("white", base))(length(labels) + 1)[-1]
    )
  }) %>%
  ungroup()

group_label_cols <- setNames(
  group_label_cols_df$col,
  group_label_cols_df$Group_label
)

saveRDS(group_label_cols, out_group_label_cols_rds)

# Optional color check
barplot(
  rep(1, length(group_label_cols)),
  col = group_label_cols,
  border = NA,
  names.arg = names(group_label_cols),
  las = 2,
  cex.names = 0.3
)

# -----------------------------
# Join PCA + metadata
# -----------------------------
pca.cluster <- pca_df %>%
  left_join(
    pops.groups.labels %>%
      left_join(
        label_owner %>% select(Group_label, owner_au),
        by = "Group_label"
      ) %>%
      rename(ind = BGP_ID),
    by = "ind"
  ) %>%
  mutate(
    Group_label = trimws(as.character(Group_label)),
    admix_group = replace_na(admix_group, "No Assignment"),
    owner_au = replace_na(owner_au, "No Assignment")
  )

# -----------------------------
# Legend order
# -----------------------------
if (!is.null(desired_order)) {
  desired_order <- as.character(desired_order)
  
  missing_from_palette <- setdiff(desired_order, names(group_label_cols))
  
  if (length(missing_from_palette) > 0) {
    warning(
      "These labels are in desired_order but not in group_label_cols: ",
      paste(missing_from_palette, collapse = ", ")
    )
  }
  
  legend_order <- c(
    desired_order[desired_order %in% names(group_label_cols)],
    sort(setdiff(names(group_label_cols), desired_order))
  )
} else {
  legend_order <- sort(names(group_label_cols))
}

group_label_cols <- group_label_cols[legend_order]

pca.cluster <- pca.cluster %>%
  mutate(Group_label = factor(Group_label, levels = legend_order))

# -----------------------------
# Plot
# -----------------------------
p <- ggplot(
  pca.cluster,
  aes(
    x = PC1,
    y = PC2,
    color = Group_label
  )
) +
  geom_point(size = 3) +
  scale_color_manual(
    values = group_label_cols,
    limits = legend_order,
    drop = FALSE,
    name = "Sampling location\nColored by AU"
  ) +
  theme_classic() +
  xlab(paste0("PC1 (", signif(pve$pve[1], 3), "%)")) +
  ylab(paste0("PC2 (", signif(pve$pve[2], 3), "%)"))

p

ggsave(
  filename = out_plot,
  plot = p,
  width = 6,
  height = 4
)

grad_out$pca_plot

pdf("../figures/pca_plots/BANS_K4_PCA_AU_coloring.pdf", width=6, height=4)
print(grad_out$pca_plot)
dev.off()
```

# GENOSCAPE SHAPE FILES
I want to create shapefiles for each genetic group to get demographic data for, either from ebird population trends or BBS. Matt has function for this in Mignette, so install and load that package if you haven't already.
```{r, message = FALSE, warning = FALSE}
library(mignette)

lamproj <- "+proj=lcc +lat_1=20 +lat_2=60 +lat_0=40 +lon_0=-100 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"

out_polygon <- "../outputs/BANS_AU_polygons_thresh.55_K3.shp"
names(genoscape_brick) <- paste0("AU", 1:K)

crs(genoscape_brick) <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"
genoscape_brick_spat <- terra::rast(genoscape_brick)
genoscape_brick_proj <- terra::project(genoscape_brick_spat, lamproj, method = "near")

genoscape_polygons <- mignette::scape_to_shape(x = genoscape_brick_proj, 
                                               prob_threshold = 0.55,
                                               d = 100,
                                               f = 100,
                                               s = 3)

polygons <- ggplot() +
  ggspatial::layer_spatial(genoscape_polygons, aes(fill = as.factor(Cluster), alpha=0.7)) +
  #geom_sf(data = breeding_range, fill = NA, linewidth = 0.1225 , alpha = 0.6) +
  geom_sf(data = countries, fill = NA, linewidth = 0.25) +
  geom_sf(data = coastlines, fill = NA, linewidth = 0.1225) +
  geom_sf(data = states, fill = NA, linewidth = 0.1225) +
  #geom_sf(data = ocean, fill = "#bddbe6") +
  #geom_sf(data = lakes, fill = "#bddbe6") +
  #geom_spatial_point(data = long_lat_tibble, mapping = aes(x = Long, y = Lat), size=2) +
  theme_void() + 
  #theme(panel.border = element_rect(color = "black", fill = NA, linewidth = 1)) +
  coord_sf(
  crs = st_crs(lamproj),
  xlim = c(bb["xmin"], bb["xmax"]),
  ylim = c(bb["ymin"], bb["ymax"]),
  expand = FALSE)  # expand = FALSE means put the axes right at the xlim and ylim

polygons
st_write(genoscape_polygons, out_polygon, delete_dsn = TRUE)
```